### 2024-12-06 Short_note_captured_phenotype_journal_version.qmd
Importing R code

---
title: "Short_note_captured_phenotype"
format: html
---

```{r loading used packages and functions}
library(stringr)#gestion chaines de caractères
library(car)#anova ou graph d'analyse
library(RVAideMemoire)#variance
library(corrplot)# graphe correlation
library(ggeffects) #effet prédit
library(ggplot2) #graphs
library(ggtext) #customization of ggplot legends
library(ggpubr) #custom graph (comparison bars)

library(multcomp) #multiple comparison of contrasts

library(data.table) #large dataset gestion

library(lme4) #mixed model
library(blme) #bayesian mixed model (handling singular fit)
library(DHARMa) #model diagnostic (mixed models/non linear models)
library(MuMIn) #r-squared for glmm (Nakagawa method)
library(rsq) #diverse rsquared including Nagelkerke

library(emmeans)#comparison of marginal means from GLM

library(sf) #spatial treatments

library(texreg) #export table in Latex format
library(kableExtra)

library(MASS) #used to generate multivariate normal distribution

library(lubridate) #handling date data

sumna <- function(x) {
    ret <- sum(x, na.rm = TRUE)
    if(ret == 0 && all(is.na(x))) {NA_real_} else {ret}
}

#fisher.test from scinded tables:

fisher.test2 <- function(a, b) {
    return(fisher.test(cbind(a, b)))
}

```

# Importing data

```{r importing data}

# Strandings:

Strandings <- fread("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/Strandings_journal_version.csv")

# Declaration data:

Declaration_Bycatch = fread("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/Declaration_Bycatch_journal_version.csv")

# Observation data (OBSMER programme):

OBSMER_Bycatch <- fread("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/OBSMER_Bycatch_journal_version.csv")

Strandings[, TRIMESTER := as.character(TRIMESTER)]
Declaration_Bycatch[, TRIMESTER := as.character(TRIMESTER)]
OBSMER_Bycatch[, TRIMESTER := as.character(TRIMESTER)] # converting trimester in categories, not numbers

Declaration_Bycatch[TARGETED_TAXA == "MZZ - Osteichthyes", TARGETED_TAXA:=NA]
OBSMER_Bycatch[TARGETED_TAXA == "MZZ - Osteichthyes", TARGETED_TAXA:=NA] # we considered the declarations of Osteichtyes to be equivalent to DECLARATIONS OF an unknown taxon, and therefore considered these values equivalent to NA for our analyses

# Merging data in common between declaration and observation datasets:

Decl_Obs_Merged_Bycatch = rbind( cbind(Declaration_Bycatch[!is.na(BODY_MASS), .(SPECIES, ICES_DIV, TRIMESTER, YEAR, FISHING_GEAR, MESH_SIZE, TARGETED_TAXA, TARGETED_ISSCAAP, BODY_MASS)], ORIGIN= "DECL"), cbind(OBSMER_Bycatch[!is.na(BODY_MASS), .(SPECIES, ICES_DIV, TRIMESTER, YEAR, FISHING_GEAR, MESH_SIZE, TARGETED_TAXA, TARGETED_ISSCAAP, BODY_MASS)], ORIGIN= "OBSMER"), use.names=F)

# Merging data in common between declaration and observation datasets, by only considering explanatory variables:

Decl_Obs_Merged_Bycatch_explanatory = rbind( cbind(Declaration_Bycatch[, .(SPECIES, ICES_DIV, TRIMESTER, YEAR, FISHING_GEAR, MESH_SIZE, TARGETED_TAXA, TARGETED_ISSCAAP)], ORIGIN= "DECL"), cbind(OBSMER_Bycatch[, .(SPECIES, ICES_DIV, TRIMESTER, YEAR, FISHING_GEAR, MESH_SIZE, TARGETED_TAXA, TARGETED_ISSCAAP)], ORIGIN= "OBSMER"), use.names=F)

# paramètres à garder:

# Decl_Obs_Merged_Bycatch = all_merged[, c("ORIGIN", "ESP_LIB_FAO", "SECT_COD_SACROIS_NIV3", "SECT_COD_SACROIS_NIV3b", "TRIMESTRE", "AN", "YEAR_DCO", "YEAR_PHO", "ENGIN2", "ENGIN4", "MAILLAGE", "TARGET_TAXON", "TARGET_TAXON3", "TARGET_TAXON4", "poids_kg")]

```

## Preliminary statistics

We first described the used datasets (sample size, removal of explanatory variables too rarely described or too collinear with others, correlation between studied response traits). We then explored the distribution of captures as a function of external variables (spatial, temporal, environmental).

```{r preliminary stat}

#### Sample sizes:

# Distribution of recorded captured species (declaration and observation datasets / Strandings):

table(Decl_Obs_Merged_Bycatch_explanatory$SPECIES)
table(Strandings$SPECIES)

# Distribution of by-catch among recording programmes:

table(Decl_Obs_Merged_Bycatch_explanatory[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),]$ORIGIN)

#### Removal of variables with two many missing values or too strong correlation patterns:

# Distribution of missing values in the used datasets:

apply(Strandings, 2, function(x) table(is.na(x))) # few missing values in proportion, except for body mass, which estimations are rarely reported

apply(Declaration_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),], 2, function(x) table(is.na(x))) # in particular, targeted taxa were reported in ~68% of cases (very few missing values otherwise)

apply(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),], 2, function(x) table(is.na(x))) # in contrast to body size, the sex and weight of individuals is often not reported / gear speed measurement was only available for trawls resulting in a particularly high proportion of missing values (~48%), others fishing techniques variables are associated with a non negligible amount of missing values (gear depth: ~14%, presence of an acoustic deterrent: ~17%)

# Particularly important correlation/association patterns between explanatory variables:

# We have only detailed here the most important correlation/association patterns that prevent variables from being used together in all subsequent analyses.

kruskal.test(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$GEAR_DIMENSION ~ OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$FISHING_GEAR)
kruskal.test(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$GEAR_DIMENSION ~ OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$TARGETED_TAXA)
cor.test(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$GEAR_DIMENSION, OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$MESH_SIZE, method = "spearman")

ggplot(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], aes(FISHING_GEAR, GEAR_DIMENSION)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
ggplot(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], aes(TARGETED_TAXA, GEAR_DIMENSION)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
ggplot(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], aes(MESH_SIZE, GEAR_DIMENSION)) + geom_point()

#Strong correlation of fishing gear dimension with the type of fishing gear (way higher for gillnets compared to trawls and seines), the targeted taxa (particularly high for benthic species as pleuronectiformes or Lophius spp) and the mesh size (globally increase with it)

kruskal.test(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$GEAR_DEPTH ~ OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$TARGETED_TAXA)
ggplot(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], aes(TARGETED_TAXA, GEAR_DEPTH)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
# as it could be expected gear depth is highly dependent of the targeted taxa habitat

# gear speed, gear, depth and gear dimension were removed from analyses due to too many missing values and/or too strong correlations with other explanatory variables

##### Correlations between measured traits:

cor.test(Strandings[SPECIES == "Delphinus delphis", BODY_LENGTH], Strandings[SPECIES == "Delphinus delphis", BODY_MASS], method = "spearman") 
cor.test(Strandings[SPECIES == "Phocoena phocoena", BODY_LENGTH], Strandings[SPECIES == "Phocoena phocoena", BODY_MASS], method = "spearman")

cor.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis", BODY_LENGTH], OBSMER_Bycatch[SPECIES == "Delphinus delphis", BODY_MASS], method = "spearman") 
cor.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena", BODY_LENGTH], OBSMER_Bycatch[SPECIES == "Phocoena phocoena", BODY_MASS], method = "spearman")

wilcox.test(Strandings[SPECIES == "Delphinus delphis", BODY_LENGTH] ~ Strandings[SPECIES == "Delphinus delphis", SEX]) 
ggplot(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX)], aes(SEX, BODY_LENGTH)) + geom_boxplot()
wilcox.test(Strandings[SPECIES == "Phocoena phocoena", BODY_LENGTH] ~ Strandings[SPECIES == "Phocoena phocoena", SEX])

wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis", BODY_LENGTH]~OBSMER_Bycatch[SPECIES == "Delphinus delphis", SEX]) 
wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena", BODY_LENGTH]~OBSMER_Bycatch[SPECIES == "Phocoena phocoena", SEX])

wilcox.test(Strandings[SPECIES == "Delphinus delphis", BODY_MASS]~Strandings[SPECIES == "Delphinus delphis", SEX]) 
ggplot(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX)], aes(SEX, BODY_MASS)) + geom_boxplot()
wilcox.test(Strandings[SPECIES == "Phocoena phocoena", BODY_MASS]~Strandings[SPECIES == "Phocoena phocoena", SEX])

wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis", BODY_MASS]~OBSMER_Bycatch[SPECIES == "Delphinus delphis", SEX]) 
wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena", BODY_MASS]~OBSMER_Bycatch[SPECIES == "Phocoena phocoena", SEX])

mean(Strandings[SPECIES == "Delphinus delphis" & SEX == "Male", BODY_LENGTH], na.rm = T)
sd(Strandings[SPECIES == "Delphinus delphis" & SEX == "Male", BODY_LENGTH], na.rm = T)
mean(Strandings[SPECIES == "Delphinus delphis" & SEX == "Female", BODY_LENGTH], na.rm = T)
sd(Strandings[SPECIES == "Delphinus delphis" & SEX == "Female", BODY_LENGTH], na.rm = T)

mean(Strandings[SPECIES == "Delphinus delphis" & SEX == "Male", BODY_MASS], na.rm = T)
sd(Strandings[SPECIES == "Delphinus delphis" & SEX == "Male", BODY_MASS], na.rm = T)
mean(Strandings[SPECIES == "Delphinus delphis" & SEX == "Female", BODY_MASS], na.rm = T)
sd(Strandings[SPECIES == "Delphinus delphis" & SEX == "Female", BODY_MASS], na.rm = T)

##### Distribution of captures as a function of external variables (spatial, temporal, environmental):

  ## Spatial distribution:

table( Decl_Obs_Merged_Bycatch_explanatory[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),]$SPECIES,  Decl_Obs_Merged_Bycatch_explanatory[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),]$ICES_DIV)

table(Strandings$SPECIES, Strandings$ICES_DIV)

# these values were then used to produce a map in QGIS, using map background provided by the ICES
# a map of abundance corrected by sampling effort (for observation data only) was also produced (see below)

# Maps of corrected abundance:

observation_effort <- readRDS("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/observation_effort.rds")

observation_effort_spat=tapply(observation_effort$DURATION_OP, observation_effort$ICES_DIV, sum) # sampling effort (OBSMER) per ICES division
observation_effort_spat=as.data.frame(observation_effort_spat)
observation_effort_spat$icesdiv=rownames(observation_effort_spat)

allgeo <- readRDS("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/allgeo_Journal_version.rds") # map of ICES division of interest

posi_bycatch = unlist(by(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$ICES_DIV, function(x) tapply(x$SPECIES, x$SPECIES, length))) # number of individuals by-caught per ICES division and species

posi_bycatch = as.data.frame(posi_bycatch)
posi_bycatch$icesdiv = str_sub(rownames(posi_bycatch), 1, 6)
posi_bycatch$Species = str_sub(rownames(posi_bycatch), 8, -1L)

posi_bycatch = merge(posi_bycatch, observation_effort_spat)
posi_bycatch$bycatch_rate = posi_bycatch$posi_bycatch / posi_bycatch$observation_effort_spat

posi_bycatch = merge(posi_bycatch, allgeo)

world_map = map_data("world")
rangex <- c(-18, 9)
rangey <- c(40, 55)

ggplot(st_as_sf(posi_bycatch)) +
  geom_sf(aes(fill = bycatch_rate)) +
  scale_fill_viridis_c(name = "By-catch rate\n(number of captured dolphin\nper minute of fishing activity\nobservation)")  +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", color = "black") + 
  coord_sf(rangex, rangey) + 
  facet_wrap(~ Species, labeller =  as_labeller(c('Delphinus delphis' = "Common dolphin", 'Phocoena phocoena' = "Harbour porpoise"))) +
  xlab("Longitude") + ylab("Latitude") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic"))

ggsave("Figure_S2.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

  ## Temporal distribution:

# Years:

ggplot(dcast(Strandings[!is.na(YEAR)], formula = YEAR + SPECIES ~ ., fun.aggregate = length), aes(YEAR, .)) +
  geom_bar(stat = "identity")  + 
  facet_wrap(~ SPECIES) +
  xlab("Year") + ylab("Number of stranded individuals") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Strandings per year")

ggsave("Figure_S3A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

ggplot(dcast(Decl_Obs_Merged_Bycatch_explanatory[!is.na(YEAR) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")],
             formula = YEAR + SPECIES ~ .,
             fun.aggregate = length),
       aes(YEAR, .)) + 
  geom_bar(stat = "identity")  + 
  facet_wrap( ~ SPECIES) + 
  xlab("Year") + ylab("Number of by-caught individuals") + 
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catches per year (declaration and observation data)")

ggsave("Figure_S3B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

observation_effort_year = tapply(observation_effort$DURATION_OP, str_sub(observation_effort$END_TIMING, 7, 10), sum) # sampling effort (OBSMER) per year
observation_effort_year = as.data.frame(observation_effort_year)
observation_effort_year$YEAR = as.integer(rownames(observation_effort_year))

bycatch_rate_year = merge(dcast(OBSMER_Bycatch[!is.na(YEAR) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = YEAR + SPECIES ~ ., fun.aggregate = length), observation_effort_year, by = "YEAR")

bycatch_rate_year$Bycatch_rate = bycatch_rate_year$. / bycatch_rate_year$observation_effort_year

ggplot(bycatch_rate_year, aes(YEAR, Bycatch_rate)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~ SPECIES)  +
  xlab("Year") + ylab("Number of by-caught individuals\nper minute of fishing activity") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catch rate per year (observation data)")

ggsave("Figure_S3C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

  ###Trimester:

ggplot(dcast(Strandings[!is.na(TRIMESTER)], formula = TRIMESTER + SPECIES ~ ., fun.aggregate = length), aes(TRIMESTER, .)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~ SPECIES)  +
  xlab("Trimester") + ylab("Number of stranded individuals") + 
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Strandings per trimester")

ggsave("Figure_S4A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

ggplot(dcast(Decl_Obs_Merged_Bycatch_explanatory[!is.na(TRIMESTER) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = TRIMESTER + SPECIES ~ ., fun.aggregate = length), aes(TRIMESTER, .)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~ SPECIES) +
  xlab("Trimester") + ylab("Number of by-caught individuals") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catches per trimester (declaration and observation data)")

ggsave("Figure_S4B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

observation_effort_trimester = tapply(observation_effort$DURATION_OP, ceiling(as.numeric(str_sub(observation_effort$END_TIMING,4,5))/3), sum) # sampling effort (OBSMER) per trimester
observation_effort_trimester = as.data.frame(observation_effort_trimester)
observation_effort_trimester$TRIMESTER = rownames(observation_effort_trimester)

bycatch_rate_trimester = merge(dcast(OBSMER_Bycatch[!is.na(TRIMESTER) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = TRIMESTER + SPECIES ~ ., fun.aggregate = length), observation_effort_trimester, by = "TRIMESTER")

bycatch_rate_trimester$Bycatch_rate = bycatch_rate_trimester$. / bycatch_rate_trimester$observation_effort_trimester

ggplot(bycatch_rate_trimester, aes(TRIMESTER, Bycatch_rate)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~ SPECIES)  +
  xlab("Trimester") + ylab("Number of by-caught individuals\nper minute of fishing activity") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catch rate per trimester (observation data)")

ggsave("Figure_S4C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

  ### Daytime :

ggplot(dcast(OBSMER_Bycatch[!is.na(TIMING_OP) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = TIMING_OP + SPECIES ~ ., fun.aggregate = length), aes(TIMING_OP, .)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~ SPECIES) +
  xlab("Daytime of hauling") + ylab("Number of by-caught individuals") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catches per hauling period (observation data)")  +
  scale_x_discrete(limit=c("Night (00:00 - 06:00)", "Morning (06:00 - 12:00)", "Afternoon (12:00 - 18:00)", "Evening (18:00 - 00:00)") ,labels=c("Night\n(00:00 - 06:00)", "Morning\n(06:00 - 12:00)", "Afternoon\n(12:00 - 18:00)", "Evening\n(18:00 - 00:00)"))

observation_effort_daytime = tapply(observation_effort$DURATION_OP, as.character(ceiling((hour(dmy_hms(observation_effort$END_TIMING)) +1)/6)), sum)  # sampling effort (OBSMER) per period of the day (6-hour periods)
observation_effort_daytime = as.data.frame(observation_effort_daytime)
observation_effort_daytime$TIMING_OP = rownames(observation_effort_daytime)

observation_effort_daytime[observation_effort_daytime$TIMING_OP == 1,]$TIMING_OP = "Night (00:00 - 06:00)"
observation_effort_daytime[observation_effort_daytime$TIMING_OP == 2,]$TIMING_OP = "Morning (06:00 - 12:00)"
observation_effort_daytime[observation_effort_daytime$TIMING_OP == 3,]$TIMING_OP = "Afternoon (12:00 - 18:00)"
observation_effort_daytime[observation_effort_daytime$TIMING_OP == 4,]$TIMING_OP = "Evening (18:00 - 00:00)"

bycatch_rate_daytime = merge(dcast(OBSMER_Bycatch[!is.na(TIMING_OP) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = TIMING_OP + SPECIES ~ ., fun.aggregate = length), observation_effort_daytime, by = "TIMING_OP")

bycatch_rate_daytime$Bycatch_rate = bycatch_rate_daytime$. / bycatch_rate_daytime$observation_effort_daytime

ggplot(bycatch_rate_daytime, aes(TIMING_OP, Bycatch_rate)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~SPECIES)  +
  xlab("Daytime of hauling") + ylab("Number of by-caught individuals\nper minute of fishing activity") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catch rate per hauling period (observation data)")  +
  scale_x_discrete(limit=c("Night (00:00 - 06:00)", "Morning (06:00 - 12:00)", "Afternoon (12:00 - 18:00)", "Evening (18:00 - 00:00)") ,labels=c("Night\n(00:00 - 06:00)", "Morning\n(06:00 - 12:00)", "Afternoon\n(12:00 - 18:00)", "Evening\n(18:00 - 00:00)"))

ggsave("Figure_S5.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# Sea state:

ggplot(dcast(OBSMER_Bycatch[SEA_STATE != "" & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = SEA_STATE + SPECIES ~ ., fun.aggregate = length), aes(SEA_STATE, .)) +
  geom_bar(stat = "identity")  +
  facet_wrap(~ SPECIES) +
  xlab("Sea condition when hauling (Douglas scale)") + ylab("Number of by-caught individuals") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catches per hauling condition (observation data)")

observation_effort_sea_state = tapply(observation_effort$DURATION_OP, observation_effort$SEA_STATE, sum)  # sampling effort (OBSMER) per sea state condition
observation_effort_sea_state = as.data.frame(observation_effort_sea_state)
observation_effort_sea_state$SEA_STATE = rownames(observation_effort_sea_state)

bycatch_rate_sea_state = merge(dcast(OBSMER_Bycatch[SEA_STATE != "" & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")], formula = SEA_STATE + SPECIES ~ ., fun.aggregate = length), observation_effort_sea_state, by = "SEA_STATE")

bycatch_rate_sea_state$Bycatch_rate = bycatch_rate_sea_state$. / bycatch_rate_sea_state$observation_effort_sea_state

ggplot(bycatch_rate_sea_state, aes(SEA_STATE, Bycatch_rate)) +
  geom_bar(stat = "identity") + 
  facet_wrap(~SPECIES) +
  xlab("Sea condition when hauling (Douglas scale)") + ylab("Number of by-caught individuals\nper minute of fishing activity") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("By-catch rate per hauling condition (observation data)")

ggsave("Figure_S6.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

```

# Phenotypic sensitivity to bycatch

We have separated the analysis of stranding data from the declaration and observation data because they may be partially redundant (with no way to verify this), and because their collection is fundamentally different (only some of the captured dolphins strand, whereas theoretically we should detect all captures through observation and reporting).

For each data sets, we first described the distribution of studied traits and compared the trait values of by-caught and stranded individuals with key population values reported in the literature for each of the species studied (for both sexes: body length at maturity, asymptotic adult body length). We then examined whether different phenotypic sensitivity profiles emerged in space or time by implementing a generalised linear model for each of the traits studied (linear model for body length, logistic regression for sex) as a function of year, trimester and ICES division of by-catch.

## Stranding data

### Distribution of traits and comparisons with key population values

```{r sensitivity stranding - distribution comparison}

# Creation of datasets per species:

data_strandings_PHO = Strandings[SPECIES == "Phocoena phocoena"]
data_strandings_DEL = Strandings[SPECIES == "Delphinus delphis"]

# sex-ratio of strandings:

ggplot(dcast(na.omit(Strandings[, .(SEX, SPECIES)]), SEX + SPECIES ~ ., fun.aggregate = length), aes( x = SPECIES, y = .)) +
  geom_bar(aes(fill = SEX), stat = "identity", position = "dodge")  +
  ggtitle("Sex distribution of strandings") + 
  labs(x = "Captured species", y = "Number of individuals", fill = "Sex") +
  scale_fill_grey() +  theme(text = element_text(size = 30))

binom.test(table(Strandings[, .(SEX, SPECIES)][SPECIES == "Delphinus delphis"][,SEX]))

binom.test(table(Strandings[, .(SEX, SPECIES)][SPECIES == "Phocoena phocoena"][,SEX]))

# strandings body mass distribution:

ggplot(Strandings[, .(BODY_MASS, SPECIES)], aes( x = SPECIES, y = BODY_MASS)) +
  geom_violin() + geom_boxplot(width = 0.1) +
  ggtitle("Body mass of stranded individuals") +
  labs(x = "Species", y = "Body mass (kg)") + 
  theme(text = element_text(size = 30))

    ### by sex:

ggplot(Strandings[!is.na(SEX), .(BODY_MASS, paste0(SPECIES, "\n(", SEX, ")"))], aes( x = V2, y = BODY_MASS)) +
  geom_violin() + geom_boxplot(width = 0.1) + 
  ggtitle("Body mass of stranded individuals") + 
  labs(x = "Species", y = "Body mass (kg)") + 
  theme(text = element_text(size = 30))

# strandings body length distribution:

biblio = data.frame(ref = c("Murphy et al. 2021", "Murphy et al. 2021", "Murphy and Rogan 2006", "Murphy and Rogan 2006", "Perrin and Reilly 1984", "Perrin and Reilly 1984", "Murphy et al. 2013", "Murphy et al. 2013","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003"),
                    param = c("Sexual maturity\n(males, Delphinus delphis)", "Sexual maturity\n(females, Delphinus delphis)", "Asymptotic length\n(males, Delphinus delphis)", "Asymptotic length\n(females, Delphinus delphis)", "Sexual maturity\n(males, Delphinus delphis)", "Sexual maturity\n(females, Delphinus delphis)", "Asymptotic length\n(males, Delphinus delphis)", "Asymptotic length\n(females, Delphinus delphis)","Sexual maturity\n(males, Phocoena phocoena)", "Sexual maturity\n(females, Phocoena phocoena)", "Asymptotic length\n(males, Phocoena phocoena)", "Asymptotic length\n(females, Phocoena phocoena)", "Sexual maturity\n(males, Phocoena phocoena)", "Sexual maturity\n(females, Phocoena phocoena)", "Asymptotic length\n(males, Phocoena phocoena)", "Asymptotic length\n(females, Phocoena phocoena)"),
                    VALEUR = c(206, 188, 211, 197, 200, 190, 209, 206, 127, 138, 141, 153, 135, 147, 149, 163))

# key population values reported in the literature for each of the species studied (for both sexes: body length at maturity, asymptotic adult body length)

ggplot(Strandings[, .(paste0(SPECIES, " (", SEX, ")"), BODY_LENGTH)], aes( x = V1, y = BODY_LENGTH)) +
  geom_violin(width = 1) + geom_boxplot(width = 0.1) +
  geom_point(data = biblio, mapping = aes(x = param, y = VALEUR, shape = ref), size = 2) +
  ggtitle("Body length of stranded individuals\n(compared with literature life-history traits)") +  
  labs(x = "Species, sex and literature estimates", y = "Body length (cm)", shape = "Bibliographic\nreferences") + 
  theme(text = element_text(size = 20))  +  theme(axis.text.x = element_text(angle = 55, hjust = 1))  +
  scale_x_discrete(limits = c("Delphinus delphis\n(Female)","Sexual maturity\n(females, Delphinus delphis)","Asymptotic length\n(females, Delphinus delphis)","Delphinus delphis\n(Male)","Sexual maturity\n(males, Delphinus delphis)","Asymptotic length\n(males, Delphinus delphis)", "Phocoena phocoena\n(Female)","Sexual maturity\n(females, Phocoena phocoena)","Asymptotic length\n(females, Phocoena phocoena)","Phocoena phocoena\n(Male)","Sexual maturity\n(males, Phocoena phocoena)","Asymptotic length\n(males, Phocoena phocoena)"))

wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=211)
wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=209)

wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=206)
wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=200)

# size significantly inferior to length at maturity and asymptotic length for male common dolphin (N=1821)

wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=206)
wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=197)

wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=190)
wilcox.test(Strandings[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=188)

# size significantly inferior to length at maturity and asymptotic length for female common dolphin (N=1348)

wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=149)
wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=141)

wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=135)
wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=127)

# size significantly inferior to length at maturity or adult asymptotic length for male harbour porpoise (N=466)

wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=163)
wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=153)

wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=147)
wilcox.test(Strandings[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=138)

# size significantly inferior to length at maturity or adult asymptotic length for female harbour porpoise (N=356)

   ### body mass and body length classes (proxy for age classes) distribution per sex and species:

ggplot(Strandings[!is.na(SEX) & ((SPECIES == "Delphinus delphis" & BODY_MASS %between% quantile(Strandings[!is.na(SEX) & SPECIES == "Delphinus delphis",BODY_MASS], c(0.025, 0.975), na.rm = T)) | (SPECIES == "Phocoena phocoena" & BODY_MASS %between% quantile(Strandings[!is.na(SEX) & SPECIES == "Phocoena phocoena" ,BODY_MASS], c(0.025, 0.975), na.rm = T)))], aes(x = BODY_MASS)) +
  geom_histogram (bins = 9, alpha = 0.5) + 
  facet_wrap(SEX~SPECIES, scales = "free_x") + 
    stat_summary(
        geom = "vline", linetype = "dashed", colour = "red",
        orientation = "y",
        # y is a required aesthetic, so use a dummy value
        aes(y = 1, xintercept = after_stat(x)),
        fun = function(x) {
            quantile(x, probs = c(0.25, 0.5, 0.75))
        }
    ) + ylab("Count") + xlab("Body mass (kg)") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) +
  ggtitle("Mass class distribution")

ggsave("Figure_S10A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

ggplot(Strandings[!is.na(SEX) & ((SPECIES == "Delphinus delphis" & BODY_LENGTH %between% quantile(Strandings[!is.na(SEX) & SPECIES == "Delphinus delphis",BODY_LENGTH], c(0.025, 0.975), na.rm = T)) | (SPECIES == "Phocoena phocoena" & BODY_LENGTH %between% quantile(Strandings[!is.na(SEX) & SPECIES == "Phocoena phocoena" ,BODY_LENGTH], c(0.025, 0.975), na.rm = T)))], aes(x = BODY_LENGTH)) +
  geom_histogram (bins = 9, alpha = 0.5) + 
  facet_wrap(SEX ~ SPECIES, scales = "free_x") + 
    stat_summary(
        geom = "vline", linetype = "dashed", colour = "red",
        orientation = "y",
        # y is a required aesthetic, so use a dummy value
        aes(y = 1, xintercept = after_stat(x)),
        fun = function(x) {
            quantile(x, probs = c(0.25, 0.5, 0.75))
        }
    ) +
  ylab("Count") + xlab("Body length (cm)") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) +
  ggtitle("Size class distribution")

ggsave("Figure_S10B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# the edges of the distribution (2.5th quantile and 97.5th quantile) were discarded to avoid including outliers when comparing distributions

```

### Spatio-temporal patterns of sensitivity

```{r sensitivity stranding - spatio-temporal variations}

# Spatio-temporal variables description:

table(is.na(Strandings$TRIMESTER))
table(is.na(Strandings$YEAR))
table(is.na(Strandings$ICES_DIV))
# Very rare missing values for spatio-temporal variables

table(data_strandings_DEL$YEAR)
table(data_strandings_PHO$YEAR)

table(data_strandings_DEL$TRIMESTER)
table(data_strandings_PHO$TRIMESTER)

table(data_strandings_DEL$ICES_DIV)
table(data_strandings_PHO$ICES_DIV)

# some years appears quite rarely and the number of categories associated with the variable is high

### we grouped year (4-year periods, except for the last three years) to facilitate interpretations and avoid cases with too few data:

data_strandings_DEL[YEAR %in% 2000:2003, YEAR_group:= "2000-2003"]
data_strandings_DEL[YEAR %in% 2004:2007, YEAR_group:= "2004-2007"]
data_strandings_DEL[YEAR %in% 2008:2011, YEAR_group:= "2008-2011"]
data_strandings_DEL[YEAR %in% 2012:2015, YEAR_group:= "2012-2015"]
data_strandings_DEL[YEAR %in% 2016:2019, YEAR_group:= "2016-2019"]
data_strandings_DEL[YEAR %in% 2020:2022, YEAR_group:= "2020-2022"]

data_strandings_DEL[,YEAR_group:=as.factor(YEAR_group)]

data_strandings_PHO[YEAR %in% 2000:2003, YEAR_group:= "2000-2003"]
data_strandings_PHO[YEAR %in% 2004:2007, YEAR_group:= "2004-2007"]
data_strandings_PHO[YEAR %in% 2008:2011, YEAR_group:= "2008-2011"]
data_strandings_PHO[YEAR %in% 2012:2015, YEAR_group:= "2012-2015"]
data_strandings_PHO[YEAR %in% 2016:2019, YEAR_group:= "2016-2019"]
data_strandings_PHO[YEAR %in% 2020:2022, YEAR_group:= "2020-2022"]

data_strandings_PHO[,YEAR_group:=as.factor(YEAR_group)]

table(data_strandings_DEL$YEAR_group, data_strandings_DEL$ICES_DIV)
table(data_strandings_PHO$YEAR_group, data_strandings_PHO$ICES_DIV)

table(data_strandings_DEL$YEAR_group, data_strandings_DEL$TRIMESTER)
table(data_strandings_PHO$YEAR_group, data_strandings_PHO$TRIMESTER)

table(data_strandings_DEL$TRIMESTER, data_strandings_DEL$ICES_DIV)
table(data_strandings_PHO$TRIMESTER, data_strandings_PHO$ICES_DIV) #some combinations are rare, preventing the use of spatio-temporal interactions in models (potentially biased)

### Common dolphin:

  # Sex:

data_strandings_DEL[, SEX := as.factor(SEX)]
data_strandings_DEL[, TRIMESTER := as.factor(TRIMESTER)]
data_strandings_DEL[, ICES_DIV := as.factor(ICES_DIV)]

SPT_sensitivity_DEL_sex <- glm(data = data_strandings_DEL, SEX ~ YEAR_group + ICES_DIV + TRIMESTER, family = "binomial")

vif(SPT_sensitivity_DEL_sex)
summary(SPT_sensitivity_DEL_sex)
Anova(SPT_sensitivity_DEL_sex)

plot(ggemmeans(SPT_sensitivity_DEL_sex, terms = "TRIMESTER")) +
  ggtitle("Predicted probabilities of by-caught individual's sex\nper trimester") +
  ylab("By-caught individual's sex (probability)") +
  theme(text = element_text(size = 30))

ggsave("Figure_2A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(SPT_sensitivity_DEL_sex, terms = "ICES_DIV")) +
  ggtitle("Predicted probabilities of by-caught individual's sex\nper ICES division") +
  ylab("By-caught individual's sex (probability)") +
  xlab("ICES division") + 
  theme(text = element_text(size = 30))

ggsave("Figure_2B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(SPT_sensitivity_DEL_sex, "TRIMESTER"))

pairs(emmeans(SPT_sensitivity_DEL_sex, "ICES_DIV"))
# post-hoc pairwise comparison tests

simulateResiduals(fittedModel = SPT_sensitivity_DEL_sex, plot = T) 
plot(cooks.distance(SPT_sensitivity_DEL_sex) )  
# diagnostic plots of model: OK

  # Size:

# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in this specific case

# Models on body size were performed for both sex separately to account for potential sex-specific spatio-temporal sensitivity profiles

    # Male:

SPT_sensitivity_DEL_sizeM <- lm(data = data_strandings_DEL[SEX == "Male"], BODY_LENGTH ~ YEAR_group + ICES_DIV + TRIMESTER)

vif(SPT_sensitivity_DEL_sizeM)
summary(SPT_sensitivity_DEL_sizeM)
Anova(SPT_sensitivity_DEL_sizeM)

plot(ggemmeans(SPT_sensitivity_DEL_sizeM, terms = "YEAR_group")) +
  ggtitle("Predicted body length of by-caught male\nper time period") +
  ylab("By-caught male's body length (cm)") + xlab("Time period (year interval)") +
  theme(text = element_text(size = 30))

ggsave("Figure_2C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(SPT_sensitivity_DEL_sizeM, terms = "TRIMESTER")) +
  ggtitle("Predicted body length of by-caught male\nper trimester") +
  ylab("By-caught male's body length (cm)") + xlab("Trimester") +
  theme(text = element_text(size = 30))

ggsave("Figure_2D.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(SPT_sensitivity_DEL_sizeM, terms = "ICES_DIV")) +
  ggtitle("Predicted body length of by-caught male\nper ICES division") +
  ylab("By-caught male's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

ggsave("Figure_2E.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(SPT_sensitivity_DEL_sizeM, "YEAR_group"))

pairs(emmeans(SPT_sensitivity_DEL_sizeM, "TRIMESTER"))

pairs(emmeans(SPT_sensitivity_DEL_sizeM, "ICES_DIV"))
# post-hoc pairwise comparison tests

plot(SPT_sensitivity_DEL_sizeM)
plot(cooks.distance(SPT_sensitivity_DEL_sizeM))
simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_DEL_sizeM, plot = T) 
# Diagnostic plots revealed deviations from statistical assumptions

hist(data_strandings_DEL[SEX == "Male"]$BODY_LENGTH) # negatively skewed distribution (same for residuals)
hist(data_strandings_DEL[SEX == "Male"]$BODY_LENGTH^3) # way better after cubic transformation

SPT_sensitivity_DEL_sizeM_cubed <- lm(data = data_strandings_DEL[SEX== "Male"], I(BODY_LENGTH^3) ~ YEAR_group + ICES_DIV + TRIMESTER)

summary(SPT_sensitivity_DEL_sizeM_cubed)
Anova(SPT_sensitivity_DEL_sizeM_cubed)

plot(ggemmeans(SPT_sensitivity_DEL_sizeM_cubed, terms = "YEAR_group")) +
  ggtitle("Predicted body length of by-caught male\nper time period") +
  ylab("By-caught male's body length (cm)") + xlab("Time period (year interval)") +
  theme(text = element_text(size = 30)) 
plot(ggemmeans(SPT_sensitivity_DEL_sizeM_cubed, terms = "TRIMESTER")) +
  ggtitle("Predicted body length of by-caught male\nper trimester") +
  ylab("By-caught male's body length (cm)") +
  theme(text = element_text(size = 30))
plot(ggemmeans(SPT_sensitivity_DEL_sizeM_cubed, terms = "ICES_DIV")) +
  ggtitle("Predicted body length of by-caught male\nper ICES division") +
  ylab("By-caught male's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

# output nearly identical than before transforming data

simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_DEL_sizeM_cubed, plot = T)
# good model diagnostic plots (still significant deviation for first quantile, but reasonable and combined adjusted quantile NS)

    # Females:

SPT_sensitivity_DEL_sizeF <- lm(data = data_strandings_DEL[SEX == "Female"], BODY_LENGTH ~ YEAR_group + ICES_DIV + TRIMESTER)

vif(SPT_sensitivity_DEL_sizeF)
summary(SPT_sensitivity_DEL_sizeF)
Anova(SPT_sensitivity_DEL_sizeF)

plot(ggemmeans(SPT_sensitivity_DEL_sizeF, terms = "ICES_DIV")) + 
  ggtitle("Predicted body length of by-caught female\nper ICES division") +
  ylab("By-caught female's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

ggsave("Figure_2F.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(SPT_sensitivity_DEL_sizeF, "ICES_DIV"))
# post-hoc pairwise comparison tests

plot(SPT_sensitivity_DEL_sizeF)
plot(cooks.distance(SPT_sensitivity_DEL_sizeF))
simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_DEL_sizeF, plot = T) 
# Diagnostic plots revealed deviations from statistical assumptions

hist(data_strandings_DEL[SEX == "Female"]$BODY_LENGTH) #negatively skewed distribution (same for residuals)
hist(data_strandings_DEL[SEX == "Female"]$BODY_LENGTH^3) #way better after cubic transformation

SPT_sensitivity_DEL_sizeF_cubed <- lm(data = data_strandings_DEL[SEX== "Female"], I(BODY_LENGTH^3) ~ YEAR_group + ICES_DIV + TRIMESTER)

summary(SPT_sensitivity_DEL_sizeF_cubed)
Anova(SPT_sensitivity_DEL_sizeF_cubed)

plot(ggemmeans(SPT_sensitivity_DEL_sizeF_cubed, terms = "ICES_DIV")) + 
  ggtitle("Predicted body length of by-caught female\nper ICES division") +
  ylab("By-caught female's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

# output nearly identical than before transforming data

simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_DEL_sizeF_cubed, plot = T)
# good model diagnostic plots

### Harbour porpoise:

  # Sex:

data_strandings_PHO[, SEX := as.factor(SEX)]
data_strandings_PHO[, TRIMESTER := as.factor(TRIMESTER)]
data_strandings_PHO[, ICES_DIV := as.factor(ICES_DIV)]

SPT_sensitivity_PHO_sex <- glm(data = data_strandings_PHO, SEX ~ YEAR_group + ICES_DIV + TRIMESTER, family = "binomial")

vif(SPT_sensitivity_PHO_sex)
summary(SPT_sensitivity_PHO_sex)
Anova(SPT_sensitivity_PHO_sex)

plot(ggemmeans(SPT_sensitivity_PHO_sex, terms = "ICES_DIV"))  +
  ggtitle("Predicted probabilities of by-caught individual's sex\nper ICES division") +
  ylab("By-caught individual's sex (probability)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

ggsave("Figure_3A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(SPT_sensitivity_PHO_sex, "ICES_DIV"))
# post-hoc pairwise comparison tests

simulateResiduals(fittedModel = SPT_sensitivity_PHO_sex, plot = T) 
plot(cooks.distance(SPT_sensitivity_PHO_sex) ) 
# diagnostic plot of models: OK

  # Size:

# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in this specific case

# Models on body size were performed for both sex separately to account for potential sex-specific spatio-temporal sensitivity profiles

    # Males:

SPT_sensitivity_PHO_sizeM <- lm(data = data_strandings_PHO[SEX == "Male"], BODY_LENGTH ~ YEAR_group + ICES_DIV + TRIMESTER)

vif(SPT_sensitivity_PHO_sizeM)
summary(SPT_sensitivity_PHO_sizeM)
Anova(SPT_sensitivity_PHO_sizeM)

plot(ggemmeans(SPT_sensitivity_PHO_sizeM, terms = "TRIMESTER"))  +
  ggtitle("Predicted body length of by-caught male\nper trimester") +
  ylab("By-caught male's body length (cm)") +  xlab("Trimester") +
  theme(text = element_text(size = 30))

ggsave("Figure_3B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(SPT_sensitivity_PHO_sizeM, terms = "ICES_DIV"))  + 
  ggtitle("Predicted body length of by-caught male\nper ICES division") +
  ylab("By-caught male's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

ggsave("Figure_3C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(SPT_sensitivity_PHO_sex, "TRIMESTER"))
pairs(emmeans(SPT_sensitivity_PHO_sex, "ICES_DIV"))
# post-hoc pairwise comparison tests

plot(SPT_sensitivity_PHO_sizeM)
cooks.distance(SPT_sensitivity_PHO_sizeM)
simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_PHO_sizeM, plot = T)

# Diagnostic plot nearly ok, with small deviation for third quantile

hist(data_strandings_PHO[SEX == "Male"]$BODY_LENGTH) # small positive skewness
hist(sqrt(data_strandings_PHO[SEX == "Male"]$BODY_LENGTH)) # way better after squared root transformation

SPT_sensitivity_PHO_sizeM_sqrt <- lm(data = data_strandings_PHO[SEX == "Male"], sqrt(BODY_LENGTH) ~ YEAR_group + ICES_DIV + TRIMESTER)

summary(SPT_sensitivity_PHO_sizeM_sqrt)
Anova(SPT_sensitivity_PHO_sizeM_sqrt)

plot(ggemmeans(SPT_sensitivity_PHO_sizeM_sqrt, terms = "TRIMESTER"))  +
  ggtitle("Predicted body length of by-caught male\nper trimester") +
  ylab("By-caught male's body length (cm)") +
  theme(text = element_text(size = 30)) 
plot(ggemmeans(SPT_sensitivity_PHO_sizeM_sqrt, terms = "ICES_DIV"))  +
  ggtitle("Predicted body length of by-caught male\nperICES division") + 
  ylab("By-caught male's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

# output nearly identical than before transforming data

simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_PHO_sizeM, plot = T) 
#diagnostic plots of model: OK (except for a small deviation of the third quantile)

    # Females:

SPT_sensitivity_PHO_sizeF <- lm(data = data_strandings_PHO[SEX == "Female"], BODY_LENGTH ~ YEAR_group + ICES_DIV + TRIMESTER)

vif(SPT_sensitivity_PHO_sizeF)
summary(SPT_sensitivity_PHO_sizeF)
Anova(SPT_sensitivity_PHO_sizeF)

plot(ggemmeans(SPT_sensitivity_PHO_sizeF, terms = "ICES_DIV"))  +
  ggtitle("Predicted body length of by-caught female\nper ICES division") +
  ylab("By-caught female's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))

ggsave("Figure_3D.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(SPT_sensitivity_PHO_sizeF, terms = "YEAR_group")) +
  ggtitle("Predicted body length of by-caught female\nper time period") +
  ylab("By-caught female's body length (cm)") + xlab("Time period (year interval)") +
  theme(text = element_text(size = 30))

ggsave("Figure_3E.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(SPT_sensitivity_PHO_sizeF, "YEAR_group"))
pairs(emmeans(SPT_sensitivity_PHO_sizeF, "ICES_DIV"))
# post-hoc pairwise comparison tests

plot(SPT_sensitivity_PHO_sizeF)
cooks.distance(SPT_sensitivity_PHO_sizeF)
simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_PHO_sizeF, plot = T) 
# Model diagnostic plot nearly ok, with small deviation for normal distribution

hist(data_strandings_PHO[SEX == "Female"]$BODY_LENGTH) # small positive skewness of distribution
hist(sqrt(data_strandings_PHO[SEX == "Female"]$BODY_LENGTH)) # way better after squared root transformation

SPT_sensitivity_PHO_sizeF_sqrt <- lm(data = data_strandings_PHO[SEX == "Female"], sqrt(BODY_LENGTH) ~ YEAR_group + ICES_DIV + TRIMESTER)

summary(SPT_sensitivity_PHO_sizeF_sqrt)
Anova(SPT_sensitivity_PHO_sizeF_sqrt)

plot(ggemmeans(SPT_sensitivity_PHO_sizeF, terms = "ICES_DIV"))  +
  ggtitle("Predicted body length of by-caught female\nper ICES division") +
  ylab("By-caught female's body length (cm)") + xlab("ICES division") +
  theme(text = element_text(size = 30))    
plot(ggemmeans(SPT_sensitivity_PHO_sizeF, terms = "YEAR_group")) +
  ggtitle("Predicted body length of by-caught female\nper time period") +
  ylab("By-caught female's body length (cm)") + xlab("Time period (year interval)") +
  theme(text = element_text(size = 30))

# output nearly identical than before transforming data

simulationOutput <- simulateResiduals(fittedModel = SPT_sensitivity_PHO_sizeF_sqrt, plot = T) 
# diagnostic plots of model: OK

```

#### Output statististic tables (spatio-temporal regressions)

```{r sensitivity stranding - spatio-temporal variations Tables}

### Table (S2) combining all models' analysis of deviance (type II) statistics:

TableS2 = cbind(Anova(SPT_sensitivity_DEL_sex), Anova(SPT_sensitivity_DEL_sizeM)[-4,-1], Anova(SPT_sensitivity_DEL_sizeM_cubed)[-4,-1], Anova(SPT_sensitivity_DEL_sizeF)[-4,-1], Anova(SPT_sensitivity_DEL_sizeF_cubed)[-4,-1], Anova(SPT_sensitivity_PHO_sex), Anova(SPT_sensitivity_PHO_sizeM)[-4,-1], Anova(SPT_sensitivity_PHO_sizeM_sqrt)[-4,-1], Anova(SPT_sensitivity_PHO_sizeF)[-4,-1], Anova(SPT_sensitivity_PHO_sizeF_sqrt)[-4,-1])

TableS2 = TableS2[, c(2,1,3,5,6,8,9,11,12,14,15,16,18,20,21,23,24,26,27,29,30)] # reordering statistics values in the same order (and removing duplicated columns)

TableS2 = cbind(rownames(TableS2), TableS2[,1], mapply(function(x, y) paste0(signif(x, digits = 3), " (", signif(y, digits = 3), ")"), TableS2[, seq(2, 21, 2)], TableS2[, seq(3, 21, 2)])) # combining statistics on a same line

TableS2 = cbind(TableS2[, 1:3], paste0(TableS2[, 4], " / ", TableS2[, 5]), paste0(TableS2[, 6], " / ", TableS2[, 7]), TableS2[, 8], paste0(TableS2[, 9], " / ", TableS2[, 10]), paste0(TableS2[, 11], " / ", TableS2[, 12])) #combining model before/after transformations on the same line

colnames(TableS2) = c("", "Df", "LR Chisq (P-value)", rep("F Value (P-value)", 2), "LR Chisq (P-value)", rep("F Value (P-value)", 2))

TableS2[, 1] = c("Time period (year interval)", "ICES division", "Trimester")

TableS2 = rbind(TableS2, c("Number of observations", "", SPT_sensitivity_DEL_sex$df.null, SPT_sensitivity_DEL_sizeM$df.residual+11, SPT_sensitivity_DEL_sizeF$df.residual+11, SPT_sensitivity_PHO_sex$df.null, SPT_sensitivity_PHO_sizeM$df.residual+11, SPT_sensitivity_PHO_sizeF$df.residual+11)) # adding a line containing sample size for each model

TableS2 = rbind(TableS2, c("R-squared", "", signif(rsq.n(SPT_sensitivity_DEL_sex), digits = 3), paste0(signif(summary(SPT_sensitivity_DEL_sizeM)$r.squared, digits = 3), " / ", signif(summary(SPT_sensitivity_DEL_sizeM_cubed)$r.squared, digits = 3)), paste0(signif(summary(SPT_sensitivity_DEL_sizeF)$r.squared, digits = 3), " / ", signif(summary(SPT_sensitivity_DEL_sizeF_cubed)$r.squared, digits = 3)), signif(rsq.n(SPT_sensitivity_PHO_sex), digits = 3), paste0(signif(summary(SPT_sensitivity_PHO_sizeM)$r.squared, digits = 3), " / ", signif(summary(SPT_sensitivity_PHO_sizeM_sqrt)$r.squared, digits = 3)), paste0(signif(summary(SPT_sensitivity_PHO_sizeF)$r.squared, digits = 3), " / ", signif(summary(SPT_sensitivity_PHO_sizeF_sqrt)$r.squared, digits = 3)))) # adding a line containing R-squared values for each model

TableS2 = rbind(colnames(TableS2), TableS2)

colnames(TableS2) = c("", "", "Sex (logistic regression)", "Males size (linear regression)", "Females size (linear regression)", "Sex (logistic regression)", "Males size (linear regression)", "Females size (linear regression)")

kable_classic(add_header_above(kbl(TableS2, format = "latex", vline= ""), c("", "", "Delphinus delphis" = 3 , "Phocoena phocoena" = 3)))
# Latex format for exporting the table

### Table combining pairwise contrast comparison test for each significant variable either in models studying sex distribution (S3) or size distribution (S4):

TableS3 = rbind( 
  cbind(Species = "Delphinus delphis", Variable = "Trimester", as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sex, "TRIMESTER"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sex, "TRIMESTER")))$p.value),
  cbind(Species = "Delphinus delphis", Variable = "ICES division", as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sex, "ICES_DIV"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sex, "ICES_DIV")))$p.value),
  cbind(Species = "Phocoena phocoena", Variable = "ICES division", as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sex, "ICES_DIV"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sex, "ICES_DIV")))$p.value)
  ) # statistics of pairwise comparison tests for all models on sex (with and without correction for multiple testing)

TableS3$df = NULL

TableS3[sapply(TableS3, is.numeric)] = lapply(TableS3[sapply(TableS3, is.numeric)], round, 3) # round all results to three decimal places

TableS3 = TableS3[TableS3$p.value < 0.05, ] # keeping only significant pairwise comparison tests

rownames(TableS3) = NULL

 kable_classic(collapse_rows(kbl(TableS3, format = "latex"), columns = 1:2, target = 1, valign = "middle"))
# Latex format for exporting the table

TableS4 = rbind( 
  cbind(Species = "Delphinus delphis", Sex = "Male", Variable = "Year interval", as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM, "YEAR_group"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM, "YEAR_group")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM_cubed, "YEAR_group")))$p.value),
  cbind(Species = "Delphinus delphis", Sex = "Male", Variable = "Trimester", as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM, "TRIMESTER"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM, "TRIMESTER")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM_cubed, "TRIMESTER")))$p.value),
  cbind(Species = "Delphinus delphis", Sex = "Male", Variable = "ICES division", as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM, "ICES_DIV"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM, "ICES_DIV")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeM_cubed, "ICES_DIV")))$p.value),
  cbind(Species = "Delphinus delphis", Sex = "Female", Variable = "ICES division", as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeF, "ICES_DIV"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeF, "ICES_DIV")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_DEL_sizeF_cubed, "ICES_DIV")))$p.value),  
  cbind(Species = "Phocoena phocoena", Sex = "Male", Variable = "Trimester", as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeM, "TRIMESTER"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeM, "TRIMESTER")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeM_sqrt, "TRIMESTER")))$p.value),
  cbind(Species = "Phocoena phocoena", Sex = "Male", Variable = "ICES division", as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeM, "ICES_DIV"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeM, "ICES_DIV")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeM_sqrt, "ICES_DIV")))$p.value),
  cbind(Species = "Phocoena phocoena", Sex = "Female", Variable = "Year interval", as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeF, "YEAR_group"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeF, "YEAR_group")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeF_sqrt, "YEAR_group")))$p.value),
  cbind(Species = "Phocoena phocoena", Sex = "Female", Variable = "ICES division", as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeF, "ICES_DIV"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeF, "ICES_DIV")))$p.value, adj.p.value_transf = as.data.frame(pairs(emmeans(SPT_sensitivity_PHO_sizeF_sqrt, "ICES_DIV")))$p.value)
) # statistics of pairwise comparison tests for all models on size (with and without correction for multiple testing)

TableS4$df = NULL

TableS4[sapply(TableS4, is.numeric)] = lapply(TableS4[sapply(TableS4, is.numeric)], round, 3) # round all results to three decimal places

TableS4=TableS4[TableS4$p.value < 0.05, ]  # keeping only significant pairwise comparison tests

rownames(TableS4) = NULL

 kable_classic(collapse_rows(kbl(TableS4, format = "latex"), columns = 1:3, target = 1, valign = "middle"))
# Latex format for exporting the table

```

## Observation and declaration data

### Distribution of traits and comparisons with key population values

```{r sensitivity observation and declaration- distribution comparison}

# sex-ratio of recorded incidental captures:

ggplot(dcast(na.omit(OBSMER_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(SEX, SPECIES)]), SEX + SPECIES ~ ., fun.aggregate = length), aes( x = SPECIES, y = .)) +
  geom_bar(aes(fill = SEX), stat = "identity", position = "dodge")  +
  ggtitle("Sex distribution of captures") +
  labs(x = "Captured species", y = "Number of individuals", fill = "Sex") + 
  scale_fill_grey() +  theme(text = element_text(size = 30))

binom.test(table(OBSMER_Bycatch[, .(SEX, SPECIES)][SPECIES == "Delphinus delphis"][,SEX]))

binom.test(table(OBSMER_Bycatch[, .(SEX, SPECIES)][SPECIES == "Phocoena phocoena"][,SEX]))

########## Combined figure with both strandings and recorded captures data:

ggplot(dcast(na.omit(rbind(OBSMER_Bycatch[!is.na(SEX) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(SEX, SPECIES, "Captures")], Strandings[!is.na(SEX), .(SEX, SPECIES, "Strandings")])), SEX + SPECIES + V3 ~ ., fun.aggregate = length), aes( x =V3, y = .)) +
  geom_bar(aes(fill = SEX), stat = "identity", position = "dodge") + 
  facet_wrap(~ SPECIES, scales = "free_y") +
  ggtitle("Sex distribution of by-catches/strandings") + 
  labs(x = "Species and data sources", y = "Number of individuals", fill = "Sex") + 
  scale_fill_grey() +  theme(text = element_text(size = 30), strip.text = element_text(face = "italic")) +  theme(axis.text.x = element_text(angle = 55, hjust = 1))

ggsave("Figure_1A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# body mass distribution of recorded incidental captures:

ggplot(Decl_Obs_Merged_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(BODY_MASS, SPECIES)], aes( x = SPECIES, y = BODY_MASS)) +
  geom_violin() + geom_boxplot(width = 0.1) + 
  ggtitle("Body mass of captured individuals") +  
  labs(x = "Species", y = "Body mass (kg)") +
  theme(text = element_text(size = 30))

    ### by sex:

ggplot(OBSMER_Bycatch[!is.na(SEX) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(BODY_MASS, paste0(SPECIES, " (", SEX, ")"))], aes( x = V2, y = BODY_MASS)) + 
  geom_violin() + geom_boxplot(width = 0.1) +
  ggtitle("Body mass of captured individuals") +  
  labs(x = "Species", y = "Body mass (kg)") +
  theme(text = element_text(size = 30))

########## Combined figure with both strandings and recorded captures data:

ggplot(rbind(Decl_Obs_Merged_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(BODY_MASS, SPECIES, "Captures")], Strandings[!is.na(SEX), .(BODY_MASS, SPECIES, paste0("Strandings,\n", SEX))]), aes( x = V3, y = BODY_MASS)) +
  geom_violin() + geom_boxplot(width = 0.1) +  
  ggtitle("Body mass of captured/stranded individuals") + 
  labs(x = "Species, data sources and sex", y = "Body mass (kg)") + 
  theme(text = element_text(size = 30), strip.text = element_text(face = "italic")) +  theme(axis.text.x = element_text(angle = 55, hjust = 1))  +
  facet_wrap(~ SPECIES)

ggsave("Figure_1B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# body length distribution of recorded incidental captures:

biblio = data.frame(ref = c("Murphy et al. 2021", "Murphy et al. 2021", "Murphy and Rogan 2006", "Murphy and Rogan 2006", "Perrin and Reilly 1984", "Perrin and Reilly 1984", "Murphy et al. 2013", "Murphy et al. 2013","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003","Loycker et al. 2003"),
                    param = c("Sexual maturity<br>(males, *Delphinus delphis*)", "Sexual maturity<br>(females, *Delphinus delphis*)", "Asymptotic length<br>(males, *Delphinus delphis*)", "Asymptotic length<br>(females, *Delphinus delphis*)", "Sexual maturity<br>(males, *Delphinus delphis*)", "Sexual maturity<br>(females, *Delphinus delphis*)", "Asymptotic length<br>(males, *Delphinus delphis*)", "Asymptotic length<br>(females, *Delphinus delphis*)","Sexual maturity<br>(males, *Phocoena phocoena*)", "Sexual maturity<br>(females, *Phocoena phocoena*)", "Asymptotic length<br>(males, *Phocoena phocoena*)", "Asymptotic length<br>(females, *Phocoena phocoena*)", "Sexual maturity<br>(males, *Phocoena phocoena*)", "Sexual maturity<br>(females, *Phocoena phocoena*)", "Asymptotic length<br>(males, *Phocoena phocoena*)", "Asymptotic length<br>(females, *Phocoena phocoena*)"), 
                    VALEUR = c(206, 188, 211, 197, 200, 190, 209, 206, 127, 138, 141, 153, 135, 147, 149, 163))

ggplot(OBSMER_Bycatch[, .(paste0("*", SPECIES, "*", " (", SEX, ")"), BODY_LENGTH)], aes( x = V1, y = BODY_LENGTH)) + 
  geom_violin(width = 1) + geom_boxplot(width = 0.1) + geom_point(data = biblio, mapping = aes(x = param, y = VALEUR, shape = ref), size = 2) + 
  ggtitle("Body length of stranded individuals\n(compared with literature life-history traits)") + 
  labs(x = "Species, sex and literature estimates", y = "Body length (cm)", shape= "Bibliographic\nreferences") +
  theme(text = element_text(size = 15))  +  theme(axis.text.x = element_text(angle = 60, hjust = 1))  +
  scale_x_discrete(limits = c("*Delphinus delphis* (Female)","Sexual maturity<br>(females, *Delphinus delphis*)","Asymptotic length<br>(females, *Delphinus delphis*)","*Delphinus delphis* (Male)","Sexual maturity<br>(males, *Delphinus delphis*)","Asymptotic length<br>(males, *Delphinus delphis*)", "*Phocoena phocoena* (Female)","Sexual maturity<br>(females, *Phocoena phocoena*)","Asymptotic length<br>(females, *Phocoena phocoena*)","*Phocoena phocoena* (Male)","Sexual maturity<br>(males, *Phocoena phocoena*)","Asymptotic length<br>(males, *Phocoena phocoena*)"))  +
    theme(axis.text = element_markdown())

########## Combined figure with both strandings and recorded captures data:

ggplot(rbind(OBSMER_Bycatch[, .(paste0("*", SPECIES, "*", ", ", SEX, "<br>Captures"), BODY_LENGTH)], Strandings[, .(paste0("*", SPECIES, "*", ", ", SEX, "<br>Strandings"), BODY_LENGTH)]), aes( x = V1, y = BODY_LENGTH)) + 
  geom_violin(width = 1) + geom_boxplot(width = 0.1) + geom_point(data=biblio, mapping=aes(x = param, y = VALEUR, shape=ref), size=2) + 
  ggtitle("Body length of captured/stranded individuals\n(compared with literature life-history traits)") +
  labs(x = "Species, sex, data sources and literature estimates", y = "Body length (cm)", shape= "Bibliographic\nreferences") +  
  theme(text = element_text(size = 15))  +  theme(axis.text.x = element_text(angle = 60, hjust = 1))  + 
  scale_x_discrete(limits = c("*Delphinus delphis*, Female<br>Strandings","*Delphinus delphis*, Female<br>Captures","Sexual maturity<br>(females, *Delphinus delphis*)","Asymptotic length<br>(females, *Delphinus delphis*)","*Delphinus delphis*, Male<br>Strandings","*Delphinus delphis*, Male<br>Captures","Sexual maturity<br>(males, *Delphinus delphis*)","Asymptotic length<br>(males, *Delphinus delphis*)", "*Phocoena phocoena*, Female<br>Strandings","*Phocoena phocoena*, Female<br>Captures","Sexual maturity<br>(females, *Phocoena phocoena*)","Asymptotic length<br>(females, *Phocoena phocoena*)","*Phocoena phocoena*, Male<br>Strandings","*Phocoena phocoena*, Male<br>Captures","Sexual maturity<br>(males, *Phocoena phocoena*)","Asymptotic length<br>(males, *Phocoena phocoena*)")) +
    theme(axis.text = element_markdown())

ggsave("Figure_1C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=211)
wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=209)

wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=206)
wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Male", BODY_LENGTH],mu=200)

# size significantly inferior to length at maturity and asymptotic length for male common dolphin (N=79)

wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=206)
wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=197)

wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=190)
wilcox.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & SEX== "Female", BODY_LENGTH],mu=188)

# size significantly inferior to length at maturity and asymptotic length for female common dolphin (N=79)

wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=149)
wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=141)

wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=135)
wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Male", BODY_LENGTH],mu=127)

# size significantly superior to length at maturity  for male harbour porpoise (N=18)

wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=163)
wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=153)

wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=147)
wilcox.test(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & SEX== "Female", BODY_LENGTH],mu=138)
# no significant differences (N=13)

   ### body length classes (proxy for age classes) distribution per sex and species:

ggplot(OBSMER_Bycatch[!is.na(SEX) & ((SPECIES == "Delphinus delphis" & BODY_LENGTH %between% quantile(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis", BODY_LENGTH], c(0.025, 0.975), na.rm = T)) | (SPECIES == "Phocoena phocoena" & BODY_LENGTH %between% quantile(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena" ,BODY_LENGTH], c(0.025, 0.975), na.rm = T)))], aes(x = BODY_LENGTH)) + geom_histogram (bins = 5, alpha = 0.5) + facet_wrap(SEX~SPECIES, scales = "free_x") + 
    stat_summary(
        geom = "vline", linetype = "dashed", colour = "red", 
        orientation = "y",
        # y is a required aesthetic, so use a dummy value
        aes(y = 1, xintercept = after_stat(x)),
        fun = function(x) {
            quantile(x, probs = c(0.25, 0.5, 0.75))
        }
    ) + ylab("Count") + xlab("Body length (cm)") +
  theme(text = element_text(size = 20), strip.text = element_text(face = "italic")) +
  ggtitle("Size class distribution")

ggsave("Figure_S11.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# we have not produced the same graph for body mass classes because the declaration data do not record sex, so the sample size with only observations is too small to produce such a graph

```

### Spatio-temporal patterns of sensitivity

```{r sensitivity observation and declaration - spatio-temporal variations}

# Spatio-temporal variables description:

anyNA(OBSMER_Bycatch$TRIMESTER)
anyNA(OBSMER_Bycatch$YEAR)
anyNA(OBSMER_Bycatch$ICES_DIV)

anyNA(Decl_Obs_Merged_Bycatch$TRIMESTER)
anyNA(Decl_Obs_Merged_Bycatch$YEAR)
anyNA(Decl_Obs_Merged_Bycatch$ICES_DIV)
# No missing values for spatio-temporal variables

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis"]$TRIMESTER)
table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$TRIMESTER)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$TRIMESTER)
table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$TRIMESTER)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV)
table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV)
table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV)

# The distribution of the number of captures per ICES division is highly variable, with some underrepresented divisions. Due to rare categories and the fact that the sample size of recorded captures is too small to support analyses with categorical variables described by a high number of classes, we decided to group years into four different groups.

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis"]$YEAR)
table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR)
table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR)

# The distribution of the number of captures per year is highly variable, with some underrepresented years, and differs greatly between common dolphins and harbour porpoises. Due to rare categories and the fact that the sample size of recorded captures is too small to support analyses with categorical variables described by a high number of classes, we decided to group years into four different groups. We used different groupings for each species to better account for temporal differences in the distribution of captures.

OBSMER_Bycatch$ICES_DIV_grouped=OBSMER_Bycatch$ICES_DIV
OBSMER_Bycatch[ICES_DIV %in% c("27.4.c", "27.7.e", "27.7.d"), ICES_DIV_grouped:= "27.7.de/4.c"]
OBSMER_Bycatch[ICES_DIV %in% c("27.7.f", "27.7.g", "27.7.h","27.7.j","27.7.k"), ICES_DIV_grouped:= "27.7.fghjk"]
OBSMER_Bycatch[ICES_DIV %in% c("27.8.a", "27.8.d"), ICES_DIV_grouped:= "27.8.ad"]
OBSMER_Bycatch[ICES_DIV %in% c("27.8.b", "27.8.c"), ICES_DIV_grouped:= "27.8.bc"]

Decl_Obs_Merged_Bycatch$ICES_DIV_grouped=Decl_Obs_Merged_Bycatch$ICES_DIV
Decl_Obs_Merged_Bycatch[ICES_DIV %in% c("27.4.c", "27.7.e", "27.7.d"), ICES_DIV_grouped:= "27.7.de/4.c"]
Decl_Obs_Merged_Bycatch[ICES_DIV %in% c("27.7.f", "27.7.g", "27.7.h","27.7.j","27.7.k"), ICES_DIV_grouped:= "27.7.fghjk"]
Decl_Obs_Merged_Bycatch[ICES_DIV %in% c("27.8.a", "27.8.d"), ICES_DIV_grouped:= "27.8.ad"]
Decl_Obs_Merged_Bycatch[ICES_DIV %in% c("27.8.b", "27.8.c"), ICES_DIV_grouped:= "27.8.bc"]

OBSMER_Bycatch$YEAR = as.character(OBSMER_Bycatch$YEAR)

OBSMER_Bycatch$YEAR_DCO=OBSMER_Bycatch$YEAR

OBSMER_Bycatch[YEAR %in% c("2005", "2007", "2008", "2009", "2010","2011", "2012"), YEAR_DCO:= "2005-2012"]
OBSMER_Bycatch[YEAR %in% c("2013", "2014", "2015", "2016"), YEAR_DCO:= "2013-2016"]
OBSMER_Bycatch[YEAR %in% c("2017", "2018", "2019", "2020"), YEAR_DCO:= "2017-2020"]
OBSMER_Bycatch[YEAR %in% c("2021", "2022","2023"), YEAR_DCO:= "2021-2023"]

OBSMER_Bycatch$YEAR_PHO=OBSMER_Bycatch$YEAR

OBSMER_Bycatch[YEAR %in% c("2005", "2007", "2008", "2009", "2010"), YEAR_PHO:= "2005-2010"]
OBSMER_Bycatch[YEAR %in% c("2011", "2012", "2013", "2014"), YEAR_PHO:= "2011-2014"]
OBSMER_Bycatch[YEAR %in% c("2015", "2016", "2017", "2018"), YEAR_PHO:= "2015-2018"]
OBSMER_Bycatch[YEAR %in% c("2019", "2020", "2021", "2022","2023"), YEAR_PHO:= "2019-2023"]

Decl_Obs_Merged_Bycatch$YEAR = as.character(Decl_Obs_Merged_Bycatch$YEAR)

Decl_Obs_Merged_Bycatch$YEAR_DCO=Decl_Obs_Merged_Bycatch$YEAR

Decl_Obs_Merged_Bycatch[YEAR %in% c("2005", "2007", "2008", "2009", "2010","2011", "2012"), YEAR_DCO:= "2005-2012"]
Decl_Obs_Merged_Bycatch[YEAR %in% c("2013", "2014", "2015", "2016"), YEAR_DCO:= "2013-2016"]
Decl_Obs_Merged_Bycatch[YEAR %in% c("2017", "2018", "2019", "2020"), YEAR_DCO:= "2017-2020"]
Decl_Obs_Merged_Bycatch[YEAR %in% c("2021", "2022","2023"), YEAR_DCO:= "2021-2023"]

Decl_Obs_Merged_Bycatch$YEAR_PHO=Decl_Obs_Merged_Bycatch$YEAR

Decl_Obs_Merged_Bycatch[YEAR %in% c("2005", "2007", "2008", "2009", "2010"), YEAR_PHO:= "2005-2010"]
Decl_Obs_Merged_Bycatch[YEAR %in% c("2011", "2012", "2013", "2014"), YEAR_PHO:= "2011-2014"]
Decl_Obs_Merged_Bycatch[YEAR %in% c("2015", "2016", "2017", "2018"), YEAR_PHO:= "2015-2018"]
Decl_Obs_Merged_Bycatch[YEAR %in% c("2019", "2020", "2021", "2022","2023"), YEAR_PHO:= "2019-2023"]

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)
table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_PHO)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)
table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_PHO)

# These groupings minimise the variance between sample sizes per group without creating groups with too few catches (N<6) and keeping consistent homogeneity (consecutive years and adjacent divisions with no major discontinuity).

# Exploration of sample size per analyses

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$SPECIES)
table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$SPECIES)
table(Decl_Obs_Merged_Bycatch[!is.na(BODY_MASS) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$SPECIES)

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$SPECIES, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena")]$SEX)

# Spatio-temporal variables used for models are composed of four categories each, regression models (see stranding analyses for details) parameters are then associated with ten degrees of freedom. If we consider that each explanatory parameter should be associated at least to 10 points of data to avoid over-parametrized models, we need at least 100 data points per model. This is not the case for all the analyses of recorded harbour porpoise captures, but also for all the models on body length of captured individuals by species and sex.

# To avoid over-parametrized models and harmonize analyses we only performed exploratory analyses on recorded captures using non parametric tests: Fisher tests for sex analyses and Kruskal-Wallis tests for body mass and length analyses, without considering sex separately for the latter analyses and with pairwise comparison tests if significant (respectively Fisher and Wilcoxon test)

# Common dolphin:

  # Sex:

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped) # more female capture north of the bay of Biscay (not significant)

ggplot(dcast(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"], SEX + ICES_DIV_grouped ~ ., fun.aggregate = length), aes( x = ICES_DIV_grouped, y = .)) +
  geom_bar(aes(fill = SEX), stat = "identity", position = "dodge") + ggtitle("Common dolphin - Sex model - OBSMER dataset") +  labs(x = "Area (ICES divisions groups)", y = "Number of captured individuals", fill = "Sex") + scale_fill_grey(label = c("Female", "Male")) +  theme(text = element_text(size = 30))  + scale_y_continuous(limits = c(0, 41))

ggsave("Figure_S13.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

fisher.test(table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$YEAR_DCO, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped), simulate.p.value = T) #strong correlation making interpretation difficult

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$YEAR_DCO)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$YEAR_DCO)

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$TRIMESTER)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$TRIMESTER)

  # weight:

# Prior to spatio-temporal test we checked whether body mass variation depended on the data sources:

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ORIGIN)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ORIGIN)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)

boxplot(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO) #larger individuals during recent years, particularly between 2017 and 2023

ggplot(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_MASS)], aes(BODY_MASS, YEAR_DCO)) + geom_violin() + geom_boxplot(width = 0.1) +  stat_compare_means(label = "p.format", comparisons = list(c("2017-2020", "2013-2016"), c("2021-2023", "2013-2016")), size=9) + ggtitle("Delphinus delphis - BM model - Merged datasets") +  labs(y = "Time period (years)", x = "Captured individuals' body mass (kg)") +  theme(text = element_text(size = 30))

ggsave("Figure_S12A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

fisher.test(table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO, Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped), simulate.p.value = T) #strong correlation making interpretation difficult

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$TRIMESTER)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$TRIMESTER)

  # size:

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$YEAR_DCO)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$YEAR_DCO)

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$TRIMESTER)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$TRIMESTER) 
boxplot(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"]$TRIMESTER) # smaller individuals captured during spring

ggplot(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis"], aes(BODY_LENGTH, TRIMESTER)) + geom_violin() + geom_boxplot(width = 0.1) +  stat_compare_means(label = "p.format", comparisons = list(c("2", "1"),c("2", "3"),c("2", "4")), size=9) + ggtitle("Common dolphin - BL model - OBSMER dataset") +  labs(y = "Trimester", x = "Captured individuals' body length (cm)") +  theme(text = element_text(size = 30))

ggsave("Figure_S12B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# Harbour porpoise:

  # Sex:

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$YEAR_DCO)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$YEAR_DCO)

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$TRIMESTER)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena"]$TRIMESTER)

  # weight:

# Prior to spatio-temporal test we checked whether body mass variation depended on the data sources:

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ORIGIN)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ORIGIN)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_DCO)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_DCO)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$TRIMESTER)
kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$TRIMESTER)

  # size:

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$YEAR_DCO)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$YEAR_DCO)

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$TRIMESTER)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena"]$TRIMESTER)

#### for by-catch observation only we also explored whether others external parameters (sea state and hauling timing during the day) correlated with by-caught individuals' traits:

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis"]$SEA_STATE)
table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$SEA_STATE)
# as their is some sea state categories with too little data (highest and lowest sea state measured) we grouped the most extreme conditions together for analyses:

OBSMER_Bycatch$SEA_STATE_grouped=OBSMER_Bycatch$SEA_STATE
OBSMER_Bycatch[SEA_STATE== "", SEA_STATE_grouped:=NA]
OBSMER_Bycatch[SEA_STATE %in% c("0 - Calm (glassy)", "1 - Calm (rippled)"), SEA_STATE_grouped:= "0, 1 - Calm"]
OBSMER_Bycatch[SEA_STATE %in% c("5 - Rough", "6 - Very rough"), SEA_STATE_grouped:= "5, 6 - Rough to very rough"]

# Common dolphin:

  # Sex:

fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped)
table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped) #more male than females captured when calm sea, converse dynamic when agitated sea

ggplot(dcast(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",], SEX + SEA_STATE_grouped ~ ., fun.aggregate = length), aes( x = ., y = SEA_STATE_grouped)) +
  geom_bar(aes(fill = SEX), stat = "identity", position = "dodge") +
  geom_signif(comparisons = list(c("0, 1 - Calm","4 - Moderate"), c("0, 1 - Calm","5, 6 - Rough to very rough")), test = "fisher.test2", y_position = c(35, 41), textsize = 9)  +
  ggtitle("Common dolphin - Sex model - OBSMER dataset") + 
  labs(y = "Sea conditions (Douglas scale)", x = "Number of captured individuals", fill = "Sex") +
  scale_fill_grey(label = c("Female", "Male")) +  theme(text = element_text(size = 30))   +
  scale_x_continuous(limits = c(0,46)) +
  scale_y_discrete(labels=c("Calm (0-1)", "Smooth (2)", "Slight (3)", "Moderate (4)", "Rough (>5)"))

ggsave("Figure_S16A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$TRIMESTER, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, simulate.p.value = T) 
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$ICES_DIV_grouped, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, simulate.p.value = T) 
# To be interpreted cautiously due to existing intercollinearity

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$TIMING_OP)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$TIMING_OP)

  # Body size:

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped)
boxplot(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped) #higher body length captured in calm sea (compared to low agitated sea)

ggplot(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",], aes(BODY_LENGTH, SEA_STATE_grouped)) + geom_violin() + geom_boxplot(width = 0.1) +  stat_compare_means(label = "p.format", comparisons = list(c("2 - belle, vagues de 0.1 à 0.5 mètres", "Calme ou ridée"), c("3 - peu agitée, vagues de 0.5 à 1.25 mètres", "Calme ou ridée"), c("4 - agitée, vagues de 1.25 à 2.5 mètres", "2 - belle, vagues de 0.1 à 0.5 mètres"), c("Forte à grosse", "Calme ou ridée")), size=9) + ggtitle("Common dolphin - BL model - OBSMER dataset") +  labs(y = "Sea conditions (Douglas scale)", x = "Captured individuals' body length (cm)") +  theme(text = element_text(size = 30)) + scale_y_discrete(labels=c("Calm (0-1)", "Smooth (2)", "Slight (3)", "Moderate (4)", "Rough (>5)"))

 ggsave("Figure_S16B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

fisher.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TRIMESTER, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, simulate.p.value = T) 
fisher.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$ICES_DIV_grouped, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, simulate.p.value = T) 
# To be interpreted cautiously due to existing intercollinearity

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TIMING_OP)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TIMING_OP)

# Harbour porpoise:

  # Sex:

fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEA_STATE_grouped)
table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEA_STATE_grouped)

table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$TIMING_OP)
fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$TIMING_OP)

  # Body size:

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$SEA_STATE_grouped)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$SEA_STATE_grouped)

table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$TIMING_OP)
kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$TIMING_OP)

```

#### Output statististic tables (spatio-temporal regressions)

```{r sensitivity by-catch - spatio-temporal variations Tables}

### Table (S5) combining all statistics of non-parametric tests on phenotypic sensitivity profiles:

TableS5 = cbind(
  "Variable" = c("ICES division", "Trimester", "Time period (year interval)", "Data source"),
  "Df" = c(3, 3, 3, 1),
  "P-value" = c(
    round(fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$ICES_DIV_grouped)$p.value, 3),
    round(fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$TRIMESTER)$p.value, 3),
    round(fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$YEAR_DCO)$p.value, 3),
    ""
  ),
  "Chi-squared (p-value)" = c(
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)$p.value, 3), ")"),
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$TRIMESTER)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$TRIMESTER)$p.value, 3), ")"),
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)$p.value, 3), ")"),
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ORIGIN)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$ORIGIN)$p.value, 3), ")")
  ),
  "Chi-squared (p-value)" = c(
    paste0(round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$ICES_DIV_grouped)$statistic, 3), " (", round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$ICES_DIV_grouped)$p.value, 3), ")"),
    paste0(round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TRIMESTER)$statistic, 3), " (", round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TRIMESTER)$p.value, 3), ")"),
    paste0(round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$YEAR_DCO)$statistic, 3), " (", round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$YEAR_DCO)$p.value, 3), ")"),
    ""
  ), 
    "P-value" = c(
    round(fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$ICES_DIV_grouped)$p.value, 3),
    round(fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$TRIMESTER)$p.value, 3),
    round(fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$YEAR_PHO)$p.value, 3),
    ""
  ),
  "Chi-squared (p-value)" = c(
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ICES_DIV_grouped)$p.value, 3), ")"),
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$TRIMESTER)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$TRIMESTER)$p.value, 3), ")"),
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_PHO)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_PHO)$p.value, 3), ")"),
    paste0(round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ORIGIN)$statistic, 3), " (", round(kruskal.test(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS ~ Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$ORIGIN)$p.value, 3), ")")
  ),
  "Chi-squared (p-value)" = c(
    paste0(round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$ICES_DIV_grouped)$statistic, 3), " (", round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$ICES_DIV_grouped)$p.value, 3), ")"),
    paste0(round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$TRIMESTER)$statistic, 3), " (", round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$TRIMESTER)$p.value, 3), ")"),
    paste0(round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$YEAR_PHO)$statistic, 3), " (", round(kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$YEAR_PHO)$p.value, 3), ")"), 
    ""
  )
)

TableS5=rbind(TableS5, 
              c("Number of observations",
                "",
                sum(table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis",]$YEAR_DCO)),
                sum(table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$BODY_MASS, Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis"]$YEAR_DCO)),
                sum(table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$YEAR_DCO)),
                sum(table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Phocoena phocoena",]$YEAR_PHO)),
                sum(table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$BODY_MASS, Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena"]$YEAR_PHO)),
                sum(table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$BODY_LENGTH, OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Phocoena phocoena",]$YEAR_PHO)) ))

TableS5=rbind(colnames(TableS5), TableS5)

colnames(TableS5)=c("", "", "Sex (Fisher test)", "BM (Kruskal-Wallis test)", "BL (Kruskal-Wallis test)", "Sex (Fisher test)", "BM (Kruskal-Wallis test)", "BL (Kruskal-Wallis test)")

kable_classic(add_header_above(kbl(TableS5, format = "latex", vline= ""), c("", "", "Common dolphin" = 3 , "Harbour porpoise" = 3)))

```

# Susceptibility to be captured

We then explored the susceptibility of individuals to be captured as a function of fishing technique. These analyses initially considered only the risk of being captured (see first graphical outputs, this part was scanned quickly as it had already been thoroughly studied) and how this susceptibility varies with individual traits (i.e. phenotypic vulnerabilities to fishing technique, see subsequent regression models).

## Bycatch distribution as a function of fishing techniques

```{r susceptibility}

# We first considered only the risk of being captured as a function of fishing technique. This part was scanned quickly as it had already been thoroughly studied by other researches. We only produced graphical outputs illustrating the relationships between fishing behaviours and the bumber of captured individuals.

# In the original dataset (combining both by-catch declarations and observations) we grouped some categories to facilitate interpretation of the distribution results (for targeted or most fished taxa, used groups were based on ISSCAAP classification from the FAO):

Decl_Obs_Merged_Bycatch_explanatory[
  FISHING_GEAR %in% c("GNS - Set gillnets", "GTN - Combined gillnets-trammel nets", "GTR - Trammel nets"), FISHING_GEAR_grouped := "Gillnets"][
    FISHING_GEAR %in% c("PS - Purse seines", "SDN - Danish seine"), FISHING_GEAR_grouped := "Seines"][
      FISHING_GEAR == "LLS - Set longlines", FISHING_GEAR_grouped := "Set longlines"][
        FISHING_GEAR %in% c("OTB - Single boat bottom otter trawls", "OTT - Twin bottom otter trawls", "PTB - Bottom pair trawls"), FISHING_GEAR_grouped := "Bottom trawls"][
          FISHING_GEAR %in% c("OTM - Single boat midwater otter trawls", "PTM - Midwater pair trawls"), FISHING_GEAR_grouped := "Midwater trawls"]

table(Decl_Obs_Merged_Bycatch_explanatory$FISHING_GEAR_grouped, Decl_Obs_Merged_Bycatch_explanatory$SPECIES)

table(Decl_Obs_Merged_Bycatch_explanatory$TARGETED_ISSCAAP, Decl_Obs_Merged_Bycatch_explanatory$SPECIES)

table(OBSMER_Bycatch$MOST_FISHED_ISSCAAP, OBSMER_Bycatch$SPECIES)

# Distribution of fishing gears associated with by-catch (Figure S7):

ggplot(Decl_Obs_Merged_Bycatch_explanatory[!is.na(FISHING_GEAR) & SPECIES == "Delphinus delphis"], aes(x = forcats::fct_infreq(FISHING_GEAR), fill = forcats::fct_infreq(FISHING_GEAR_grouped))) +
  geom_bar()  +
  labs(y = "Number of captures", x = "Fishing gear", fill = "Main groups") +
  ggtitle("Common dolphin") +
  theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_fill_discrete(labels = c("Gillnets (N=186)", "Midwater trawls (N=143)", "Bottom trawls (N=32)", "Seines (N=3)", "Set longlines (N=1)"))

ggsave("Figure_S7A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

ggplot(Decl_Obs_Merged_Bycatch_explanatory[!is.na(FISHING_GEAR) & SPECIES == "Phocoena phocoena"], aes(x = forcats::fct_infreq(FISHING_GEAR), fill = forcats::fct_infreq(FISHING_GEAR_grouped))) +
  geom_bar()  +
  labs(y = "Number of captures", x = "Fishing gear", fill = "Main groups") +
  ggtitle("Harbour porpoise") +
  theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_fill_discrete(labels = c("Gillnets (N=103)", "Midwater trawls (N=3)", "Bottom trawls (N=5)", "Seines (N=1)"))

ggsave("Figure_S7B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# Distribution of targeted taxa associated with by-catch (Figure S8):

ggplot(Decl_Obs_Merged_Bycatch_explanatory[!is.na(TARGETED_TAXA) & SPECIES == "Delphinus delphis"], aes(x = forcats::fct_infreq(TARGETED_TAXA), fill = forcats::fct_infreq(TARGETED_ISSCAAP))) +
  geom_bar()  +
  labs(y = "Number of captures", x = "Targeted taxa", fill = "ISSCAAP groups") +
  ggtitle("Merged data - Common dolphin") +
  theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  scale_fill_discrete(labels = c("31 - Pleuronectiformes (N=105)", "32 - Gadiformes (N=91)", "33 - Miscellaneous coastal fishes (N=68)", "34 - Miscellaneous demersal fishes  (N=17)",  "36 - Tunas, bonitos (N=17)", "37 - Miscellaneous pelagic fishes (N=12)", "57 - Squids, cuttlefishes, octopuses (N=5)", "35 - Herrings, sardines, anchovies  (N=3)", "43 - Lobsters, spiny-rock lobsters (N=3)",  "38 - Sharks, rays (N=2)"))

ggsave("Figure_S8A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

ggplot(Decl_Obs_Merged_Bycatch_explanatory[!is.na(TARGETED_TAXA) & SPECIES == "Phocoena phocoena"], aes(x = forcats::fct_infreq(TARGETED_TAXA), fill = forcats::fct_infreq(TARGETED_ISSCAAP))) +
  geom_bar()  +
  labs(y = "Number of captures", x = "Targeted taxa", fill = "ISSCAAP groups") +
  ggtitle("Merged data - Harbour porpoise") +
  theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  scale_fill_discrete(labels = c("31 - Pleuronectiformes (N=44)", "34 - Miscellaneous demersal fishes  (N=32)", "32 - Gadiformes (N=12)", "33 - Miscellaneous coastal fishes (N=3)",  "38 - Sharks, rays (N=2)", "37 - Miscellaneous pelagic fishes (N=1)", "43 - Lobsters, spiny-rock lobsters (N=1)", "57 - Squids, cuttlefishes, octopuses (N=1)"))

ggsave("Figure_S8B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

# Distribution of actually mostly fished taxa associated with by-catch (Figure S9):

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis", MOST_FISHED_SP], OBSMER_Bycatch[SPECIES == "Delphinus delphis", MOST_FISHED_ISSCAAP])

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis", MOST_FISHED_SPP], OBSMER_Bycatch[SPECIES == "Delphinus delphis", MOST_FISHED_ISSCAAP])

OBSMER_Bycatch[
  SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "31 - Pleuronectiformes", Most_fished_ISSCAAP_associated_taxa := "Solea solea (97%)"][
    SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "32 - Gadiformes", Most_fished_ISSCAAP_associated_taxa := "Merluccius merluccius (64%)\nand Gadidae (33%, haddock, whiting, pollack)"][
      SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "33 - Miscellaneous coastal fishes", Most_fished_ISSCAAP_associated_taxa := "Dicentrarchus spp (93%)"][
        SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "34 - Miscellaneous demersal fishes", Most_fished_ISSCAAP_associated_taxa := "Lophius spp (58%) and Zeidae (33%)"][
          SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "36 - Tunas, bonitos", Most_fished_ISSCAAP_associated_taxa := "Thunnus spp (90%)"][
            SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "37 - Miscellaneous pelagic fishes", Most_fished_ISSCAAP_associated_taxa := "Trachurus spp (78%) and Scomber spp (22%)"][
              SPECIES == "Delphinus delphis" & MOST_FISHED_ISSCAAP == "38 - Sharks, rays", Most_fished_ISSCAAP_associated_taxa := "Raja spp (42%) and diverse shark species\n(50%, Scyliorhinus, Alopias and Mustelus spp)"]

# To better illustrate the distribution of mostly fished taxa associated with by-catch, we defined the proportion of most fished species or group of species mainly associated with mostly fished ISSCAAP groups (only for ISSCAAP groups with more than 6 by-catch observations, as other groups will not be considered for analyses because too rare). The proportion were computed from the the above tables and the total number of data available for the most fished associated ISSCAAP group.

ggplot(OBSMER_Bycatch[!is.na(MOST_FISHED_ISSCAAP) & SPECIES == "Delphinus delphis"], aes(x = forcats::fct_infreq(MOST_FISHED_ISSCAAP), fill = forcats::fct_infreq(Most_fished_ISSCAAP_associated_taxa))) +
  geom_bar()  +
  labs(y = "Number of captures", x = "Mostly fished ISSCAAP group", fill = "Main associated fished taxa") +
  ggtitle("OBSMER - Common dolphin") +
  theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
   scale_fill_discrete(breaks = c("Merluccius merluccius (64%)\nand Gadidae (33%, haddock, whiting, pollack)", "Dicentrarchus spp (93%)", "Solea solea (97%)", "Lophius spp (58%) and Zeidae (33%)", "Raja spp (42%) and diverse shark species\n(50%, Scyliorhinus, Alopias and Mustelus spp)", "Thunnus spp (90%)", "Trachurus spp (78%) and Scomber spp (22%)"),labels = c("*Merluccius merluccius* (64%)<br>and Gadidae (33%, haddock, whiting, pollack)", "Dicentrarchus spp (93%)", "*Solea solea* (97%)", "Lophius spp (58%) and Zeidae (33%)", "Raja spp (42%) and diverse shark species<br>(50%, Scyliorhinus, Alopias and Mustelus spp)", "Thunnus spp (90%)", "Trachurus spp (78%) and Scomber spp (22%)")) +
    theme(legend.text = element_markdown())

ggsave("Figure_S9A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena", MOST_FISHED_SP], OBSMER_Bycatch[SPECIES == "Phocoena phocoena", MOST_FISHED_ISSCAAP])

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena", MOST_FISHED_SPP], OBSMER_Bycatch[SPECIES == "Phocoena phocoena", MOST_FISHED_ISSCAAP])

OBSMER_Bycatch[
  SPECIES == "Phocoena phocoena" & MOST_FISHED_ISSCAAP == "31 - Pleuronectiformes", Most_fished_ISSCAAP_associated_taxa := "Solea spp (79%) and Scophthalmidae (21%)"][
    SPECIES == "Phocoena phocoena" & MOST_FISHED_ISSCAAP == "32 - Gadiformes", Most_fished_ISSCAAP_associated_taxa := "Gadidae (88%, mostly pollack)"][
      SPECIES == "Phocoena phocoena" & MOST_FISHED_ISSCAAP == "45 - Crabs, sea-spiders", Most_fished_ISSCAAP_associated_taxa := "Cancer pagurus (71%)"][
        SPECIES == "Phocoena phocoena" & MOST_FISHED_ISSCAAP == "34 - Miscellaneous demersal fishes", Most_fished_ISSCAAP_associated_taxa := "Lophius spp (84%)"]

# To better illustrate the distribution of mostly fished taxa associated with by-catch, we defined the proportion of most fished species or group of species mainly associated with mostly fished ISSCAAP groups (only for ISSCAAP groups with more than 6 by-catch observations, as other groups will not be considered for analyses because too rare). The proportion were computed from the the above tables and the total number of data available for the most fished associated ISSCAAP group.

ggplot(OBSMER_Bycatch[!is.na(MOST_FISHED_ISSCAAP) & SPECIES == "Phocoena phocoena"], aes(x = forcats::fct_infreq(MOST_FISHED_ISSCAAP), fill = forcats::fct_infreq(Most_fished_ISSCAAP_associated_taxa))) +
  geom_bar()  +
  labs(y = "Number of captures", x = "Mostly fished ISSCAAP group", fill = "Main associated fished taxa") +
  ggtitle("OBSMER - Harbour porpoise") +
  theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
   scale_fill_discrete(breaks = c("Solea spp (79%) and Scophthalmidae (21%)","Lophius spp (84%)","Gadidae (88%, mostly pollack)","Cancer pagurus (71%)"), labels = c("Solea spp (79%) and Scophthalmidae (21%)","Lophius spp (84%)","Gadidae (87%, mostly pollack)","*Cancer pagurus* (71%)")) +
    theme(legend.text = element_markdown())

ggsave("Figure_S9B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

```

## Distribution of by-caught individuals' traits as a function of fishing techniques (vulnerabilities to fishing techniques)

We then studied how traits of by-caught individuals varied with fishing techniques used so as to identify potential phenotypic vulnerabilities. Three phenotypic traits were studied: the by-caught individuals' body mass (using both declaration and observation data), body length and sex (using observation data only).

### Weight model (observation and declaration data combined)

```{r weight model}

# To avoid too rare categories to be included in our analyses and to facilitate the interpretation of potential effects, we have combined the different fishing gear categories into two main groups: trawls and seines (hereafter referred to as trawls), and set gillnets and trammel nets (hereafter referred to as gillnets)

Decl_Obs_Merged_Bycatch[
  FISHING_GEAR %in% c("GNS - Set gillnets", "GTN - Combined gillnets-trammel nets", "GTR - Trammel nets"), FISHING_GEAR_grouped := "Gillnets"][
    FISHING_GEAR %in% c("OTB - Single boat bottom otter trawls", "OTT - Twin bottom otter trawls", "PTB - Bottom pair trawls", "PS - Purse seines", "SDN - Danish seine", "OTM - Single boat midwater otter trawls", "PTM - Midwater pair trawls"), FISHING_GEAR_grouped := "Trawls"]

Decl_Obs_Merged_Bycatch[, FISHING_GEAR_grouped := as.factor(FISHING_GEAR_grouped)]
Decl_Obs_Merged_Bycatch[, TARGETED_ISSCAAP := as.factor(TARGETED_ISSCAAP)]

######## Harbour porpoise:

# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped)]$TARGETED_ISSCAAP)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped)]$FISHING_GEAR_grouped)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & TARGETED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)
# in this specific case there is rare categories in both categorical variables: removing rare categories in targeted ISSCAAP groups categories lead to the presence of too rare categories in used fishing gears. We therefore cannot use both variables in the model. In this particular case we chose to favor the use of used fishing gear in explanatory variables.

# dataset used for body mass regression analysis on harbour porpoises:
Decl_Obs_Merged_Bycatch_BM_PHO = Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(FISHING_GEAR_grouped) & !is.na(MESH_SIZE)]

# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):

vif(lm(data = Decl_Obs_Merged_Bycatch_BM_PHO, BODY_MASS ~ MESH_SIZE + FISHING_GEAR_grouped))
# correlation level between explanatory variables reasonable (GVIF^(1/(2*Df)) < 2.24)

# implementation and analyses of the regression model:

model_BM_PHO <- lm(data = Decl_Obs_Merged_Bycatch_BM_PHO, BODY_MASS ~ MESH_SIZE + FISHING_GEAR_grouped)

summary(model_BM_PHO)
Anova(model_BM_PHO)

plot(ggemmeans(model_BM_PHO, terms = "MESH_SIZE"), show_data = T, jitter = 2, dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body mass\nas a function of used mesh size") +
  labs(y = "Body mass (kg)", x = "Mesh size (mm)")  +
  scale_y_continuous(limits = c(75, 300)) +
  theme(text = element_text(size = 30))

ggsave("Figure_4A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

simulateResiduals(fittedModel = model_BM_PHO, plot = T) 
plot(cooks.distance(model_BM_PHO) )  
# diagnostic plots of model: OK

######## Common dolphin:

# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped)]$TARGETED_ISSCAAP)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped)]$FISHING_GEAR_grouped)

table(Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & TARGETED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)

# dataset used for body mass regression analysis on common dolphin:
Decl_Obs_Merged_Bycatch_BM_DCO = Decl_Obs_Merged_Bycatch[SPECIES == "Delphinus delphis" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & TARGETED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "34 - Miscellaneous demersal fishes")]

# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):

vif(lm(data = Decl_Obs_Merged_Bycatch_BM_DCO, BODY_MASS ~ MESH_SIZE + FISHING_GEAR_grouped + TARGETED_ISSCAAP))
# correlation level between explanatory variables reasonable (GVIF^(1/(2*Df)) < 2.24)

anova(lmer(data = Decl_Obs_Merged_Bycatch_BM_DCO, BODY_MASS ~ 1 + (1|YEAR_DCO)), lm(data = Decl_Obs_Merged_Bycatch_BM_DCO, BODY_MASS ~ 1))
 #random effects does not seems necessary (to be noticed: no difference if included)

# reordering variable to improve plot readability:

Decl_Obs_Merged_Bycatch_BM_DCO$TARGETED_ISSCAAP = factor(Decl_Obs_Merged_Bycatch_BM_DCO$TARGETED_ISSCAAP, levels = c("34 - Miscellaneous demersal fishes", "31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes"), labels = c("34 - Miscellaneous\ndemersal fishes\n(mainly monkfishes)", "31 - Pleuronectiformes\n(mainly soles)", "32 - Gadiformes\n(mainly hake)", "33 - Miscellaneous\ncoastal fishes\n(mainly sea bass)"))

# implementation and analyses of the regression model:

model_BM_DCO <- lm(data = Decl_Obs_Merged_Bycatch_BM_DCO, BODY_MASS ~ MESH_SIZE + FISHING_GEAR_grouped + TARGETED_ISSCAAP)

summary(model_BM_DCO)
Anova(model_BM_DCO)

plot(ggemmeans(model_BM_DCO, terms = "FISHING_GEAR_grouped"), show_data = T, jitter=c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body mass\nas a function of used fishing gear") +
  labs(y = "Body mass (kg)", x = "Fishing gears")  +
  theme(text = element_text(size = 30))

ggsave("Figure_5A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BM_DCO, terms = "TARGETED_ISSCAAP"), show_data = T, jitter=c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body mass\nas a function of targeted taxa") +
  labs(y = "Body mass (kg)", x = "Targeted taxa (ISSCAAP group)")  +
  theme(text = element_text(size = 30))   +  theme(axis.text.x = element_text(angle = 55, hjust = 1))

ggsave("Figure_5B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(model_BM_DCO, "TARGETED_ISSCAAP"))
pairs(emmeans(model_BM_DCO, "TARGETED_ISSCAAP"), adjust = "none")
# post-hoc pairwise comparison tests

simulateResiduals(fittedModel = model_BM_DCO, plot = T) 
plot(cooks.distance(model_BM_DCO) )  
# Diagnostic plots revealed deviations from statistical assumptions

hist(Decl_Obs_Merged_Bycatch_BM_DCO$BODY_MASS) # positively skewed distribution (same for residuals)
hist(sqrt(Decl_Obs_Merged_Bycatch_BM_DCO$BODY_MASS)) # way better after squared root transformation

model_BM_DCO_sqrt <- lm(data = Decl_Obs_Merged_Bycatch_BM_DCO, sqrt(BODY_MASS) ~ MESH_SIZE + FISHING_GEAR_grouped + TARGETED_ISSCAAP)

summary(model_BM_DCO_sqrt)
Anova(model_BM_DCO_sqrt)

plot(ggemmeans(model_BM_DCO_sqrt, terms = "FISHING_GEAR_grouped"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body mass\nas a function of used fishing gear") +
  labs(y = "Body mass (kg)", x = "Fishing gears")  +
  theme(text = element_text(size = 30)) 
ggsave("Figure_S14A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BM_DCO_sqrt, terms = "TARGETED_ISSCAAP"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body mass\nas a function of targeted taxa") +
  labs(y = "Body mass (kg)", x = "Targeted taxa (ISSCAAP group)")  +
  theme(text = element_text(size = 30))   +  theme(axis.text.x = element_text(angle = 55, hjust = 1))
ggsave("Figure_S14B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(model_BM_DCO_sqrt, "TARGETED_ISSCAAP"))
pairs(emmeans(model_BM_DCO_sqrt, "TARGETED_ISSCAAP"), adjust = "none")
# post-hoc pairwise comparison tests

# output nearly identical than before transforming data

simulateResiduals(fittedModel = model_BM_DCO_sqrt, plot = T) 
plot(cooks.distance(model_BM_DCO_sqrt) )  
#diagnostic plots of model: OK

```

### Size model (OBSMER)

```{r size model}

# To avoid too rare categories to be included in our analyses and to facilitate the interpretation of potential effects, we have combined the different fishing gear categories into two main groups: trawls and seines (hereafter referred to as trawls), and set gillnets and trammel nets (hereafter referred to as gillnets)

OBSMER_Bycatch[
  FISHING_GEAR %in% c("GNS - Set gillnets", "GTN - Combined gillnets-trammel nets", "GTR - Trammel nets"), FISHING_GEAR_grouped := "Gillnets"][
    FISHING_GEAR %in% c("OTB - Single boat bottom otter trawls", "OTT - Twin bottom otter trawls", "PTB - Bottom pair trawls", "PS - Purse seines", "SDN - Danish seine", "OTM - Single boat midwater otter trawls", "PTM - Midwater pair trawls"), FISHING_GEAR_grouped := "Trawls"]

OBSMER_Bycatch[, FISHING_GEAR_grouped := as.factor(FISHING_GEAR_grouped)]
OBSMER_Bycatch[, MOST_FISHED_ISSCAAP := as.factor(MOST_FISHED_ISSCAAP)]
OBSMER_Bycatch[, DETERRENT_DEVICE := as.factor(DETERRENT_DEVICE)]
OBSMER_Bycatch[, SEX := as.factor(SEX)]

# OBSMER data contained more detailed descriptions of fishing techniques. Some of them contained a large part of missing values or were highly inter-correlated. We removed such variables from analyses (see details below).

apply(OBSMER_Bycatch, 2, function(x) table(is.na(x)))
# some variable present too many missing values to be included in analyses (as gear speed only available for trawls)

cor.test(OBSMER_Bycatch$GEAR_DIMENSION, OBSMER_Bycatch$MESH_SIZE)
wilcox.test(OBSMER_Bycatch$GEAR_DIMENSION ~ OBSMER_Bycatch$FISHING_GEAR_grouped)
kruskal.test(OBSMER_Bycatch$GEAR_DIMENSION, OBSMER_Bycatch$MOST_FISHED_ISSCAAP)
# gear dimension strongly correlated with gear type, targeted taxa, and mesh size

kruskal.test(OBSMER_Bycatch$GEAR_DEPTH, OBSMER_Bycatch$MOST_FISHED_ISSCAAP)
# gear depth was well described by the targeted taxa (and was also associated with 14% of missing value)

# gear speed, dimension and depth were discarded from analyses due to too many missing values and/or because they were too well described by other variables (in this latter case including these varaibles in regression models resulted in too high VIF values)

# Other remaining variables (namely: gear type, mesh size, most caught taxa, total catch weight, soaking time, and presence of an acoustic deterrent device) were used in all subsequent analyses using OBSMER data

######## Harbour porpoise:

# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$MOST_FISHED_ISSCAAP)

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$FISHING_GEAR_grouped)

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$DETERRENT_DEVICE)
# There is not enough diversity in the presence of an acoustic deterrent to include this variable in the present analysis

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & TARGETED_ISSCAAP %in% c("31 - Pleuronectiformes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)
# in this specific case there is rare categories in both categorical variables: removing rare categories in targeted ISSCAAP groups categories lead to the presence of too rare categories in used fishing gears. We therefore cannot use both variables in the model. In this particular case we chose to favor the use of used fishing gear in explanatory variables.

# dataset used for body length regression analysis on harbour porpoises:
OBSMER_Bycatch_BL_PHO = OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(FISHING_GEAR_grouped) & !is.na(MESH_SIZE) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT)]

# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):

vif(lm(data = OBSMER_Bycatch_BL_PHO, BODY_LENGTH ~ MESH_SIZE + FISHING_GEAR_grouped + DURATION_OP + TOTAL_CATCH_WEIGHT))
# correlation level between explanatory variables reasonable (GVIF < 5)

# implementation and analyses of the regression model:

model_BL_PHO <- lm(data = OBSMER_Bycatch_BL_PHO, BODY_LENGTH ~ MESH_SIZE + FISHING_GEAR_grouped + DURATION_OP + TOTAL_CATCH_WEIGHT)

summary(model_BL_PHO)
Anova(model_BL_PHO)

plot(ggemmeans(model_BL_PHO, terms = "MESH_SIZE"), show_data = T, jitter = 2, dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of used mesh size") +
  labs(y = "Body length (cm)", x = "Mesh size (mm)")  +
  theme(text = element_text(size = 30))

ggsave("Figure_4B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BL_PHO, terms = "FISHING_GEAR_grouped"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of used fishing gear") +
  labs(y = "Body length (cm)", x = "Fishing gear")  +
  theme(text = element_text(size = 30))

ggsave("Figure_4C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

simulateResiduals(fittedModel = model_BL_PHO, plot = T) 
plot(cooks.distance(model_BL_PHO) )  
# diagnostic plots of model: OK

######## Common dolphin:

# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$MOST_FISHED_ISSCAAP)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$FISHING_GEAR_grouped)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$DETERRENT_DEVICE)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "34 - Miscellaneous demersal fishes", "36 - Tunas, bonitos", "37 - Miscellaneous pelagic fishes", "38 - Sharks, rays")]$FISHING_GEAR_grouped)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "34 - Miscellaneous demersal fishes", "36 - Tunas, bonitos", "37 - Miscellaneous pelagic fishes", "38 - Sharks, rays")]$DETERRENT_DEVICE)

# dataset used for body length regression analysis on common dolphin:
OBSMER_Bycatch_BL_DCO = OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "34 - Miscellaneous demersal fishes", "36 - Tunas, bonitos", "37 - Miscellaneous pelagic fishes", "38 - Sharks, rays")]

# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):

vif(lm(data = OBSMER_Bycatch_BL_DCO, BODY_LENGTH ~ MESH_SIZE + FISHING_GEAR_grouped + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT + DETERRENT_DEVICE))
# correlation level between explanatory variables reasonable (GVIF^(1/(2*Df)) < 2.24)

anova(lmer(data = OBSMER_Bycatch_BL_DCO, BODY_LENGTH ~ 1 + (1|TRIMESTER)), lm(data = OBSMER_Bycatch_BL_DCO, BODY_LENGTH ~ 1))
 #random effects does not seems necessary (to be noticed: no difference if included)

# reordering variable to improve plot readability:

OBSMER_Bycatch_BL_DCO$MOST_FISHED_ISSCAAP = factor(OBSMER_Bycatch_BL_DCO$MOST_FISHED_ISSCAAP, levels = c("31 - Pleuronectiformes", "38 - Sharks, rays", "32 - Gadiformes", "36 - Tunas, bonitos", "37 - Miscellaneous pelagic fishes", "33 - Miscellaneous coastal fishes", "34 - Miscellaneous demersal fishes"), labels = c("31 - Pleuronectiformes\n(mainly soles)", "38 - Sharks, rays", "32 - Gadiformes\n(mainly hake)", "36 - Tunas, bonitos", "37 - Miscellaneous\npelagic fishes\n(mainly mackerells)", "33 - Miscellaneous\ncoastal fishes\n(mainly sea bass)", "34 - Miscellaneous\ndemersal fishes\n(mainly monkfishes)"))

# implementation and analyses of the regression model:

model_BL_DCO <- lm(data = OBSMER_Bycatch_BL_DCO, BODY_LENGTH ~  MESH_SIZE + FISHING_GEAR_grouped + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT + DETERRENT_DEVICE)

summary(model_BL_DCO)
Anova(model_BL_DCO)

plot(ggemmeans(model_BL_DCO, terms = "FISHING_GEAR_grouped"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of used fishing gear") +
  labs(y = "Body length (cm)", x = "Fishing gears")  +
  theme(text = element_text(size = 30))

ggsave("Figure_5C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BL_DCO, terms = "MOST_FISHED_ISSCAAP"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of most fished taxa") +
  labs(y = "Body length (cm)", x = "Most fished taxa (ISSCAAP group)")  +
  theme(text = element_text(size = 30))   +  theme(axis.text.x = element_text(angle = 55, hjust = 1))

ggsave("Figure_5D.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BL_DCO, terms = "DETERRENT_DEVICE"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of deterrent device presence") +
  labs(y = "Body length (cm)", x = "Presence of a deterrent device")  +
  theme(text = element_text(size = 30))   +
  scale_x_continuous(breaks=c(1:2), labels=c("Absence", "Presence"))

ggsave("Figure_5E.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(model_BL_DCO, "MOST_FISHED_ISSCAAP"))
pairs(emmeans(model_BL_DCO, "MOST_FISHED_ISSCAAP"), adjust = "none")
# post-hoc pairwise comparison tests

simulateResiduals(fittedModel = model_BL_DCO, plot = T) 
plot(cooks.distance(model_BL_DCO) )  
# Diagnostic plots revealed deviations from statistical assumptions

hist(OBSMER_Bycatch_BL_DCO$BODY_LENGTH) # negatively skewed distribution (same for residuals)
hist(OBSMER_Bycatch_BL_DCO$BODY_LENGTH^2) # way better after squared transformation

model_BL_DCO_squared <- lm(data = OBSMER_Bycatch_BL_DCO, I(BODY_LENGTH^2) ~ MESH_SIZE + FISHING_GEAR_grouped + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT + DETERRENT_DEVICE)

summary(model_BL_DCO_squared)
Anova(model_BL_DCO_squared)

plot(ggemmeans(model_BL_DCO_squared, terms = "FISHING_GEAR_grouped"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of used fishing gear") +
  labs(y = "Body length (cm)", x = "Fishing gears")  +
  theme(text = element_text(size = 30))

ggsave("Figure_S14C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BL_DCO_squared, terms = "MOST_FISHED_ISSCAAP"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of most fished taxa") +
  labs(y = "Body length (cm)", x = "Most fished taxa (ISSCAAP group)")  +
  theme(text = element_text(size = 30))   +  theme(axis.text.x = element_text(angle = 55, hjust = 1))

ggsave("Figure_S14D.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

plot(ggemmeans(model_BL_DCO_squared, terms = "DETERRENT_DEVICE"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's body length\nas a function of deterrent device presence") +
  labs(y = "Body length (cm)", x = "Presence of a deterrent device")  +
  theme(text = element_text(size = 30))   +
  scale_x_continuous(breaks=c(1:2), labels=c("Absence", "Presence"))

ggsave("Figure_S14E.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

pairs(emmeans(model_BL_DCO_squared, "MOST_FISHED_ISSCAAP"))
pairs(emmeans(model_BL_DCO_squared, "MOST_FISHED_ISSCAAP"), adjust = "none")
# post-hoc pairwise comparison tests

# output nearly identical than before transforming data

simulateResiduals(fittedModel = model_BL_DCO_squared, plot = T) 
plot(cooks.distance(model_BL_DCO_squared) )  
#diagnostic plots of model: OK (still significant deviation for first quantile, but reasonable and combined adjusted quantile NS)

############ Modifier legende axe abscisse pour simplifier interpretation ?????

#reordering variables for plots:

# OBSMER_size_del$ISSCAAP=factor(OBSMER_size_del$ISSCAAP, levels = c("31","38","32","36","37","33","34"), labels=c("Pleuronectiformes\n(mainly soles)", "Sharks, rays", "Gadiformes\n(mainly hake and gadoids)", "Tunas, bonitos", "Miscellaneous pelagic fishes\n(mainly mackerells)", "Miscellaneous coastal fishes\n(mainly sea bass)", "Miscellaneous demersal fishes\n(mainly monkfishes or St Pierre)")) # 31:poisson plats, 32:merlus et gadidés, 33: mainly bass, 34: démerseaux (baudroies, saint pierre), 36: thon, bonite, 37: maqueraux, chinchard, 38: requins, raies

```

### Sex model (OBSMER)

```{r sex model}

######## Harbour porpoise:

# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$MOST_FISHED_ISSCAAP)

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$FISHING_GEAR_grouped)
# There is not enough diversity in the different used fishing gears to include this variable in the present analysis

table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$DETERRENT_DEVICE)
# There is not enough diversity in the presence of an acoustic deterrent to include this variable in the present analysis

# dataset used for sex regression analysis on harbour porpoises:
OBSMER_Bycatch_Sex_PHO = OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "34 - Miscellaneous demersal fishes")]

# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):

vif(glm(data = OBSMER_Bycatch_Sex_PHO, SEX ~ MESH_SIZE + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT, family = "binomial"))
# correlation level between explanatory variables is too high for mesh size and most fished taxa variables (GVIF > 5). We removed from the analysis the variable with the highest VIF: the mesh size.

model_sex_PHO_high_vif <- glm(data = OBSMER_Bycatch_Sex_PHO, SEX ~ MESH_SIZE + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT, family = "binomial")

vif(glm(data = OBSMER_Bycatch_Sex_PHO, SEX ~ MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT, family = "binomial"))
# correlation level between explanatory variables reasonable (GVIF < 5)

# implementation and analyses of the regression model:

model_sex_PHO <- glm(data = OBSMER_Bycatch_Sex_PHO, SEX ~ MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT, family = "binomial")

summary(model_sex_PHO)
Anova(model_sex_PHO)

simulateResiduals(fittedModel = model_BL_PHO, plot = T) 
plot(cooks.distance(model_BL_PHO) )  
# diagnostic plots of model: OK

######## Common dolphin:

# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$MOST_FISHED_ISSCAAP)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$FISHING_GEAR_grouped)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE)]$DETERRENT_DEVICE)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "38 - Sharks, rays")]$FISHING_GEAR_grouped)

table(OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "38 - Sharks, rays")]$DETERRENT_DEVICE)

# dataset used for sex regression analysis on common dolphin:
OBSMER_Bycatch_Sex_DCO = OBSMER_Bycatch[SPECIES == "Delphinus delphis" & !is.na(SEX) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & !is.na(DURATION_OP) & !is.na(TOTAL_CATCH_WEIGHT) & !is.na(DETERRENT_DEVICE) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "33 - Miscellaneous coastal fishes", "38 - Sharks, rays")]

# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):

vif(glm(data = OBSMER_Bycatch_Sex_DCO, SEX ~ MESH_SIZE + FISHING_GEAR_grouped + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT + DETERRENT_DEVICE, family = "binomial"))
# correlation level between explanatory variables reasonable (GVIF^(1/(2*Df)) < 2.24)

# implementation and analyses of the regression model:

model_sex_DCO <- glm(data = OBSMER_Bycatch_Sex_DCO, SEX ~  MESH_SIZE + FISHING_GEAR_grouped + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT + DETERRENT_DEVICE, family = "binomial")

summary(model_sex_DCO)
Anova(model_sex_DCO)

plot(ggemmeans(model_sex_DCO, terms = "TOTAL_CATCH_WEIGHT"), show_data = T, jitter= c(0, 0.1), dot_alpha = 0.2) +
  ggtitle("Predicted by-caught individual's sex\nas a function of total catch weight") +
  labs(y = "By-caught individual's sex (probability)", x = "Total catch weight (kg)")  +
  theme(text = element_text(size = 30))  #more females captures when high fishing weight, but this effect is driven by only one capture events for a group only constituted of females (highly sensitive to one outlier event, without it there is no significant effects anymore of this variable: see below)

ggsave("Figure_S15.pdf", device = "pdf", units = "in", width = 14, height = 8.5)

model_sex_DCO_wo_outlier <- glm(data = OBSMER_Bycatch_Sex_DCO[TOTAL_CATCH_WEIGHT<5000], SEX ~  MESH_SIZE + FISHING_GEAR_grouped + MOST_FISHED_ISSCAAP + DURATION_OP + TOTAL_CATCH_WEIGHT + DETERRENT_DEVICE, family = "binomial")

summary(model_sex_DCO_wo_outlier)
Anova(model_sex_DCO_wo_outlier)

simulateResiduals(fittedModel = model_sex_DCO, plot = T) 
plot(cooks.distance(model_sex_DCO) )  
# Diagnostic plots OK

```

### Output statististic tables (vulnerabilities regressions)

```{r tables susceptibility models}

### Global table (Table S6) of vulnerabilities regression models (ANOVA, type II):

TableS6 = merge(merge(merge(merge(merge(merge(merge(
  as.data.table(Anova(model_sex_DCO)[, -2], keep.rownames = T), as.data.table(Anova(model_BM_DCO)[-4, c(-1, -2)], keep.rownames = T), by = "rn", all = T),
  as.data.table(Anova(model_BM_DCO_sqrt)[-4, c(-1, -2)], keep.rownames = T), by = "rn", all = T),
  as.data.table(Anova(model_BL_DCO)[-7, c(-1, -2)], keep.rownames = T), by = "rn", all = T),
  as.data.table(Anova(model_BL_DCO_squared)[-7, c(-1, -2)], keep.rownames = T), by = "rn", all = T)[, setnames(.SD, c("F value.x", "Pr(>F).x", "F value.y", "Pr(>F).y"), c("F value.xz", "Pr(>F).xz", "F value.yz", "Pr(>F).yz"))],
  as.data.table(Anova(model_sex_PHO)[, -2], keep.rownames = T), by = "rn", all = T),
  as.data.table(Anova(model_BM_PHO)[-3, c(-1, -2)], keep.rownames = T), by = "rn", all = T),
  as.data.table(Anova(model_BL_PHO)[-5, c(-1, -2)], keep.rownames = T), by = "rn", all = T)

TableS6$Df = c(1, 1, 1, 1, "3/6/1", 3, 1)

TableS6 = TableS6[c(3, 4, 6, 5, 2, 7, 1), c(1, 18, 2:17)]

TableS6 = cbind(TableS6[, 1:2], mapply(function(x, y) paste0(signif(x, digits = 3), " (", signif(y, digits = 3), ")"), TableS6[, seq(3, 18, 2), with = F], TableS6[, seq(4, 18, 2), with = F])) # grouping of statistical values with associated p-values

setDF(TableS6)

TableS6 = cbind(TableS6[, 1:3], paste0(TableS6[, 4], " / ", TableS6[, 5]), paste0(TableS6[, 6], " / ", TableS6[, 7]), TableS6[8:10]) # grouping of models before/after variable transformation

setDT(TableS6)

TableS6 = TableS6[, lapply(.SD, function(x) ifelse(x %in% c("NA (NA) / NA (NA)", "NA (NA)"), "", x))]

colnames(TableS6) = c("", "Df", "LR Chisq (P-value)", rep("F Value (P-value)", 2), "LR Chisq (P-value)", rep("F Value (P-value)", 2))

TableS6[,1] = c("Used gear", "Mesh size", "Targeted taxa (ISSCAAP code)", "Fished taxa (ISSCAAP code)", "Soaking time", "Total catch weight", "Presence of an acoustic repellent")

setDF(TableS6)

TableS6 = rbind(TableS6, c("Number of observations", "", model_sex_DCO$df.null, model_BM_DCO$df.residual+7, model_BL_DCO$df.residual+11, model_sex_PHO$df.null, model_BM_PHO$df.residual+3, model_BL_PHO$df.residual+4))

TableS6 = rbind(TableS6, c("R-squared", "", signif(rsq.n(model_sex_DCO), digits = 3),
                           paste0(signif(summary(model_BM_DCO)$r.squared, digits = 3), " / ", signif(summary(model_BM_DCO_sqrt)$r.squared, digits = 3)),
                           paste0(signif(summary(model_BL_DCO)$r.squared, digits = 3), " / ", signif(summary(model_BL_DCO_squared)$r.squared, digits = 3)), 
                           signif(rsq.n(model_sex_PHO), digits = 3),
                           signif(summary(model_BM_PHO)$r.squared, digits = 3),
                           signif(summary(model_BL_PHO)$r.squared, digits = 3)))

TableS6 = rbind(colnames(TableS6), TableS6)

colnames(TableS6) = c("", "", "Sex (logistic regression)", "Body mass (linear regression)", "Body size (linear regression)", "Sex (logistic regression)", "Body mass (linear regression)", "Body size (linear regression)")

kable_classic(add_header_above(kbl(TableS6, format = "latex", vline= ""), c("", "", "Common dolphin" = 3 , "Harbour porpoise" = 3)))

### Table (S7) combining pairwise contrast comparison test for each significant variable in vulnerabilities regression models:

TableS7 = rbind( 
  cbind('Response/Explanatory variables'= "BM / Targeted taxa", 'Response variable transformation'= "None", as.data.frame(pairs(emmeans(model_BM_DCO, "TARGETED_ISSCAAP"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(model_BM_DCO, "TARGETED_ISSCAAP")))$p.value),
    cbind('Response/Explanatory variables'= "BM / Targeted taxa", 'Response variable transformation'= "Squared root", as.data.frame(pairs(emmeans(model_BM_DCO_sqrt, "TARGETED_ISSCAAP"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(model_BM_DCO_sqrt, "TARGETED_ISSCAAP")))$p.value),
  cbind('Response/Explanatory variables'= "BL / Fished taxa (ISSCAAP code)", 'Response variable transformation'= "None", as.data.frame(pairs(emmeans(model_BL_DCO, "MOST_FISHED_ISSCAAP"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(model_BL_DCO, "MOST_FISHED_ISSCAAP")))$p.value),
    cbind('Response/Explanatory variables'= "BL / Fished taxa (ISSCAAP code)", 'Response variable transformation'= "Squared", as.data.frame(pairs(emmeans(model_BL_DCO_squared, "MOST_FISHED_ISSCAAP"), adjust = "none")), adj.p.value = as.data.frame(pairs(emmeans(model_BL_DCO_squared, "MOST_FISHED_ISSCAAP")))$p.value)
  )

TableS7$df = NULL

TableS7[sapply(TableS7, is.numeric)] = lapply(TableS7[sapply(TableS7, is.numeric)], round, 3)

TableS7 = TableS7[TableS7$p.value < 0.05,]

rownames(TableS7) = NULL

setDT(TableS7)
TableS7[, lapply(.SD, function(x) ifelse(str_detect(x, "\n"), str_replace_all(x, "\n", " "), x))]
setDF(TableS7)

 kable_classic(collapse_rows(kbl(TableS7, format = "latex"), columns = 1:2, target = 1, valign= "middle"))

 ### VIF table (table S1):

TableS1 = merge(merge(merge(merge(merge(merge(
  as.data.table(vif(model_sex_DCO)[, -2], keep.rownames = T), as.data.table(vif(model_BM_DCO)[, -2], keep.rownames = T), by = "rn", all = T),
  as.data.table(vif(model_BL_DCO)[, -2], keep.rownames = T), by = "rn", all = T, suffixes = c("a", "b")),
  as.data.table(vif(model_sex_PHO_high_vif), keep.rownames = T), by.x = "rn", by.y = "V1", all = T),
  as.data.table(vif(model_sex_PHO), keep.rownames = T), by.x = "rn", by.y = "V1", all = T),
  as.data.table(vif(model_BM_PHO), keep.rownames = T), by.x = "rn", by.y = "V1", all = T),
  as.data.table(vif(model_BL_PHO), keep.rownames = T), by.x = "rn", by.y = "V1", all = T, suffixes = c("c", "d"))

TableS1$Df = c(1, 1, 1, 1, "3/6/1", 3, 1)

TableS1 = TableS1[c(3, 4, 6, 5, 2, 7, 1), c(1, 12, 2:11)]

setDF(TableS1)

TableS1[sapply(TableS1, is.numeric)] = lapply(TableS1[sapply(TableS1, is.numeric)], round, 3)

TableS1 = cbind(TableS1[, 1:2], paste0(TableS1[, 3], " (", TableS1[, 4], ")"), paste0(TableS1[, 5], " (", TableS1[, 6], ")"), paste0(TableS1[, 7], " (", TableS1[, 8], ")"), TableS1[9:12])

TableS1 = cbind(TableS1[, 1:5], paste0(TableS1[, 6], " / ", TableS1[, 7]), TableS1[, 8:9])

setDT(TableS1)

TableS1=TableS1[, lapply(.SD, function(x) ifelse(x %in% c(NA, "NA / NA", "NA (NA)", "NA (NA) / NA (NA)"), "", x))]

colnames(TableS1)=c("", "Df", rep(c("GVIF (GVIF^(1/(2*Df)))"), 3), rep("GVIF", 3))

TableS1[,1] = c("Used gear", "Mesh size", "Targeted taxa", "Fished taxa (ISSCAAP code)", "Soaking time", "Total catch weight", "Presence of an acoustic repellent")

setDF(TableS1)

TableS1 = rbind(colnames(TableS1), TableS1)

colnames(TableS1) = c("", "", "Sex (logistic regression)", "Body mass (linear regression)", "Body size (linear regression)", "Sex (logistic regression)", "Body mass (linear regression)", "Body size (linear regression)")

kable_classic(add_header_above(kbl(TableS1, format = "latex", vline= ""), c("", "", "Common dolphin" = 3 , "Harbour porpoise" = 3)))

```

### 2024-12-09 Short_note_captured_phenotype_journal_version.qmd
Update libraries

@@
 ```{r loading used packages and functions}
-library(stringr)#gestion chaines de caractères
-library(car)#anova ou graph d'analyse
-library(RVAideMemoire)#variance
-library(corrplot)# graphe correlation
-library(ggeffects) #effet prédit
+library(stringr)# handling string format in R 
+library(car) # Anova analysis
+library(ggeffects) # prediction of effects from regression models
 library(ggplot2) #graphs
@@
 
-library(multcomp) #multiple comparison of contrasts
-
-library(data.table) #large dataset gestion
-
-library(lme4) #mixed model
-library(blme) #bayesian mixed model (handling singular fit)
+library(data.table) #large dataset handling
+
+library(lme4) #mixed regression models
 library(DHARMa) #model diagnostic (mixed models/non linear models)
-library(MuMIn) #r-squared for glmm (Nakagawa method)
 library(rsq) #diverse rsquared including Nagelkerke
@@
 
-library(sf) #spatial treatments
+library(sf) #spatial operations
 
@@
 
-library(MASS) #used to generate multivariate normal distribution
-
 library(lubridate) #handling date data
-
-sumna <- function(x) {
-    ret <- sum(x, na.rm = TRUE)
-    if(ret == 0 && all(is.na(x))) {NA_real_} else {ret}
-}
 
@@
 
-ggplot(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",], aes(BODY_LENGTH, SEA_STATE_grouped)) + geom_violin() + geom_boxplot(width = 0.1) +  stat_compare_means(label = "p.format", comparisons = list(c("2 - belle, vagues de 0.1 à 0.5 mètres", "Calme ou ridée"), c("3 - peu agitée, vagues de 0.5 à 1.25 mètres", "Calme ou ridée"), c("4 - agitée, vagues de 1.25 à 2.5 mètres", "2 - belle, vagues de 0.1 à 0.5 mètres"), c("Forte à grosse", "Calme ou ridée")), size=9) + ggtitle("Common dolphin - BL model - OBSMER dataset") +  labs(y = "Sea conditions (Douglas scale)", x = "Captured individuals' body length (cm)") +  theme(text = element_text(size = 30)) + scale_y_discrete(labels=c("Calm (0-1)", "Smooth (2)", "Slight (3)", "Moderate (4)", "Rough (>5)"))
+ggplot(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",], aes(BODY_LENGTH, SEA_STATE_grouped)) + 
+  geom_violin() + geom_boxplot(width = 0.1) +
+  stat_compare_means(label = "p.format", comparisons = list(c("2 - Smooth", "0, 1 - Calm"), c("3 - Slight", "0, 1 - Calm"), c("4 - Moderate", "2 - Smooth"), c("5, 6 - Rough to very rough", "0, 1 - Calm")), size=9) +
+  ggtitle("Common dolphin - BL model - OBSMER dataset") +
+  labs(y = "Sea conditions (Douglas scale)", x = "Captured individuals' body length (cm)") +  theme(text = element_text(size = 30)) + scale_y_discrete(labels=c("Calm (0-1)", "Smooth (2)", "Slight (3)", "Moderate (4)", "Rough (>5)"))
 
@@
 setDT(TableS7)
-TableS7[, lapply(.SD, function(x) ifelse(str_detect(x, "\n"), str_replace_all(x, "\n", " "), x))]
+TableS7 = TableS7[, lapply(.SD, function(x) ifelse(str_detect(x, "\n"), str_replace_all(x, "\n", " "), x))]
 setDF(TableS7)
 
- kable_classic(collapse_rows(kbl(TableS7, format = "latex"), columns = 1:2, target = 1, valign= "middle"))
+ kable_classic(collapse_rows(kbl(TableS7, format = "latex"), columns = 1:2, target = 1, valign = "middle"))
 

### 2024-12-10 Short_note_captured_phenotype_journal_version.qmd
Adding supplementary graphs

@@
 table(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TIMING_OP)
-kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH~OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TIMING_OP)
+kruskal.test(OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$BODY_LENGTH ~ OBSMER_Bycatch[!is.na(BODY_LENGTH) & SPECIES == "Delphinus delphis",]$TIMING_OP)
+
+table(OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$TRIMESTER)
+table(OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$ICES_DIV_grouped)
+chisq.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$TRIMESTER)
+chisq.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$ICES_DIV_grouped)
+# For the interpretation of sea state effects: trimester and spatial areas are highly correlated with sea conditions: calm sea conditions were mostly reported in the Bay of Biscay and rough sea conditions mainly in the Channel Sea; these two latter conditions were also mostly reported in winter
 
@@
 
+ggplot(data = Decl_Obs_Merged_Bycatch_BM_PHO, aes(BODY_MASS, MESH_SIZE, color=TARGETED_ISSCAAP)) + geom_jitter()
+# For the interpretation of the latter effect: a large majority of large mesh size (>200mm) are associated with targeting miscellaneous demersal fishes (i.e. mainly mofishes or Zeidae)
+
 simulateResiduals(fittedModel = model_BM_PHO, plot = T) 
@@
 
-# gear speed, dimension and depth were discarded from analyses due to too many missing values and/or because they were too well described by other variables (in this latter case including these varaibles in regression models resulted in too high VIF values)
+# gear speed, dimension and depth were discarded from analyses due to too many missing values and/or because they were too well described by other variables (in this latter case including these variables in regression models resulted in too high VIF values)
 
@@
 
-table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & TARGETED_ISSCAAP %in% c("31 - Pleuronectiformes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)
+table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)
 # in this specific case there is rare categories in both categorical variables: removing rare categories in targeted ISSCAAP groups categories lead to the presence of too rare categories in used fishing gears.
@@
 ggsave("Figure_4B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+
+ggplot(data = OBSMER_Bycatch_BL_PHO, aes(BODY_MASS, MESH_SIZE, color = TARGETED_ISSCAAP)) + geom_jitter()
+# For the interpretation of the latter effect: a large majority of large mesh size (>200mm) are associated with targeting miscellaneous demersal fishes (i.e. mainly mofishes or Zeidae)
 

### 2024-12-11 Short_note_captured_phenotype_journal_version.qmd
Corrections supplementary graphs

@@
 
+ggplot(data = Decl_Obs_Merged_Bycatch_BM_DCO, aes(BODY_MASS, MESH_SIZE, color=TARGETED_ISSCAAP)) + geom_jitter()
+# For the interpretation of the latter effect: a large majority of large mesh size (>200mm) are associated with targeting miscellaneous demersal fishes (i.e. mainly mofishes or Zeidae)
+
 simulateResiduals(fittedModel = model_BM_DCO, plot = T) 
@@
 
-ggplot(data = OBSMER_Bycatch_BL_PHO, aes(BODY_MASS, MESH_SIZE, color = TARGETED_ISSCAAP)) + geom_jitter()
+ggplot(data = OBSMER_Bycatch_BL_PHO, aes(BODY_LENGTH, MESH_SIZE, color = TARGETED_ISSCAAP)) + geom_jitter()
 # For the interpretation of the latter effect: a large majority of large mesh size (>200mm) are associated with targeting miscellaneous demersal fishes (i.e. mainly mofishes or Zeidae)

### 2025-01-30 Manuscript_data_analyses.qmd
Adding definitive code version

--- 2024-12-11 Short_note_captured_phenotype_journal_version.qmd
@@
 
-Strandings <- fread("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/Strandings_journal_version.csv")
+Strandings <- fread("~/Strandings_journal_version.csv")
 
@@
 
-Declaration_Bycatch = fread("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/Declaration_Bycatch_journal_version.csv")
+Declaration_Bycatch = fread("~/Declaration_Bycatch_journal_version.csv")
 
@@
 
-OBSMER_Bycatch <- fread("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/OBSMER_Bycatch_journal_version.csv")
+OBSMER_Bycatch <- fread("~/OBSMER_Bycatch_journal_version.csv")
 
@@
 
-# paramètres à garder:
-
-# Decl_Obs_Merged_Bycatch = all_merged[, c("ORIGIN", "ESP_LIB_FAO", "SECT_COD_SACROIS_NIV3", "SECT_COD_SACROIS_NIV3b", "TRIMESTRE", "AN", "YEAR_DCO", "YEAR_PHO", "ENGIN2", "ENGIN4", "MAILLAGE", "TARGET_TAXON", "TARGET_TAXON3", "TARGET_TAXON4", "poids_kg")]
-
 ```
@@
 
-apply(Declaration_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),], 2, function(x) table(is.na(x))) # in particular, targeted taxa were reported in ~68% of cases (very few missing values otherwise)
+apply(Declaration_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"),], 2, function(x) table(is.na(x))) # in particular, targeted taxa were reported in ~75% of cases (very few missing values otherwise)
 
@@
 
-observation_effort <- readRDS("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/observation_effort.rds")
+observation_effort <- readRDS("~/observation_effort.rds")
 
@@
 
-allgeo <- readRDS("~/ATER PAU 2024/Papier postdoc sensibilité bycatch/allgeo_Journal_version.rds") # map of ICES division of interest
+allgeo <- readRDS("~/allgeo.rds") # spatial information file
+
+allgeo=allgeo[allgeo$offname %in% c("27.4.c", "27.7.d", "27.7.e", "27.7.f", "27.7.g", "27.7.h", "27.7.j", "27.7.k", "27.8.a", "27.8.b", "27.8.c", "27.8.d"),] #selection of only ices division along the French coast
 
@@
 
-ggsave("Figure_S2.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S3.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S3A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S4A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S3B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S4B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S3C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S4C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S4A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S5A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S4B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S5B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S4C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S5C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S5.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S6.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S6.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S7.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S10A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S12A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S10B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S12B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
   ylab("By-caught individual's sex (probability)") +
+  xlab("Trimester") +
   theme(text = element_text(size = 30))
@@
 
-ggplot(dcast(na.omit(rbind(OBSMER_Bycatch[!is.na(SEX) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(SEX, SPECIES, "Captures")], Strandings[!is.na(SEX), .(SEX, SPECIES, "Strandings")])), SEX + SPECIES + V3 ~ ., fun.aggregate = length), aes( x =V3, y = .)) +
+ggplot(dcast(na.omit(rbind(OBSMER_Bycatch[!is.na(SEX) & SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(SEX, SPECIES, "Declared/observed\nby-catches")], Strandings[!is.na(SEX), .(SEX, SPECIES, "Strandings")])), SEX + SPECIES + V3 ~ ., fun.aggregate = length), aes( x =V3, y = .)) +
   geom_bar(aes(fill = SEX), stat = "identity", position = "dodge") + 
@@
 
-ggplot(rbind(Decl_Obs_Merged_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(BODY_MASS, SPECIES, "Captures")], Strandings[!is.na(SEX), .(BODY_MASS, SPECIES, paste0("Strandings,\n", SEX))]), aes( x = V3, y = BODY_MASS)) +
+ggplot(rbind(Decl_Obs_Merged_Bycatch[SPECIES %in% c("Delphinus delphis", "Phocoena phocoena"), .(BODY_MASS, SPECIES, "Declared/observed\nby-catches")], Strandings[!is.na(SEX), .(BODY_MASS, SPECIES, paste0("Strandings,\n", SEX))]), aes( x = V3, y = BODY_MASS)) +
   geom_violin() + geom_boxplot(width = 0.1) +  
-  ggtitle("Body mass of captured/stranded individuals") + 
+  ggtitle("Body mass of by-caught/stranded individuals") + 
   labs(x = "Species, data sources and sex", y = "Body mass (kg)") + 
@@
 
-ggplot(rbind(OBSMER_Bycatch[, .(paste0("*", SPECIES, "*", ", ", SEX, "<br>Captures"), BODY_LENGTH)], Strandings[, .(paste0("*", SPECIES, "*", ", ", SEX, "<br>Strandings"), BODY_LENGTH)]), aes( x = V1, y = BODY_LENGTH)) +
+ggplot(rbind(OBSMER_Bycatch[, .(paste0("*", SPECIES, "*", ", ", SEX, "<br>Declared/observed by-catches"), BODY_LENGTH)], Strandings[, .(paste0("*", SPECIES, "*", ", ", SEX, "<br>Strandings"), BODY_LENGTH)]), aes( x = V1, y = BODY_LENGTH)) +
   geom_violin(width = 1) + geom_boxplot(width = 0.1) + geom_point(data=biblio, mapping=aes(x = param, y = VALEUR, shape=ref), size=2) + 
-  ggtitle("Body length of captured/stranded individuals\n(compared with literature life-history traits)") +
-  labs(x = "Species, sex, data sources and literature estimates", y = "Body length (cm)", shape= "Bibliographic\nreferences") +  
+  ggtitle("Body length of by-caught/stranded individuals\n(compared with literature life-history traits)") +
+  labs(x = "Species, sex, data sources and literature estimates", y = "Body length (cm)", shape= "Bibliographic references") +  
   theme(text = element_text(size = 15))  +  theme(axis.text.x = element_text(angle = 60, hjust = 1))  + 
-  scale_x_discrete(limits = c("*Delphinus delphis*, Female<br>Strandings","*Delphinus delphis*, Female<br>Captures","Sexual maturity<br>(females, *Delphinus delphis*)","Asymptotic length<br>(females, *Delphinus delphis*)","*Delphinus delphis*, Male<br>Strandings","*Delphinus delphis*, Male<br>Captures","Sexual maturity<br>(males, *Delphinus delphis*)","Asymptotic length<br>(males, *Delphinus delphis*)", "*Phocoena phocoena*, Female<br>Strandings","*Phocoena phocoena*, Female<br>Captures","Sexual maturity<br>(females, *Phocoena phocoena*)","Asymptotic length<br>(females, *Phocoena phocoena*)","*Phocoena phocoena*, Male<br>Strandings","*Phocoena phocoena*, Male<br>Captures","Sexual maturity<br>(males, *Phocoena phocoena*)","Asymptotic length<br>(males, *Phocoena phocoena*)")) +
+  scale_x_discrete(limits = c("*Delphinus delphis*, Female<br>Strandings","*Delphinus delphis*, Female<br>Declared/observed by-catches","Sexual maturity<br>(females, *Delphinus delphis*)","Asymptotic length<br>(females, *Delphinus delphis*)","*Delphinus delphis*, Male<br>Strandings","*Delphinus delphis*, Male<br>Declared/observed by-catches","Sexual maturity<br>(males, *Delphinus delphis*)","Asymptotic length<br>(males, *Delphinus delphis*)", "*Phocoena phocoena*, Female<br>Strandings","*Phocoena phocoena*, Female<br>Declared/observed by-catches","Sexual maturity<br>(females, *Phocoena phocoena*)","Asymptotic length<br>(females, *Phocoena phocoena*)","*Phocoena phocoena*, Male<br>Strandings","*Phocoena phocoena*, Male<br>Declared/observed by-catches","Sexual maturity<br>(males, *Phocoena phocoena*)","Asymptotic length<br>(males, *Phocoena phocoena*)")) +
     theme(axis.text = element_markdown())
@@
 
-ggsave("Figure_S11.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S13.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S13.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S15.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S12A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S14A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S12B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S14B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S16A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S17A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
- ggsave("Figure_S16B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ ggsave("Figure_S17B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-# Susceptibility to be captured
+# Vulnerability to by-cath
 
@@
 
-ggsave("Figure_S7A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S8A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S7B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S8B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
   theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
-  scale_fill_discrete(labels = c("31 - Pleuronectiformes (N=105)", "32 - Gadiformes (N=91)", "33 - Miscellaneous coastal fishes (N=68)", "34 - Miscellaneous demersal fishes  (N=17)",  "36 - Tunas, bonitos (N=17)", "37 - Miscellaneous pelagic fishes (N=12)", "57 - Squids, cuttlefishes, octopuses (N=5)", "35 - Herrings, sardines, anchovies  (N=3)", "43 - Lobsters, spiny-rock lobsters (N=3)",  "38 - Sharks, rays (N=2)"))
-
-ggsave("Figure_S8A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+  scale_fill_discrete(labels = c("31 - Pleuronectiformes (N=109)", "32 - Gadiformes (N=92)", "33 - Miscellaneous coastal fishes (N=68)", "34 - Miscellaneous demersal fishes  (N=18)",  "36 - Tunas, bonitos (N=17)", "37 - Miscellaneous pelagic fishes (N=12)", "57 - Squids, cuttlefishes, octopuses (N=5)", "35 - Herrings, sardines, anchovies  (N=3)", "43 - Lobsters, spiny-rock lobsters (N=3)",  "38 - Sharks, rays (N=2)"))
+
+ggsave("Figure_S9A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
   theme(text = element_text(size = 20))   +    theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
-  scale_fill_discrete(labels = c("31 - Pleuronectiformes (N=44)", "34 - Miscellaneous demersal fishes  (N=32)", "32 - Gadiformes (N=12)", "33 - Miscellaneous coastal fishes (N=3)",  "38 - Sharks, rays (N=2)", "37 - Miscellaneous pelagic fishes (N=1)", "43 - Lobsters, spiny-rock lobsters (N=1)", "57 - Squids, cuttlefishes, octopuses (N=1)"))
-
-ggsave("Figure_S8B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+  scale_fill_discrete(labels = c("31 - Pleuronectiformes (N=47)", "34 - Miscellaneous demersal fishes  (N=33)", "32 - Gadiformes (N=13)", "33 - Miscellaneous coastal fishes (N=3)",  "38 - Sharks, rays (N=2)", "37 - Miscellaneous pelagic fishes (N=1)", "43 - Lobsters, spiny-rock lobsters (N=1)", "57 - Squids, cuttlefishes, octopuses (N=1)"))
+
+ggsave("Figure_S9B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S9A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S10A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S9B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S10B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 ggplot(data = Decl_Obs_Merged_Bycatch_BM_DCO, aes(BODY_MASS, MESH_SIZE, color=TARGETED_ISSCAAP)) + geom_jitter()
-# For the interpretation of the latter effect: a large majority of large mesh size (>200mm) are associated with targeting miscellaneous demersal fishes (i.e. mainly mofishes or Zeidae)
+# For the interpretation of the latter effect: a large majority of large mesh size (>200mm) are associated with targeting miscellaneous demersal fishes (i.e. mainly monkfishes or Zeidae)
 
@@
   theme(text = element_text(size = 30)) 
-ggsave("Figure_S14A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S11A.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
   theme(text = element_text(size = 30))   +  theme(axis.text.x = element_text(angle = 55, hjust = 1))
-ggsave("Figure_S14B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S11B.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S14C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S11C.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S14D.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S11D.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S14E.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S11E.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 
@@
 
-ggsave("Figure_S15.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
+ggsave("Figure_S16.pdf", device = "pdf", units = "in", width = 14, height = 8.5)
 

### 2025-01-30 Manuscript_data_curation.qmd
Adding definitive code version

---
title: "Short_note_captured_phenotype"
format: html
---

```{r loading used packages and functions}

library(stringr)# handling string format in R 
library(data.table) #large dataset gestion 
library(sf) #spatial treatments
library(lubridate) #handling date data

sumna <- function(x) {
    ret <- sum(x, na.rm = TRUE)
    if(ret == 0 && all(is.na(x))) {NA_real_} else {ret}
}

```

# Retrieval of French fishing activity information

```{r retrieval of SACROIS data}

SACROIS=data.table()

# we only retrieved here fishing activity information that will use latter on for complement our original dataset

for (i in 2000:2023) {  #the year 2023 is included to take into account the marees beginning in 2022 and ending in 2023 (including not negligible amount of data about the end of 2022)

  setwd(paste0('~/',i));

  SACROIS<-rbind(fread(paste0("NAVIRES-MOIS-MAREES-JOUR-IFR_",i,".txt"), dec=",", select= c("NAVS_COD","MAREE_ID","LIEU_COD_RET_SACROIS", "SEQ_ID","DATE_SEQ","METIER_COD_SACROIS")), SACROIS);

  SACROIS=SACROIS[NAVS_COD!="",];         #removal of vessels without id
}

SACROIS[,NAVS_COD_YEAR:=paste(SACROIS[,NAVS_COD], str_sub(SACROIS[,DATE_SEQ],7,10), sep="_")]  #produce a unique ID per vessel ad per year

#minimal requirement for being able to run the previous section: 16GB of RAM

#important notice: the last studied year (here 2022) does not take into account overlapping marees (between this year and the following one): have to be corrected by taking the following year into account during computation and then removing it at the end !!!! Important to do  because represent a non-negligible amount of data at the end of the year

```

# Data curation

The purpose of this code is to format the data so that it can be analysed later. Three datasets were used: the fisheries observer data (OBSMER data) and the French data on reported fishing activities (SACROIS data, which also includes the skipper declaration of by-catches). We prepared data to retain only variables used for analysis, time and areas of interest, while removing potentially error-prone data.

## Declaration data

```{r declaration data}

### We used all available skipper declaration data, from 2019 to 2023:

DECL_BYC=data.table()

for (i in 2019:2023) {
  setwd(paste0('~/',i));
  DECL_BYC<-rbind(fread(paste0("CAPTURES-ACC-IFR_",i,".txt"), dec=",", encoding="Latin-1", colClasses=c("SECT_COD_SACROIS_NIV5"="character"))[,-c("DATE_SEQ", "ZSV_COD")], DECL_BYC)
}

### Data selection

DECL_BYC=DECL_BYC[SECT_COD_SACROIS_NIV1==27,] # We only retained data from Northeast Atlantic ("27" ICES zone)

DECL_BYC=DECL_BYC[ESP_LIB_FAO %in% c("Dauphin à flancs blancs Atlan.", "Dauphin bleu et blanc","Dauphin commun","Dauphin commun a bec large","Dauphin tacheté de l'Atlantiq.","Dauphins nca","Grand dauphin", "Marsouin commun")] # we only retained by-catch data on small cetacean

DECL_BYC$NAVS_COD_YEAR=paste(DECL_BYC[,NAVS_COD], DECL_BYC[,AN], sep="_") # creation of an unique ID per vessel and per year

# there is no individual traits measurements for declaration of multiple capture, we then discarded such declarations from analyses:

DECL_BYC=DECL_BYC[NB_INDIVIDUS==1,]

#### Corrections in data format:

# We discarded all data with individuals weighing less than 3 kg or more than 300 kg (range of values outside the weight ranges existing in natural conditions)

DECL_BYC[QUANT_POIDS_VIF_CACC_DECLARES<=3, QUANT_POIDS_VIF_CACC_DECLARES:=NA]
DECL_BYC[QUANT_POIDS_VIF_CACC_DECLARES>=300, QUANT_POIDS_VIF_CACC_DECLARES:=NA]

# Extracting targeted taxa from métiers declarations:

DECL_BYC$TARGET_TAXON=str_sub(DECL_BYC$METIER_COD_SACROIS,4,6)

# Replacement of empty strings with NA values:

DECL_BYC[ENGIN_COD=="", ENGIN_COD:=NA]
DECL_BYC[TARGET_TAXON=="", TARGET_TAXON:=NA]

# retrieval of missing targeted taxa, using global fisheries activity declaration:

for (i in unique(DECL_BYC[METIER_COD_SACROIS=="", MAREE_ID])) {
  if (length(unique(str_sub(SACROIS[MAREE_ID==i, METIER_COD_SACROIS],4,6))[unique(str_sub(SACROIS[MAREE_ID==i, METIER_COD_SACROIS],4,6))!=""])==1) {
    DECL_BYC[MAREE_ID==i, TARGET_TAXON:=unique(str_sub(SACROIS[MAREE_ID==i, METIER_COD_SACROIS],4,6))[unique(str_sub(SACROIS[MAREE_ID==i, METIER_COD_SACROIS],4,6))!=""]]
  }
}

DECL_BYC_Short_note=DECL_BYC[ !is.na(QUANT_POIDS_VIF_CACC_DECLARES), .(ESP_LIB_FAO, QUANT_POIDS_VIF_CACC_DECLARES, AN, TRIMESTRE, SECT_COD_SACROIS_NIV3, ENGIN_COD, MAILLAGE, TARGET_TAXON, MAREE_ID, NAVS_COD_YEAR)] # selection of all by-catch data with information on body mass; and of all variables used for subsequent analyses

```

## Strandings data

```{r standings data}

### We used all reported strandings data, on both common dolphin and harbour porpoise, from 2000 to 2022:

data_strandings_DEL=read.delim("data_byc_MB_2000-2022.txt", encoding = "latin1")

data_strandings_PHO=read.delim("data_byc_Pp_MB_2000-2022.txt", encoding = "latin1")

# conversion of continuous variables in numeric format:

data_strandings_DEL$long_cm=as.numeric(data_strandings_DEL$long_cm)
data_strandings_DEL$poids_kg=as.numeric(data_strandings_DEL$poids_kg)

data_strandings_PHO$long_cm=as.numeric(data_strandings_PHO$long_cm)
data_strandings_PHO$poids_kg=as.numeric(data_strandings_PHO$poids_kg)

# We discarded all data with individuals weighing more than 150 kg (range of values outside the weight ranges existing in natural conditions)

data_strandings_PHO[data_strandings_PHO$poids_kg>=150 & !is.na(data_strandings_PHO$poids_kg), ]$poids_kg=NA

# Recoding individual's sex explicitly:

data_strandings_DEL$SEXE=data_strandings_DEL$code_sexe
data_strandings_DEL[data_strandings_DEL$code_sexe==1,]$SEXE="Male"
data_strandings_DEL[data_strandings_DEL$code_sexe %in% 2:3,]$SEXE="Female"
data_strandings_DEL[data_strandings_DEL$code_sexe==0,]$SEXE=NA

data_strandings_PHO$SEXE=data_strandings_PHO$code_sexe
data_strandings_PHO[data_strandings_PHO$code_sexe==1,]$SEXE="Male"
data_strandings_PHO[data_strandings_PHO$code_sexe %in% 2:3,]$SEXE="Female"
data_strandings_PHO[data_strandings_PHO$code_sexe==0,]$SEXE=NA

# retrieving approximate dates of death (year and trimester with the greatest overlap with estimated date of death interval):

data_strandings_DEL$date_mort_max=as.Date(data_strandings_DEL$date_mort_max, format = "%d/%m/%Y")
data_strandings_DEL$date_mort_min=as.Date(data_strandings_DEL$date_mort_min, format = "%d/%m/%Y")

data_strandings_PHO$date_mort_max=as.Date(data_strandings_PHO$date_mort_max, format = "%d/%m/%Y")
data_strandings_PHO$date_mort_min=as.Date(data_strandings_PHO$date_mort_min, format = "%d/%m/%Y")

data_strandings_DEL$date_mean_death=do.call("c", apply(data_strandings_DEL[,c("date_mort_max", "date_mort_min")], 1, function(x) mean.Date(as.Date(x)), simplify = F))

data_strandings_PHO$date_mean_death=do.call("c", apply(data_strandings_PHO[,c("date_mort_max", "date_mort_min")], 1, function(x) mean.Date(as.Date(x)), simplify = F))

data_strandings_DEL$YEAR=str_sub(data_strandings_DEL$date_mean_death, 1, 4)

data_strandings_PHO$YEAR=str_sub(data_strandings_PHO$date_mean_death, 1, 4)

data_strandings_DEL$MONTH=str_sub(data_strandings_DEL$date_mean_death, 6, 7)

data_strandings_PHO$MONTH=str_sub(data_strandings_PHO$date_mean_death, 6, 7)

data_strandings_DEL$Trimester=as.character(ceiling(as.integer(data_strandings_DEL$MONTH)/3))

data_strandings_PHO$Trimester=as.character(ceiling(as.integer(data_strandings_PHO$MONTH)/3))

# Using observation ID as rownames:

data_strandings_DEL$ID=rownames(data_strandings_DEL)
data_strandings_PHO$ID=rownames(data_strandings_PHO)

# Selection of areas of interest and retrieval of ICES division associated with location points of strandings:

allgeo <- readRDS("~/allgeo.rds")
allgeo2=allgeo[allgeo$offname %in% c("27.8.a", "27.8.b", "27.7.e", "27.7.d",  "27.4.c"),] #selection of only ices division along the French coast

pos=na.omit(data_strandings_DEL[,c("com_coord_lat","com_coord_long", "ID")])

pos=st_as_sf(pos, coords = c("com_coord_long", "com_coord_lat"), crs="WGS84")
posi=st_intersection(pos, allgeo2)

posi=setDT(posi[, c("ID","icesdiv")])[,-3]
setDF(posi)

data_strandings_DEL=merge(data_strandings_DEL, posi, by="ID", all.x = T)

pos=na.omit(data_strandings_PHO[,c("com_coord_lat","com_coord_long", "ID")])

pos=st_as_sf(pos, coords = c("com_coord_long", "com_coord_lat"), crs="WGS84")
posi=st_intersection(pos, allgeo2)

posi=setDT(posi[, c("ID","icesdiv")])[,-3]
setDF(posi)

data_strandings_PHO=merge(data_strandings_PHO, posi, by="ID", all.x = T)

rm(list=c("pos", "posi"))

setDT(data_strandings_DEL)
setDT(data_strandings_PHO)

# creation of an unique dataset containing stranding data for both harbour porpoise and common dolphin

Strandings=rbind(data_strandings_DEL, data_strandings_PHO)

Strandings[lb_nom=="Delphinus delphis", lb_nom:="Common dolphin"]
Strandings[lb_nom=="Phocoena phocoena", lb_nom:="Harbour porpoise"]

# Removal of data without any phenotypic measure:

Strandings=Strandings[!is.na(SEXE) | !is.na(long_cm) | !is.na(poids_kg)]

```

## OBSMER data

```{r OBSMER data}

#### Extraction of data of interest:

OBSMER_BYC<-read.csv2("CAPTURES.csv", fileEncoding="ISO-8859-3")

setDT(OBSMER_BYC)

OBSMER_BYC=OBSMER_BYC[ESPECE %in% c("Delphinus delphis", "Stenella coeruleoalba","Tursiops truncatus","Globicephala melas","Grampus griseus","Delphinidae","Phocoena phocoena"),] # selection of small cetacean species by-catch

strat<-read.csv2("STRATEGIE.csv", fileEncoding="ISO-8859-3")

setDT(strat)

OBSMER_BYC=merge(OBSMER_BYC, strat[,.(ID_VRAC,ID_OP)], by="ID_VRAC", all.x = T, all.y=F) #retrieval of information on fishing activity during fishing operation

operation<-read.csv2("OPERATION_PECHE.csv", fileEncoding="ISO-8859-3")

setDT(operation)

OBSMER_BYC=merge(OBSMER_BYC, operation, by="ID_OP", all.x = T, all.y=F) #retrieval of information on fishing operation

OBSMER_BYC=OBSMER_BYC[str_detect(ZONE,"27"),] # selection of data from Northeast Atlantic ("27" ICES zone)

marees<-read.csv2("MAREE.csv", fileEncoding="ISO-8859-3")

setDT(marees)

OBSMER_BYC=merge(OBSMER_BYC, marees, by="ID_MAREE", all.x = T, all.y=F)

OBSMER_BYC=OBSMER_BYC[VALID_PROGRAMME=="Données exploitables",] # keeping only data validated by data curator

# removing variables not used in subsequent analysis:

OBSMER_BYC=OBSMER_BYC[ ,c("TYPE","TYPE.x","TYPE.y","ESPECE_COM","CAT","PRESENTATION","TAUX_ECH","WEIGHT_RTP","PRODUCT_DESTINATION","CONVERSION_COEFFICIENT","ALL_TAUX_ECH","METIER1","METIER2","METIER3","METIER_DCF6","DISPO_SELECT2","CAPTURE","MARINE_LITTER_LANDING") := NULL]

#renaming confusing variables:

colnames(OBSMER_BYC)[colnames(OBSMER_BYC) %in% c("COMMENTS", "DATE_FIN.x", "POIDS_REF_CAPT", "COMMENTAIRE.x", "DATE_DEBUT", "DATE_FIN.y", "COMMENTAIRE.y")]=c("COMMENTS_CAPT", "DATE_FIN_OP", "POIDS_REF_CAPT_OP", "COMMENTS_OP", "DATE_DEBUT_MAREE", "DATE_FIN_MAREE", "COMMENTS_MAREE")

mesure<-read.csv2("MESURES.csv", fileEncoding="ISO-8859-3")

setDT(mesure)

OBSMER_BYC=merge(OBSMER_BYC, mesure[ID_CAPTURE %in% OBSMER_BYC$ID_CAPTURE,], by="ID_CAPTURE", all = T)  #retrieval of individual measurements

# removing measured variables not used in subsequent analysis:

OBSMER_BYC=OBSMER_BYC[ ,c("TYPE","TAUX_ECH","WEIGHT_RTP","VALEUR_ALPHANUMERIQUE") := NULL]

#### retrieving catch weights of the fishing operation:

capt<-read.csv2("CAPTURES.csv", fileEncoding="ISO-8859-3")
setDT(capt)

capt=merge(capt, strat[,.(ID_OP, ID_VRAC)])
capt=capt[(ID_VRAC %in% strat[ID_OP %in% OBSMER_BYC$ID_OP, ID_VRAC]) & !(ID_CAPTURE %in% OBSMER_BYC$ID_CAPTURE)] #retrieving catch information (on all taxa other than small cetaceans) on all fishing operation with observed by-catch

Code_FAO<-fread("ISIH-504549-espece_fao-20230320105303.txt") # table of correspondence between FAO nomenclature and taxa names

capt=capt[!(str_detect(ESPECE, "Macrodéchets|Inerte|Cetorhinus maximus|Animaux|Halichoerus grypus|Uria aalge"))] #removing catch information on waste, other megafauna bycatch (not commercialized species), or undetermined taxa

# retrieving catch weight for each captured taxa

capt[POIDS_REF=="", POIDS_REF:=WEIGHT_RTP]
capt[, POIDS_REF:=as.numeric(POIDS_REF)]
capt[, TAUX_ECH:=as.numeric(TAUX_ECH)]
capt[, CONVERSION_COEFFICIENT:=as.numeric(CONVERSION_COEFFICIENT)]
capt[, POIDS_REF:=(POIDS_REF/sapply(str_replace_all(str_replace_all(capt$ALL_TAUX_ECH, ",", "."), "#", "*"), function(x) eval(parse(text=x))))*CONVERSION_COEFFICIENT]

# updating out of date taxa names:

capt[ESPECE=="Trigloporus lastoviza", ESPECE:="Chelidonichthys lastoviza"]
capt[ESPECE=="Rhombosepion orbignyanum", ESPECE:="Sepia orbignyana"]

# For all taxa at the species level, retrieval of the higher order classification (spp.):

capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")) & !is.na(ESPECE), SPP:= 
             capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")) & !is.na(ESPECE), Code_FAO[ESPF_SCI_LIB==ESPECE, ESPF_PARENT_SCI_LIB], by=
                          row.names(capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")) & !is.na(ESPECE)])]$V1]

# harmonizing nomenclature at the group of species (i.e. spp, mostly genus) level:

capt[ESPECE=="Thunnus", SPP:= "Thunnus spp"]
capt[ESPECE=="Solea", SPP:= "Solea spp"]
capt[ESPECE=="Lophius", SPP:= "Lophius spp"]
capt[ESPECE=="Mullus", SPP:= "Mullus spp"]
capt[ESPECE=="Octopus", SPP:= "Octopus spp"]
capt[ESPECE=="Liza", SPP:= "Liza spp"]
capt[ESPECE=="Sepia", SPP:= "Sepia spp"]
capt[ESPECE=="Mustelus", SPP:= "Mustelus spp"]
capt[ESPECE=="Chelidonichthys", SPP:= "Chelidonichthys spp"]

# For all taxa at the species level, retrieval of the family classification level:

capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")) & !is.na(ESPECE), FAMILY:= 
             capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")), Code_FAO[ESPF_SCI_LIB==ESPECE, FAMILLE], by=
                          row.names(capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys"))])]$V1]

# retrieval of family classification for data already registered at the group of species or family level:

capt[ESPECE=="Thunnus", FAMILY:= "Scombridae"]
capt[ESPECE=="Solea", FAMILY:= "Soleidae"]
capt[ESPECE=="Lophius", FAMILY:= "Lophiidae"]
capt[ESPECE=="Mullus", FAMILY:= "Mullidae"]
capt[ESPECE=="Octopus", FAMILY:= "Octopodidae"]
capt[ESPECE=="Liza", FAMILY:= "Mugilidae"]
capt[ESPECE=="Sepia", FAMILY:= "Sepiidae"]
capt[ESPECE=="Mustelus", FAMILY:= "Triakidae"]
capt[ESPECE=="Chelidonichthys", FAMILY:= "Triglidae"]
capt[ESPECE=="Mugilidae", FAMILY:= "Mugilidae"]

# For all taxa at the species level, retrieval of the order classification level:

capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")) & !is.na(ESPECE), ORDER:= 
             capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")), Code_FAO[ESPF_SCI_LIB==ESPECE, ORDRE], by=
                          row.names(capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys"))])]$V1]

# retrieval of order classification for data already registered at the group of species or family level:

capt[ESPECE=="Thunnus", ORDER:= "Perciformes"]
capt[ESPECE=="Solea", ORDER:= "Pleuronectiformes"]
capt[ESPECE=="Lophius", ORDER:= "Lophiiformes"]
capt[ESPECE=="Mullus", ORDER:= "Perciformes"]
capt[ESPECE=="Octopus", ORDER:= "Octopoda"]
capt[ESPECE=="Liza", ORDER:= "Mugiliformes"]
capt[ESPECE=="Sepia", ORDER:= "Sepiida"]
capt[ESPECE=="Mustelus", ORDER:= "Carcharhiniformes"]
capt[ESPECE=="Chelidonichthys", ORDER:= "Scorpaeniformes"]
capt[ESPECE=="Mugilidae", ORDER:= "Mugiliformes"]

# For all taxa at the species level, retrieval of the ISSCAAP FAO classification values:

capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")) & !is.na(ESPECE), ISSCAAP:= 
             capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys")), Code_FAO[ESPF_SCI_LIB==ESPECE, ESPF_ISSCAP], by=
                          row.names(capt[!(ESPECE %in% c("Thunnus","Lophius","Solea","Mugilidae","Mullus","Octopus","Liza","Sepia","Mustelus","Pisces","Actinopterygii","Chelidonichthys"))])]$V1]

# retrieval of ISSCAAP FAO classification values for data already registered at the group of species or family level:

capt[ESPECE=="Thunnus", ISSCAAP:= 36]
capt[ESPECE=="Solea", ISSCAAP:= 31]
capt[ESPECE=="Lophius", ISSCAAP:= 34]
capt[ESPECE=="Mullus", ISSCAAP:= 33]
capt[ESPECE=="Octopus", ISSCAAP:= 57]
capt[ESPECE=="Liza", ISSCAAP:= 33]
capt[ESPECE=="Sepia", ISSCAAP:= 57]
capt[ESPECE=="Mustelus", ISSCAAP:= 38]
capt[ESPECE=="Chelidonichthys", ISSCAAP:= 34]
capt[ESPECE=="Mugilidae", ISSCAAP:= 33]

# Computation of total catch weight per fishing operation:

 poids_op = data.frame(POIDS_OP = tapply(capt$POIDS_REF,capt$ID_OP,  sumna), ID_OP = as.numeric(names(tapply(capt$POIDS_REF,capt$ID_OP,  sumna))))

OBSMER_BYC = merge(OBSMER_BYC, poids_op, all=T, by="ID_OP")

### retrieving main fished taxa (by weight, for each taxonomic level):

capt[ESPECE=="Actinopterygii", POIDS_REF:=NA] # not considering unidentified fish in most fished taxa

first_sp=dcast(capt,
      ID_OP+ESPECE~., 
      fun.aggregate = sumna,
      value.var = "POIDS_REF")[!is.na(ESPECE) & !is.na(.)][,.(.==max(.),ESPECE),by = ID_OP][V1==T][!duplicated(ID_OP), .(ID_OP, ESPECE)]

first_spp=dcast(capt,
      ID_OP+SPP~., 
      fun.aggregate = sumna,
      value.var = "POIDS_REF")[!is.na(SPP) & !is.na(.)][,.(.==max(.),SPP),by = ID_OP][V1==T][!duplicated(ID_OP), .(ID_OP, SPP)]

first_family=dcast(capt,
      ID_OP+FAMILY~., 
      fun.aggregate = sumna,
      value.var = "POIDS_REF")[!is.na(FAMILY) & !is.na(.)][,.(.==max(.),FAMILY),by = ID_OP][V1==T][!duplicated(ID_OP), .(ID_OP, FAMILY)]

first_order=dcast(capt,
      ID_OP+ORDER~., 
      fun.aggregate = sumna,
      value.var = "POIDS_REF")[!is.na(ORDER) & !is.na(.)][,.(.==max(.),ORDER),by = ID_OP][V1==T][!duplicated(ID_OP), .(ID_OP, ORDER)]

first_isscaap=dcast(capt,
      ID_OP+ISSCAAP~., 
      fun.aggregate = sumna,
      value.var = "POIDS_REF")[!is.na(ISSCAAP) & !is.na(.)][,.(.==max(.),ISSCAAP),by = ID_OP][V1==T][!duplicated(ID_OP), .(ID_OP, ISSCAAP)]

OBSMER_BYC=merge(OBSMER_BYC, first_sp, all=T, by="ID_OP", suffixes = c("", "_first"))
OBSMER_BYC=merge(OBSMER_BYC, first_spp, all=T, by="ID_OP")
OBSMER_BYC=merge(OBSMER_BYC, first_family, all=T, by="ID_OP")
OBSMER_BYC=merge(OBSMER_BYC, first_order, all=T, by="ID_OP")
OBSMER_BYC=merge(OBSMER_BYC, first_isscaap, all=T, by="ID_OP")

# removing temporary variables:

rm(list=c("marees","strat","mesure","capt","poids_op","first_sp","first_spp","first_family","first_order","first_isscaap", "Code_FAO"))
gc()

# Discarding all groups with undistinguished individuals (i.e. not individually measured):

OBSMER_BYC=OBSMER_BYC[is.na(NOMBRE.x) | NOMBRE.x==1]

OBSMER_BYC=rbind(OBSMER_BYC[NOMBRE.y==1 | is.na(NOMBRE.y)], OBSMER_BYC[NOMBRE.y>1][rep(sequence(nrow(OBSMER_BYC[NOMBRE.y>1])), OBSMER_BYC[NOMBRE.y>1, NOMBRE.y]),]) # for the number indicated in measurements, number superior to one correspond to dolphin having the same measurements (we therefore repeated these lines)

# Retrieving individual IDs for each measured individuals:

# Important note: for individuals measured in groups, with the same capture ID, there is no way to associate multiple measures to the same individual (e.g. if both Circonference and Total Length are measured, there are no existing IDs to associate the two measures to one individual), in these cases we are currently forced to consider the different measures as different individuals.

setorder(OBSMER_BYC, CRITERE)

OBSMER_BYC[CRITERE=="Longueur standard", CRITERE:="Longueur totale (LT)"] # correcting an error in "CRITERE" denomination:

u=c()
for (i in unique(OBSMER_BYC$ID_CAPTURE)) {u=c(u,max(c(table(OBSMER_BYC[ID_CAPTURE==i,]$CRITERE),1)))} # u: number of individuals described per capture event
unique(OBSMER_BYC$ID_CAPTURE)[u>1] # all capture events with more than one individual described

cat=c()
for (i in unique(OBSMER_BYC$ID_CAPTURE)) {cat=c(cat, length(unique(OBSMER_BYC[ID_CAPTURE==i,]$CRITERE)))} # for each capture event, number of measurements conducted

OBSMER_BYC$ID_IND=NA_character_

for (i in 1:length(u)) {
    if (u[i]==1 | (u[i]>1 & cat[i]==1)) {
      for (j in 1:u[i]) {
        OBSMER_BYC[ID_CAPTURE==unique(OBSMER_BYC$ID_CAPTURE)[i],][which(!duplicated(CRITERE))+(j-1),]$ID_IND=
        paste(OBSMER_BYC[ID_CAPTURE==unique(OBSMER_BYC$ID_CAPTURE)[i],]$ID_CAPTURE[1],j, sep="_")
      }
    }
    else {
      for (j in 1:length(OBSMER_BYC[ID_CAPTURE==unique(OBSMER_BYC$ID_CAPTURE)[i], ID_CAPTURE])) {
        OBSMER_BYC[ID_CAPTURE==unique(OBSMER_BYC$ID_CAPTURE)[i],][j,]$ID_IND=
          paste(OBSMER_BYC[ID_CAPTURE==unique(OBSMER_BYC$ID_CAPTURE)[i],]$ID_CAPTURE[1],j, sep="_")
      } # individuals measured in groups, with the same capture ID, and with multiple measures: considering the different measures as different individuals as we are not able to reliably assigned the different measurements to each individual
    }
}

setorder(OBSMER_BYC, ANNEE, ID_NAVIRE, ID_MAREE, ID_OP, ID_CAPTURE, ID_IND) # reordering the dataset

# correction of errors (visible from both measurements and associated comments):

OBSMER_BYC[ID_OP=="111024597",]$ESPECE="Delphinidae" # error on dolphin identification: declared unidentifiable in comments

OBSMER_BYC[ID_MAREE=="17862108",]$ESPECE="Stenella coeruleoalba" # error on dolphin identification: Stenella coreualba not delphinus delphis (comments and measurements)

OBSMER_BYC[ID_MAREE=="20527097",]$ESPECE="Tursiops truncatus" # error on dolphin identification: Tursiops truncatus not delphinus delphis (comments and measurements)

OBSMER_BYC[POIDS_REF.x %in% c("0.0", "10.0", "15.0", "999.0"), POIDS_REF.x:=""] # absurd values (also when compared to size measurements, which are much more reliable as weight are guess estimates)
OBSMER_BYC[ESPECE=="Phocoena phocoena" & POIDS_REF.x=="150.0", POIDS_REF.x:=""] # discarding all data with individuals weighing more than 150 kg (range of values outside the weight ranges existing in natural conditions)

u=c()
for (i in unique(OBSMER_BYC$ID_CAPTURE)) {u=c(u,max(c(table(OBSMER_BYC[ID_CAPTURE==i,]$CRITERE),1)))}

OBSMER_BYC[POIDS_REF.x!="" & !is.na(POIDS_REF.x) & ID_CAPTURE %in% unique(OBSMER_BYC$ID_CAPTURE)[u==1], POIDS_REF.y:=POIDS_REF.x] #when there is only one individual captured and no measure of individual weight, the capture weight value can be attributed to the individual

OBSMER_BYC[POIDS_REF.y=="0.0", POIDS_REF.y:=""] # considering zero values as missing values

OBSMER_BYC[ID_CAPTURE=="12982044" & CRITERE=="Photo de la capture", VALEUR:="115.0"] 
OBSMER_BYC[ID_CAPTURE=="12982044" & CRITERE=="Photo de la capture", CRITERE:="Longueur totale (LT)"] # body size indicated in commentary (=raw data)

OBSMER_BYC[CRITERE=="Longueur totale (LT)" & VALEUR=="0.0", VALEUR:=""] # considering zero values as missing values

OBSMER_BYC[CRITERE=="Circonférence (C)" & VALEUR=="0.0", VALEUR:=""] # considering zero values as missing values

OBSMER_BYC[SEXE %in% c("", "Non sexe", "UNK"), SEXE:=NA]

OBSMER_BYC[ID_CAPTURE==14998143 & VALEUR=="104.0", VALEUR:="188.0"]   #inversion of measures...
OBSMER_BYC[ID_CAPTURE==14998143 & VALEUR=="188.0" & CRITERE=="Circonférence (C)", VALEUR:="104.0"]   #inversion of measures...

# removal of 4 measures of circonference completely aberrant (<30cm for weight >100kg, or >180cm...)
OBSMER_BYC[CRITERE=="Circonférence (C)" & as.numeric(VALEUR)<35, VALEUR:=""]
OBSMER_BYC[CRITERE=="Circonférence (C)" & as.numeric(VALEUR)>170, VALEUR:=""]

OBSMER_BYC[DISPO_REPULS=="", DISPO_REPULS:=NA]
OBSMER_BYC[ETAT_MER=="", ETAT_MER:=NA]

OBSMER_BYC$SEXE=as.factor(OBSMER_BYC$SEXE) # sex values considered as factor data
OBSMER_BYC$ANNEE=as.character(OBSMER_BYC$ANNEE) # year of observation considered as a string variable
OBSMER_BYC$ISSCAAP=as.character(OBSMER_BYC$ISSCAAP) # ISSCAAP classification considered as a string variable
OBSMER_BYC$MAILL=as.numeric(OBSMER_BYC$MAILL) # mesh size considered as a numeric variable

OBSMER_BYC[, Trimestre:=as.character(ceiling(as.numeric(str_sub(DATE_FIN_OP,4,5))/3))] # retrieving trimester of by-catch

# retrieving fishing operation timing when hauling (i.e. when fishing operation ends):

OBSMER_BYC$TIMING_OP=as.character(ceiling((hour(dmy_hms(OBSMER_BYC$DATE_FIN_OP))+1)/6))
OBSMER_BYC[TIMING_OP==1, TIMING_OP:="Night"]
OBSMER_BYC[TIMING_OP==2, TIMING_OP:="Morning"]
OBSMER_BYC[TIMING_OP==3, TIMING_OP:="Afternoon"]
OBSMER_BYC[TIMING_OP==4, TIMING_OP:="Evening"]

OBSMER_BYC_Short_note=OBSMER_BYC[CRITERE=="Longueur totale (LT)" | (is.na(CRITERE) & !is.na(POIDS_REF.y)), .(ESPECE, SEXE, POIDS_REF.y, CRITERE, VALEUR, UNITE, ETAT_MER, DUREE_OP, ANNEE, Trimestre, ZONE, ENGIN, MAILL, EFFORT, DISPO_REPULS, PROF_ENGIN, GEAR_SPEED, POIDS_OP, ESPECE_CIBLE, ESPECE_first, SPP, FAMILY, ORDER, ISSCAAP, ID_IND, TIMING_OP)] # selection of all bycatch data with information on body length and/or body mass and/or sex; and of all variables used for subsequent analyses

```

## Checking for data replication

```{r checking replication}

#### overlap with other datasets:

# trip IDs from declaration and observation datasets are not attributed the same way

OBSMER_BYC$NAVS_COD_YEAR=paste(OBSMER_BYC$ID_NAVIRE, OBSMER_BYC$ANNEE, sep="_")

probl_vessels=intersect(DECL_BYC$NAVS_COD_YEAR, OBSMER_BYC$NAVS_COD_YEAR) #some vessels with presence in both datasets

#rapid checking in potential overlap to remove vessels with substantial doubts of replication in the data:

OBSMER_BYC_simp=OBSMER_BYC[ESPECE=="Delphinus delphis", .(ID_OP, ID_MAREE, ID_NAVIRE, ANNEE, ESPECE, ZONE, RECTANGLE, PORT_DEBARQUEMENT_CODE, DATE_DEBUT_MAREE, DATE_FIN_MAREE, NAVS_COD_YEAR)][!duplicated(ID_OP),][NAVS_COD_YEAR %in% probl_vessels]

#checking if there is any overlap between the two sources of data:

OBSMER_BYC_simp$MAREE_ID=NA_integer_

for (i in 1:length(OBSMER_BYC_simp$NAVS_COD_YEAR)) {
  OBSMER_BYC_simp[i, MAREE_ID:=c(names(which.max(table(SACROIS[NAVS_COD_YEAR==OBSMER_BYC_simp[i, NAVS_COD_YEAR] & LIEU_COD_RET_SACROIS==OBSMER_BYC_simp[i, PORT_DEBARQUEMENT_CODE] & dmy_hms(DATE_SEQ)>=dmy_hms(OBSMER_BYC_simp[i, DATE_DEBUT_MAREE]) & dmy_hms(DATE_SEQ)<=dmy_hms(OBSMER_BYC_simp[i, DATE_FIN_MAREE]) & !duplicated(SEQ_ID), MAREE_ID]))), NA_integer_)[1]] #in case of overlapping trips: select the ones with the highest number of sequence included in target trip
} # retrieving the most probable declaration fishing trip ID (on the basis of landing harbour and fishing trip timing) for each vessels of potential overlap between declaration and observation data

# Important note: some trip IDs have no correspondence, probably due to SACROIS recomputation of trip

probl_tactics=intersect(DECL_BYC$MAREE_ID, OBSMER_BYC_simp$MAREE_ID) # all declaration fishing trip IDs potentially duplicated in declaration and observation datasets

probl_vessels=OBSMER_BYC_simp[is.na(MAREE_ID), NAVS_COD_YEAR] #vessels with no sufficient data to identify the fishing trip

# cases to be removed to avoid any replication (removed from declaration data, as this dataset is less precise than the observation dataset):

DECL_BYC_Short_note = DECL_BYC_Short_note[!(MAREE_ID %in% probl_tactics) & !(NAVS_COD_YEAR %in% probl_vessels), ]

DECL_BYC_Short_note=DECL_BYC_Short_note[, .(ESP_LIB_FAO, QUANT_POIDS_VIF_CACC_DECLARES, AN, TRIMESTRE, SECT_COD_SACROIS_NIV3, ENGIN_COD, MAILLAGE, TARGET_TAXON)] # selection of all variables used for subsequent analyses

```

# Exportation of data for analyses

```{r exportation data}

# Strandings:

Strandings

# Declaration data:

Declaration_Bycatch = DECL_BYC_Short_note

# Observation data (OBSMER programme):

OBSMER_BYC_Short_note

Table_correspondence = read.csv2("ISIH-504549-espece_fao-20230320105303.txt") # taxon correspondence table

Declaration_Bycatch = merge(Declaration_Bycatch, Table_correspondence[, 1:2], by.x = "TARGET_TAXON", by.y = "ESPF_COD", all.x = T, all.y = F)
OBSMER_BYC_Short_note = merge(OBSMER_BYC_Short_note[, TARGETED_TAXA_ID := str_split_i(ESPECE_CIBLE, " - ", 1)], Table_correspondence[, 1:2], by.x = "TARGETED_TAXA_ID", by.y = "ESPF_COD", all.x = T, all.y = F) # retrieving ISSCAAP id associated with targeted taxa

### Using explicit English denominations for variables (and selection of variables used subsequently for analyses when necessary):

Strandings = Strandings[, c("lb_nom", "date_mort_min", "date_mort_max", "Trimester", "YEAR", "icesdiv", "SEXE", "poids_kg", "long_cm")]

colnames(Strandings) = c("SPECIES", "MIN_DATE_DEATH", "MAX_DATE_DEATH", "TRIMESTER", "YEAR", "ICES_DIV", "SEX", "BODY_MASS", "BODY_LENGTH")

Declaration_Bycatch = Declaration_Bycatch[, .(ESP_LIB_FAO, TRIMESTRE, AN, SECT_COD_SACROIS_NIV3, ENGIN_COD, MAILLAGE, TARGET_TAXON, ESPF_ISSCAP, QUANT_POIDS_VIF_CACC_DECLARES)]

colnames(Declaration_Bycatch) = c("SPECIES", "TRIMESTER", "YEAR", "ICES_DIV", "FISHING_GEAR", "MESH_SIZE", "TARGETED_TAXA", "TARGETED_ISSCAAP", "BODY_MASS")

OBSMER_Bycatch = OBSMER_BYC_Short_note[, c("ESPECE", "Trimestre", "ANNEE",  "ZONE", "TIMING_OP", "ETAT_MER", "ENGIN", "MAILL", "EFFORT", "PROF_ENGIN", "GEAR_SPEED", "DISPO_REPULS", "DUREE_OP", "ESPECE_CIBLE", "ESPF_ISSCAP", "ESPECE_first", "SPP", "FAMILY", "ORDER", "ISSCAAP", "POIDS_OP", "SEXE", "POIDS_REF.y", "VALEUR")]

colnames(OBSMER_Bycatch) = c("SPECIES", "TRIMESTER", "YEAR", "ICES_DIV", "TIMING_OP", "SEA_STATE", "FISHING_GEAR", "MESH_SIZE", "GEAR_DIMENSION", "GEAR_DEPTH", "GEAR_SPEED", "DETERRENT_DEVICE", "DURATION_OP", "TARGETED_TAXA", "TARGETED_ISSCAAP", "MOST_FISHED_SP", "MOST_FISHED_SPP", "MOST_FISHED_FAMILY", "MOST_FISHED_ORDER", "MOST_FISHED_ISSCAAP", "TOTAL_CATCH_WEIGHT", "SEX", "BODY_MASS", "BODY_LENGTH")

# Harmonization of terminology:

Strandings[SPECIES=="Common dolphin", SPECIES:="Delphinus delphis"]
Strandings[SPECIES=="Harbour porpoise", SPECIES:="Phocoena phocoena"]

Declaration_Bycatch[SPECIES=="Dauphin commun", SPECIES:="Delphinus delphis"]
Declaration_Bycatch[SPECIES=="Marsouin commun", SPECIES:="Phocoena phocoena"]
Declaration_Bycatch[SPECIES=="Dauphins nca", SPECIES:="Delphinidae"]
Declaration_Bycatch[SPECIES=="Dauphin à flancs blancs Atlan.", SPECIES:="Lagenorhynchus acutus"]
Declaration_Bycatch[SPECIES=="Dauphin bleu et blanc", SPECIES:="Stenella coeruleoalba"]
Declaration_Bycatch[SPECIES=="Dauphin commun a bec large", SPECIES:="Delphinus capensis"]
Declaration_Bycatch[SPECIES=="Dauphin tacheté de l'Atlantiq.", SPECIES:="Stenella frontalis"]
Declaration_Bycatch[SPECIES=="Grand dauphin", SPECIES:="Tursiops truncatus"]

Declaration_Bycatch[FISHING_GEAR=="GNS", FISHING_GEAR:="GNS - Set gillnets"]
Declaration_Bycatch[FISHING_GEAR=="GTN", FISHING_GEAR:="GTN - Combined gillnets-trammel nets"]
Declaration_Bycatch[FISHING_GEAR=="GTR", FISHING_GEAR:="GTR - Trammel nets"]
Declaration_Bycatch[FISHING_GEAR=="OTB", FISHING_GEAR:="OTB - Single boat bottom otter trawls"]
Declaration_Bycatch[FISHING_GEAR=="OTM", FISHING_GEAR:="OTM - Single boat midwater otter trawls"]
Declaration_Bycatch[FISHING_GEAR=="PTM", FISHING_GEAR:="PTM - Midwater pair trawls"]
Declaration_Bycatch[FISHING_GEAR=="SDN", FISHING_GEAR:="SDN - Danish seine"]

OBSMER_Bycatch[FISHING_GEAR=="GNS - Filets maillants calés (ancrés, filets maillants de fond à une nappe)", FISHING_GEAR:="GNS - Set gillnets"]
OBSMER_Bycatch[FISHING_GEAR=="GTR - Trémails", FISHING_GEAR:="GTR - Trammel nets"]
OBSMER_Bycatch[FISHING_GEAR=="OTT - Chaluts jumeaux à panneaux", FISHING_GEAR:="OTT - Twin bottom otter trawls"]
OBSMER_Bycatch[FISHING_GEAR=="OTB - Chaluts de fond à panneaux", FISHING_GEAR:="OTB - Single boat bottom otter trawls"]
OBSMER_Bycatch[FISHING_GEAR=="OTM - Chaluts pélagiques à panneaux", FISHING_GEAR:="OTM - Single boat midwater otter trawls"]
OBSMER_Bycatch[FISHING_GEAR=="PTM - Chaluts boeufs pélagiques", FISHING_GEAR:="PTM - Midwater pair trawls"]
OBSMER_Bycatch[FISHING_GEAR=="PTB - Chaluts boeufs de fond", FISHING_GEAR:="PTB - Bottom pair trawls"]
OBSMER_Bycatch[FISHING_GEAR=="SDN - Sennes de fond danoise", FISHING_GEAR:="SDN - Danish seine"]
OBSMER_Bycatch[FISHING_GEAR=="PS - Filets tournants avec coulisse (sennes coulissantes)", FISHING_GEAR:="PS - Purse seines"]
OBSMER_Bycatch[FISHING_GEAR=="LLS - Palangres calées (fixes)", FISHING_GEAR:="LLS - Set longlines"]

Declaration_Bycatch[, TARGETED_ISSCAAP := as.character(TARGETED_ISSCAAP)]
Declaration_Bycatch[TARGETED_ISSCAAP=="31", TARGETED_ISSCAAP:="31 - Pleuronectiformes"]
Declaration_Bycatch[TARGETED_ISSCAAP=="32", TARGETED_ISSCAAP:="32 - Gadiformes"]
Declaration_Bycatch[TARGETED_ISSCAAP=="33", TARGETED_ISSCAAP:="33 - Miscellaneous coastal fishes"]
Declaration_Bycatch[TARGETED_ISSCAAP=="34", TARGETED_ISSCAAP:="34 - Miscellaneous demersal fishes"]
Declaration_Bycatch[TARGETED_ISSCAAP=="36", TARGETED_ISSCAAP:="36 - Tunas, bonitos"]
Declaration_Bycatch[TARGETED_ISSCAAP=="39", TARGETED_ISSCAAP:=NA_character_]
Declaration_Bycatch[TARGETED_ISSCAAP=="57", TARGETED_ISSCAAP:="57 - Squids, cuttlefishes, octopuses"]

OBSMER_Bycatch[, TARGETED_ISSCAAP := as.character(TARGETED_ISSCAAP)]
OBSMER_Bycatch[TARGETED_ISSCAAP=="31", TARGETED_ISSCAAP:="31 - Pleuronectiformes"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="32", TARGETED_ISSCAAP:="32 - Gadiformes"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="33", TARGETED_ISSCAAP:="33 - Miscellaneous coastal fishes"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="34", TARGETED_ISSCAAP:="34 - Miscellaneous demersal fishes"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="35", TARGETED_ISSCAAP:="35 - Herrings, sardines, anchovies"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="36", TARGETED_ISSCAAP:="36 - Tunas, bonitos"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="37", TARGETED_ISSCAAP:="37 - Miscellaneous pelagic fishes"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="38", TARGETED_ISSCAAP:="38 - Sharks, rays"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="39", TARGETED_ISSCAAP:=NA_character_]
OBSMER_Bycatch[TARGETED_ISSCAAP=="43", TARGETED_ISSCAAP:="43 - Lobsters, spiny-rock lobsters"]
OBSMER_Bycatch[TARGETED_ISSCAAP=="57", TARGETED_ISSCAAP:="57 - Squids, cuttlefishes, octopuses"]

OBSMER_Bycatch[MOST_FISHED_ISSCAAP==31, MOST_FISHED_ISSCAAP:="31 - Pleuronectiformes"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==32, MOST_FISHED_ISSCAAP:="32 - Gadiformes"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==33, MOST_FISHED_ISSCAAP:="33 - Miscellaneous coastal fishes"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==34, MOST_FISHED_ISSCAAP:="34 - Miscellaneous demersal fishes"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==35, MOST_FISHED_ISSCAAP:="35 - Herrings, sardines, anchovies"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==36, MOST_FISHED_ISSCAAP:="36 - Tunas, bonitos"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==37, MOST_FISHED_ISSCAAP:="37 - Miscellaneous pelagic fishes"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==38, MOST_FISHED_ISSCAAP:="38 - Sharks, rays"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==42, MOST_FISHED_ISSCAAP:="45 - Crabs, sea-spiders"]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==43, MOST_FISHED_ISSCAAP:="43 - Lobsters, spiny-rock lobsters "]
OBSMER_Bycatch[MOST_FISHED_ISSCAAP==57, MOST_FISHED_ISSCAAP:="57 - Squids, cuttlefishes, octopuses"]

Declaration_Bycatch[TARGETED_TAXA=="BFT", TARGETED_TAXA:="BFT - Thunnus thynnus"]
Declaration_Bycatch[TARGETED_TAXA=="BSE", TARGETED_TAXA:="BSE - Dicentrarchus spp"]
Declaration_Bycatch[TARGETED_TAXA=="CTL", TARGETED_TAXA:="CTL - Sepiidae, Sepiolidae"]
Declaration_Bycatch[TARGETED_TAXA=="GAD", TARGETED_TAXA:="GAD - Gadiformes"]
Declaration_Bycatch[TARGETED_TAXA=="HKE", TARGETED_TAXA:="HKE - Merluccius merluccius"]
Declaration_Bycatch[TARGETED_TAXA=="MNZ", TARGETED_TAXA:="MNZ - Lophius spp"]
Declaration_Bycatch[TARGETED_TAXA=="MZZ", TARGETED_TAXA:="MZZ - Osteichthyes"]
Declaration_Bycatch[TARGETED_TAXA=="POL", TARGETED_TAXA:="POL - Pollachius pollachius"]
Declaration_Bycatch[TARGETED_TAXA=="SBG", TARGETED_TAXA:="SBG - Sparus aurata"]
Declaration_Bycatch[TARGETED_TAXA=="SOX", TARGETED_TAXA:="SOX - Soleidae"]
Declaration_Bycatch[TARGETED_TAXA=="SQU", TARGETED_TAXA:="SQU - Loliginidae, Ommastrephidae"]
Declaration_Bycatch[TARGETED_TAXA=="TUR", TARGETED_TAXA:="TUR - Scophthalmus maximus"]

OBSMER_Bycatch[TARGETED_TAXA=="ALB - Thon germon", TARGETED_TAXA:="ALB - Thunnus alalunga "]
OBSMER_Bycatch[TARGETED_TAXA=="BFT - Thon rouge", TARGETED_TAXA:="BFT - Thunnus thynnus"]
OBSMER_Bycatch[TARGETED_TAXA=="BSE - Bars", TARGETED_TAXA:="BSE - Dicentrarchus spp"]
OBSMER_Bycatch[TARGETED_TAXA=="BSS - Bar commun", TARGETED_TAXA:="BSS - Dicentrarchus labrax"]
OBSMER_Bycatch[TARGETED_TAXA=="BZX - Bonites (divers)", TARGETED_TAXA:="BZX - Sarda spp"]
OBSMER_Bycatch[TARGETED_TAXA=="CTL - Seiches, sépioles nca", TARGETED_TAXA:="CTL - Sepiidae, Sepiolidae"]
OBSMER_Bycatch[TARGETED_TAXA=="FLX - Poissons plats", TARGETED_TAXA:="FLX - Pleuronectiformes"]
OBSMER_Bycatch[TARGETED_TAXA=="GAD - Gadidés", TARGETED_TAXA:="GAD - Gadiformes"]
OBSMER_Bycatch[TARGETED_TAXA=="HKE - Merlu européen", TARGETED_TAXA:="HKE - Merluccius merluccius"]
OBSMER_Bycatch[TARGETED_TAXA=="HKX - Merlus (divers)", TARGETED_TAXA:="HKX - Merluccius spp"]
OBSMER_Bycatch[TARGETED_TAXA=="JAX - Chinchards (divers)", TARGETED_TAXA:="JAX - Trachurus spp"]
OBSMER_Bycatch[TARGETED_TAXA=="JOD - Saint-Pierre", TARGETED_TAXA:="JOD - Zeus faber"]
OBSMER_Bycatch[TARGETED_TAXA=="MAX - Maquereaux (divers)", TARGETED_TAXA:="MAX - Scomber spp"]
OBSMER_Bycatch[TARGETED_TAXA=="MGR - Maigre", TARGETED_TAXA:="MGR - Argyrosomus regius"]
OBSMER_Bycatch[TARGETED_TAXA=="MNZ - Baudroies (= Lottes) (divers)", TARGETED_TAXA:="MNZ - Lophius spp"]
OBSMER_Bycatch[TARGETED_TAXA=="MZZ - Poissons", TARGETED_TAXA:="MZZ - Osteichthyes"]
OBSMER_Bycatch[TARGETED_TAXA=="NEP - Langoustine commune", TARGETED_TAXA:="NEP - Nephrops norvegicus"]
OBSMER_Bycatch[TARGETED_TAXA=="PIL - Sardine commune", TARGETED_TAXA:="PIL - Sardina pilchardus"]
OBSMER_Bycatch[TARGETED_TAXA=="POL - Lieu jaune", TARGETED_TAXA:="POL - Pollachius pollachius"]
OBSMER_Bycatch[TARGETED_TAXA=="RAJ - Raies (divers)", TARGETED_TAXA:="RAJ - Rajidae"]
OBSMER_Bycatch[TARGETED_TAXA=="SBX - Dorades, Sparidés", TARGETED_TAXA:="SBX - Sparidae"]
OBSMER_Bycatch[TARGETED_TAXA=="SOX - Soles (divers)", TARGETED_TAXA:="SOX - Soleidae"]
OBSMER_Bycatch[TARGETED_TAXA=="SQU - Calmars, Encornets", TARGETED_TAXA:="SQU - Loliginidae, Ommastrephidae"]
OBSMER_Bycatch[TARGETED_TAXA=="TUN - Thons (divers)", TARGETED_TAXA:="TUN - Thunnini"]
OBSMER_Bycatch[TARGETED_TAXA=="TUR - Turbot", TARGETED_TAXA:="TUR - Scophthalmus maximus"]
OBSMER_Bycatch[TARGETED_TAXA=="WHG - Merlan", TARGETED_TAXA:="WHG - Merlangius merlangus"]

OBSMER_Bycatch[SEA_STATE=="", SEA_STATE:=NA]
OBSMER_Bycatch[SEA_STATE=="0 - calme, vagues absentes", SEA_STATE:="0 - Calm (glassy)"]
OBSMER_Bycatch[SEA_STATE=="1 - ridée, vagues de 0 à 0.1 mètres", SEA_STATE:="1 - Calm (rippled)"]
OBSMER_Bycatch[SEA_STATE=="2 - belle, vagues de 0.1 à 0.5 mètres", SEA_STATE:="2 - Smooth"]
OBSMER_Bycatch[SEA_STATE=="3 - peu agitée, vagues de 0.5 à 1.25 mètres", SEA_STATE:="3 - Slight"]
OBSMER_Bycatch[SEA_STATE=="4 - agitée, vagues de 1.25 à 2.5 mètres", SEA_STATE:="4 - Moderate"]
OBSMER_Bycatch[SEA_STATE=="5 - forte, vagues de 2.5 à 4 mètres", SEA_STATE:="5 - Rough"]
OBSMER_Bycatch[SEA_STATE=="6 - très forte, vagues de 4 à 6 mètres", SEA_STATE:="6 - Very rough"]

OBSMER_Bycatch[DETERRENT_DEVICE=="Oui", DETERRENT_DEVICE:=TRUE]
OBSMER_Bycatch[DETERRENT_DEVICE=="Non", DETERRENT_DEVICE:=FALSE]

OBSMER_Bycatch[MOST_FISHED_SP=="Lophius", MOST_FISHED_SP:="Lophius spp"]
OBSMER_Bycatch[MOST_FISHED_SP=="Solea", MOST_FISHED_SP:="Solea spp"]
OBSMER_Bycatch[MOST_FISHED_SP=="Thunnus", MOST_FISHED_SP:="Thunnus spp"]

OBSMER_Bycatch[TIMING_OP=="Afternoon", TIMING_OP:="Afternoon (12:00 - 18:00)"]
OBSMER_Bycatch[TIMING_OP=="Evening", TIMING_OP:="Evening (18:00 - 00:00)"]
OBSMER_Bycatch[TIMING_OP=="Morning", TIMING_OP:="Morning (06:00 - 12:00)"]
OBSMER_Bycatch[TIMING_OP=="Night", TIMING_OP:="Night (00:00 - 06:00)"]

OBSMER_Bycatch[SEX=="Femelle", SEX:="Female"]
OBSMER_Bycatch[SEX=="Male", SEX:="Male"]

# Harmonization of variable format:

str(Strandings)
str(Declaration_Bycatch)
str(OBSMER_Bycatch)

Declaration_Bycatch[, TRIMESTER:=as.character(TRIMESTER)]
Declaration_Bycatch[, YEAR:=as.character(YEAR)]
OBSMER_Bycatch[, TRIMESTER:=as.character(TRIMESTER)]
OBSMER_Bycatch[, YEAR:=as.character(YEAR)]
OBSMER_Bycatch[, GEAR_SPEED:=as.numeric(GEAR_SPEED)]
OBSMER_Bycatch[, DETERRENT_DEVICE:=as.logical(DETERRENT_DEVICE)]
OBSMER_Bycatch[, BODY_MASS:=as.numeric(BODY_MASS)]
OBSMER_Bycatch[, BODY_LENGTH:=as.numeric(BODY_LENGTH)]

# Observation effort data (OBSMER) by external variables:

observation_effort = setDT(operation[, c("DUREE_OP", "ZONE", "DATE_FIN", "ETAT_MER")])

colnames(observation_effort) = c("DURATION_OP", "ICES_DIV", "END_TIMING", "SEA_STATE")

observation_effort[SEA_STATE=="", SEA_STATE:=NA]
observation_effort[SEA_STATE=="0 - calme, vagues absentes", SEA_STATE:="0 - Calm (glassy)"]
observation_effort[SEA_STATE=="1 - ridée, vagues de 0 à 0.1 mètres", SEA_STATE:="1 - Calm (rippled)"]
observation_effort[SEA_STATE=="2 - belle, vagues de 0.1 à 0.5 mètres", SEA_STATE:="2 - Smooth"]
observation_effort[SEA_STATE=="3 - peu agitée, vagues de 0.5 à 1.25 mètres", SEA_STATE:="3 - Slight"]
observation_effort[SEA_STATE=="4 - agitée, vagues de 1.25 à 2.5 mètres", SEA_STATE:="4 - Moderate"]
observation_effort[SEA_STATE=="5 - forte, vagues de 2.5 à 4 mètres", SEA_STATE:="5 - Rough"]
observation_effort[SEA_STATE=="6 - très forte, vagues de 4 à 6 mètres", SEA_STATE:="6 - Very rough"]

# exporting dataset used for subsequent analyses:

saveRDS(observation_effort, "observation_effort.rds")
saveRDS(Strandings, "Strandings_journal_version.rds")
saveRDS(Declaration_Bycatch, "Declaration_Bycatch_journal_version.rds")
saveRDS(OBSMER_Bycatch, "OBSMER_Bycatch_journal_version.rds")

write.csv(Strandings, "Strandings_journal_version.csv", row.names = F)
write.csv(Declaration_Bycatch, "Declaration_Bycatch_journal_version.csv", row.names = F)
write.csv(OBSMER_Bycatch, "OBSMER_Bycatch_journal_version.csv", row.names = F)

```

### 2025-02-03 Manuscript_data_analyses.qmd
Last formatting changes

@@
 
-library(texreg) #export table in Latex format
-library(kableExtra)
+library(texreg) 
+library(kableExtra) #export table in Latex format
 
@@
 Declaration_Bycatch[TARGETED_TAXA == "MZZ - Osteichthyes", TARGETED_TAXA:=NA]
-OBSMER_Bycatch[TARGETED_TAXA == "MZZ - Osteichthyes", TARGETED_TAXA:=NA] # we considered the declarations of Osteichtyes to be equivalent to DECLARATIONS OF an unknown taxon, and therefore considered these values equivalent to NA for our analyses
-
-# Merging data in common between declaration and observation datasets:
+OBSMER_Bycatch[TARGETED_TAXA == "MZZ - Osteichthyes", TARGETED_TAXA:=NA] # we considered the declarations of Osteichtyes to be equivalent to declarations of an unknown taxon, and therefore considered these values equivalent to NA for our analyses
+
+# Merging data shared by declaration and observation datasets:
 
@@
 
-# Merging data in common between declaration and observation datasets, by only considering explanatory variables:
+# Merging data shared by declaration and observation datasets, but only considering explanatory variables:
 
@@
 
-# Maps of corrected abundance:
+# Map of corrected abundance:
 
@@
 
-allgeo=allgeo[allgeo$offname %in% c("27.4.c", "27.7.d", "27.7.e", "27.7.f", "27.7.g", "27.7.h", "27.7.j", "27.7.k", "27.8.a", "27.8.b", "27.8.c", "27.8.d"),] #selection of only ices division along the French coast
+allgeo=allgeo[allgeo$offname %in% c("27.4.c", "27.7.d", "27.7.e", "27.7.f", "27.7.g", "27.7.h", "27.7.j", "27.7.k", "27.8.a", "27.8.b", "27.8.c", "27.8.d"),] #selection of sampled ices divisions
 
@@
 
-We have separated the analysis of stranding data from the declaration and observation data because they may be partially redundant (with no way to verify this), and because their collection is fundamentally different (only some of the captured dolphins strand, whereas theoretically we should detect all captures through observation and reporting).
-
-For each data sets, we first described the distribution of studied traits and compared the trait values of by-caught and stranded individuals with key population values reported in the literature for each of the species studied (for both sexes: body length at maturity, asymptotic adult body length).
+We have separated the analysis of stranding data from the declaration and observation data because they may be partially redundant (with no way to verify this), and because their collection is fundamentally different (only some of the captured dolphins actually strand, whereas theoretically we should detect all captures through observation and reporting).
+
+For each dataset, we first described the distribution of studied traits and compared the trait values of by-caught and stranded individuals with key population values reported in the literature for each of the species studied (for both sexes: body length at maturity, asymptotic adult body length).
 We then examined whether different phenotypic sensitivity profiles emerged in space or time by implementing a generalised linear model for each of the traits studied (linear model for body length, logistic regression for sex) as a function of year, trimester and ICES division of by-catch.
@@
 
-# Creation of datasets per species:
+# Creation of species-specific datasets:
 
@@
 
-# key population values reported in the literature for each of the species studied (for both sexes: body length at maturity, asymptotic adult body length)
+# biblio: key population values reported in the literature for each of the species studied (for both sexes: body length at maturity, asymptotic adult body length)
 
@@
 
-# some years appears quite rarely and the number of categories associated with the variable is high
+# some years appear quite rarely and the number of categories associated with the variable is high
 
@@
 
-# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in this specific case
+# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in our specific case (body mass values are guess estimations)
 
@@
 
-# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in this specific case
+# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in our specific case
 
@@
 
-### Table combining pairwise contrast comparison test for each significant variable either in models studying sex distribution (S3) or size distribution (S4):
+### Table combining pairwise contrast comparison tests for each significant variable either in models studying sex distribution (S3) or size distribution (S4):
 
@@
 # The distribution of the number of captures per ICES division is highly variable, with some underrepresented divisions.
-Due to rare categories and the fact that the sample size of recorded captures is too small to support analyses with categorical variables described by a high number of classes, we decided to group years into four different groups.
+Due to rare categories and the fact that the sample size of recorded captures is too small to support analyses with categorical variables described by a high number of classes, we decided to group ICES divisions into four different groups.
 
@@
 fisher.test(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped)
-table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped) # more female capture north of the bay of Biscay (not significant)
+table(OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$SEX, OBSMER_Bycatch[!is.na(SEX) & SPECIES == "Delphinus delphis"]$ICES_DIV_grouped) # more female captured north of the bay of Biscay (not significant)
 
@@
 
-# Prior to spatio-temporal test we checked whether body mass variation depended on the data sources:
+# Prior to spatio-temporal tests we checked whether body mass variation depended on the data sources:
 
@@
 table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena"]$SEA_STATE)
-# as their is some sea state categories with too little data (highest and lowest sea state measured) we grouped the most extreme conditions together for analyses:
+# as there is some sea state categories with too little data (highest and lowest sea state measured) we grouped the most extreme conditions together for analyses:
 
@@
 chisq.test(OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$SEA_STATE_grouped, OBSMER_Bycatch[SPECIES == "Delphinus delphis",]$ICES_DIV_grouped)
-# For the interpretation of sea state effects: trimester and spatial areas are highly correlated with sea conditions: calm sea conditions were mostly reported in the Bay of Biscay and rough sea conditions mainly in the Channel Sea; these two latter conditions were also mostly reported in winter
+# For the interpretation of sea state effects: trimester and spatial areas are highly correlated with sea conditions; calm sea conditions were mostly reported in the Bay of Biscay and rough sea conditions mainly in the Channel Sea, these two latter conditions were also mostly reported in winter
 
@@
 
-We then explored the susceptibility of individuals to be captured as a function of fishing technique.
+We then explored the susceptibility of individuals to be captured as a function of fishing techniques.
 These analyses initially considered only the risk of being captured (see first graphical outputs, this part was scanned quickly as it had already been thoroughly studied) and how this susceptibility varies with individual traits (i.e.
@@
 
-# We first considered only the risk of being captured as a function of fishing technique.
+# We first considered only the risk of being captured as a function of fishing techniques.
 This part was scanned quickly as it had already been thoroughly studied by other researches.
-We only produced graphical outputs illustrating the relationships between fishing behaviours and the bumber of captured individuals.
+We only produced graphical outputs illustrating the relationships between fishing behaviours and the number of captured individuals.
 
@@
 # To better illustrate the distribution of mostly fished taxa associated with by-catch, we defined the proportion of most fished species or group of species mainly associated with mostly fished ISSCAAP groups (only for ISSCAAP groups with more than 6 by-catch observations, as other groups will not be considered for analyses because too rare).
-The proportion were computed from the the above tables and the total number of data available for the most fished associated ISSCAAP group.
+The proportion were computed from the above tables and the total number of data available for the most fished associated ISSCAAP group.
 
@@
 
-We then studied how traits of by-caught individuals varied with fishing techniques used so as to identify potential phenotypic vulnerabilities.
+We then studied how traits of by-caught individuals varied with used fishing techniques, so as to identify potential phenotypic vulnerabilities.
 Three phenotypic traits were studied: the by-caught individuals' body mass (using both declaration and observation data), body length and sex (using observation data only).
@@
 
-# We began to remove too rare categories (N<6) from regression analyses (while having removed all missing data with missing values for at least one of the considered explanatory variables):
+# We began to remove too rare categories (N<6) from regression analyses (while having removed all by-catch data with missing values for at least one of the considered explanatory variables):
 
@@
 table(Decl_Obs_Merged_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(TARGETED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & TARGETED_ISSCAAP %in% c("31 - Pleuronectiformes", "32 - Gadiformes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)
-# in this specific case there is rare categories in both categorical variables: removing rare categories in targeted ISSCAAP groups categories lead to the presence of too rare categories in used fishing gears.
+# in this specific case there is rare categories in both categorical variables: removing rare categories in targeted ISSCAAP group categories lead to the presence of too rare categories in used fishing gears.
 We therefore cannot use both variables in the model.
-In this particular case we chose to favor the use of used fishing gear in explanatory variables.
+In this particular case, we chose to favor the use of used fishing gear in explanatory variables.
 
@@
 apply(OBSMER_Bycatch, 2, function(x) table(is.na(x)))
-# some variable present too many missing values to be included in analyses (as gear speed only available for trawls)
+# some variables present too many missing values to be included in analyses (as gear speed only available for trawls)
 
@@
 table(OBSMER_Bycatch[SPECIES == "Phocoena phocoena" & !is.na(BODY_LENGTH) & !is.na(MOST_FISHED_ISSCAAP) & !is.na(MESH_SIZE) & !is.na(FISHING_GEAR_grouped) & MOST_FISHED_ISSCAAP %in% c("31 - Pleuronectiformes", "34 - Miscellaneous demersal fishes")]$FISHING_GEAR_grouped)
-# in this specific case there is rare categories in both categorical variables: removing rare categories in targeted ISSCAAP groups categories lead to the presence of too rare categories in used fishing gears.
+# in this specific case there are rare categories in both categorical variables: removing rare categories in targeted ISSCAAP groups categories lead to the presence of too rare categories in used fishing gears.
 We therefore cannot use both variables in the model.
@@
 
-# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there is significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):
+# we then checked for collinearity between explanatory variables (and for the inclusion of potential random effects, only when there are significant non-parametric tests of spatio-temporal (and source) effects on the phenotype of by-catch individuals):
 
@@
 
-############ Modifier legende axe abscisse pour simplifier interpretation ?????
-
-#reordering variables for plots:
-
-# OBSMER_size_del$ISSCAAP=factor(OBSMER_size_del$ISSCAAP, levels = c("31","38","32","36","37","33","34"), labels=c("Pleuronectiformes\n(mainly soles)", "Sharks, rays", "Gadiformes\n(mainly hake and gadoids)", "Tunas, bonitos", "Miscellaneous pelagic fishes\n(mainly mackerells)", "Miscellaneous coastal fishes\n(mainly sea bass)", "Miscellaneous demersal fishes\n(mainly monkfishes or St Pierre)")) # 31:poisson plats, 32:merlus et gadidés, 33: mainly bass, 34: démerseaux (baudroies, saint pierre), 36: thon, bonite, 37: maqueraux, chinchard, 38: requins, raies
-
 ```
@@
 
-### Global table (Table S6) of vulnerabilities regression models (ANOVA, type II):
+### Global table (Table S6) of phenotypic vulnerabilities regression models (ANOVA, type II):
 
@@
 
-### Table (S7) combining pairwise contrast comparison test for each significant variable in vulnerabilities regression models:
+### Table (S7) combining pairwise contrast comparison tests for each significant variable in vulnerabilities regression models:
 

### 2025-02-03 Manuscript_data_curation.qmd
Last formatting changes

@@
 library(stringr)# handling string format in R 
-library(data.table) #large dataset gestion 
-library(sf) #spatial treatments
-library(lubridate) #handling date data
+library(data.table) # large dataset gestion 
+library(sf) # spatial treatments
+library(lubridate) # handling date data
 
@@
     if(ret == 0 && all(is.na(x))) {NA_real_} else {ret}
-}
+} # sum function returning NA when summing NA values only
 
@@
 
-# we only retrieved here fishing activity information that will use latter on for complement our original dataset
+# we only retrieved here fishing activity information that will be used latter on, to complement information from our initial datasets
 
-for (i in 2000:2023) {  #the year 2023 is included to take into account the marees beginning in 2022 and ending in 2023 (including not negligible amount of data about the end of 2022)
+for (i in 2000:2023) {  #the year 2023 is included to take into account the fishing trip beginning in 2022 and ending in 2023 (including not negligible amount of data about the end of 2022)
 
@@
 
-SACROIS[,NAVS_COD_YEAR:=paste(SACROIS[,NAVS_COD], str_sub(SACROIS[,DATE_SEQ],7,10), sep="_")]  #produce a unique ID per vessel ad per year
+SACROIS[,NAVS_COD_YEAR:=paste(SACROIS[,NAVS_COD], str_sub(SACROIS[,DATE_SEQ],7,10), sep="_")]  #produce a unique ID per vessel and per year
 
-#minimal requirement for being able to run the previous section: 16GB of RAM
-
-#important notice: the last studied year (here 2022) does not take into account overlapping marees (between this year and the following one): have to be corrected by taking the following year into account during computation and then removing it at the end !!!!
-Important to do  because represent a non-negligible amount of data at the end of the year
+#minimal requirement for being able to run the previous section of code: 16GB of RAM
 
@@
 The purpose of this code is to format the data so that it can be analysed later.
-Three datasets were used: the fisheries observer data (OBSMER data) and the French data on reported fishing activities (SACROIS data, which also includes the skipper declaration of by-catches).
-We prepared data to retain only variables used for analysis, time and areas of interest, while removing potentially error-prone data.
+Three datasets were used: the fisheries observer data (OBSMER data), the French data on reported fishing activities (SACROIS data, which also includes the skipper declaration of by-catches), and French stranding records ('Réseau national d'échouages').
+We prepared data to retain only variables used for analysis, time periods and areas of interest, while removing potentially error-prone data.
 
@@
 
-DECL_BYC=DECL_BYC[ESP_LIB_FAO %in% c("Dauphin à flancs blancs Atlan.", "Dauphin bleu et blanc","Dauphin commun","Dauphin commun a bec large","Dauphin tacheté de l'Atlantiq.","Dauphins nca","Grand dauphin", "Marsouin commun")] # we only retained by-catch data on small cetacean
+DECL_BYC=DECL_BYC[ESP_LIB_FAO %in% c("Dauphin à flancs blancs Atlan.", "Dauphin bleu et blanc","Dauphin commun","Dauphin commun a bec large","Dauphin tacheté de l'Atlantiq.","Dauphins nca","Grand dauphin", "Marsouin commun")] # We only retained by-catch data on small cetacean
 
@@
 
-# there is no individual traits measurements for declaration of multiple capture, we then discarded such declarations from analyses:
+# there are no individual traits measurements for declaration of multiple captures, we then discarded such declarations from analyses:
 
@@
 
-# We discarded all data with individuals weighing less than 3 kg or more than 300 kg (range of values outside the weight ranges existing in natural conditions)
+# We discarded all data from individuals weighing less than 3 kg or more than 300 kg (range of values outside the weight ranges existing in natural conditions)
 
@@
 
-# We discarded all data with individuals weighing more than 150 kg (range of values outside the weight ranges existing in natural conditions)
+# We discarded all data from individuals weighing more than 150 kg (range of values outside the weight ranges existing in natural conditions)
 
@@
 
-# Selection of areas of interest and retrieval of ICES division associated with location points of strandings:
+# Selection of areas of interest and retrieval of ICES divisions associated with location points of strandings:
 
@@
 
-# Removal of data without any phenotypic measure:
+# Removal of data without any phenotypic measurements:
 
@@
 
-OBSMER_BYC=merge(OBSMER_BYC, strat[,.(ID_VRAC,ID_OP)], by="ID_VRAC", all.x = T, all.y=F) #retrieval of information on fishing activity during fishing operation
+OBSMER_BYC=merge(OBSMER_BYC, strat[,.(ID_VRAC,ID_OP)], by="ID_VRAC", all.x = T, all.y=F) # retrieval of information on fishing activity during fishing operation
 
@@
 
-OBSMER_BYC=merge(OBSMER_BYC, marees, by="ID_MAREE", all.x = T, all.y=F)
+OBSMER_BYC=merge(OBSMER_BYC, marees, by="ID_MAREE", all.x = T, all.y=F) #retrieval of information on fishing trip
 
@@
 
-#renaming confusing variables:
+# renaming confusing variables:
 
@@
 
-capt=capt[!(str_detect(ESPECE, "Macrodéchets|Inerte|Cetorhinus maximus|Animaux|Halichoerus grypus|Uria aalge"))] #removing catch information on waste, other megafauna bycatch (not commercialized species), or undetermined taxa
+capt=capt[!(str_detect(ESPECE, "Macrodéchets|Inerte|Cetorhinus maximus|Animaux|Halichoerus grypus|Uria aalge"))] # removing catch information on waste, other megafauna bycatch (not commercialized species), or undetermined taxa
 
-# retrieving catch weight for each captured taxa
+# retrieving catch weight for each captured taxa:
 
@@
 
-OBSMER_BYC=rbind(OBSMER_BYC[NOMBRE.y==1 | is.na(NOMBRE.y)], OBSMER_BYC[NOMBRE.y>1][rep(sequence(nrow(OBSMER_BYC[NOMBRE.y>1])), OBSMER_BYC[NOMBRE.y>1, NOMBRE.y]),]) # for the number indicated in measurements, number superior to one correspond to dolphin having the same measurements (we therefore repeated these lines)
+OBSMER_BYC=rbind(OBSMER_BYC[NOMBRE.y==1 | is.na(NOMBRE.y)], OBSMER_BYC[NOMBRE.y>1][rep(sequence(nrow(OBSMER_BYC[NOMBRE.y>1])), OBSMER_BYC[NOMBRE.y>1, NOMBRE.y]),]) # for the number of individuals indicated in measurements, number superior to one correspond to dolphin having the same measurements (we therefore repeated these lines)
 
@@
           paste(OBSMER_BYC[ID_CAPTURE==unique(OBSMER_BYC$ID_CAPTURE)[i],]$ID_CAPTURE[1],j, sep="_")
-      } # individuals measured in groups, with the same capture ID, and with multiple measures: considering the different measures as different individuals as we are not able to reliably assigned the different measurements to each individual
+      } # individuals measured in groups, with the same capture ID, and with multiple measures: considering the different measures as different individuals as we are not able to reliably assign the different measurements to each individual
     }
@@
 OBSMER_BYC[POIDS_REF.x %in% c("0.0", "10.0", "15.0", "999.0"), POIDS_REF.x:=""] # absurd values (also when compared to size measurements, which are much more reliable as weight are guess estimates)
-OBSMER_BYC[ESPECE=="Phocoena phocoena" & POIDS_REF.x=="150.0", POIDS_REF.x:=""] # discarding all data with individuals weighing more than 150 kg (range of values outside the weight ranges existing in natural conditions)
+OBSMER_BYC[ESPECE=="Phocoena phocoena" & POIDS_REF.x=="150.0", POIDS_REF.x:=""] # discarding all data from individuals weighing more than 150 kg (range of values outside the weight ranges existing in natural conditions)
 
@@
 
-# removal of 4 measures of circonference completely aberrant (<30cm for weight >100kg, or >180cm...)
+# removal of 4 measures of circumference completely aberrant (<30cm for weight >100kg, or >180cm...)
 OBSMER_BYC[CRITERE=="Circonférence (C)" & as.numeric(VALEUR)<35, VALEUR:=""]
@@
 
+# Important note: trip IDs from declaration and observation datasets are not attributed the same way
+
 #### overlap with other datasets:
-
-# trip IDs from declaration and observation datasets are not attributed the same way
 
@@
   OBSMER_BYC_simp[i, MAREE_ID:=c(names(which.max(table(SACROIS[NAVS_COD_YEAR==OBSMER_BYC_simp[i, NAVS_COD_YEAR] & LIEU_COD_RET_SACROIS==OBSMER_BYC_simp[i, PORT_DEBARQUEMENT_CODE] & dmy_hms(DATE_SEQ)>=dmy_hms(OBSMER_BYC_simp[i, DATE_DEBUT_MAREE]) & dmy_hms(DATE_SEQ)<=dmy_hms(OBSMER_BYC_simp[i, DATE_FIN_MAREE]) & !duplicated(SEQ_ID), MAREE_ID]))), NA_integer_)[1]] #in case of overlapping trips: select the ones with the highest number of sequence included in target trip
-} # retrieving the most probable declaration fishing trip ID (on the basis of landing harbour and fishing trip timing) for each vessels of potential overlap between declaration and observation data
+} # retrieving the most probable declaration fishing trip ID (on the basis of landing harbour and fishing trip timing) for each vessel with potential overlap between declaration and observation data
 
@@
 
-# Declaration data:
+# Declaration data (from SACROIS database):
 
@@
 Declaration_Bycatch = merge(Declaration_Bycatch, Table_correspondence[, 1:2], by.x = "TARGET_TAXON", by.y = "ESPF_COD", all.x = T, all.y = F)
-OBSMER_BYC_Short_note = merge(OBSMER_BYC_Short_note[, TARGETED_TAXA_ID := str_split_i(ESPECE_CIBLE, " - ", 1)], Table_correspondence[, 1:2], by.x = "TARGETED_TAXA_ID", by.y = "ESPF_COD", all.x = T, all.y = F) # retrieving ISSCAAP id associated with targeted taxa
+OBSMER_BYC_Short_note = merge(OBSMER_BYC_Short_note[, TARGETED_TAXA_ID := str_split_i(ESPECE_CIBLE, " - ", 1)], Table_correspondence[, 1:2], by.x = "TARGETED_TAXA_ID", by.y = "ESPF_COD", all.x = T, all.y = F)
+# retrieving ISSCAAP id associated with targeted taxa
 

### 2025-02-04 Manuscript_data_analyses.qmd
Adding NA proportion for BM

@@
 
-# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in our specific case (body mass values are guess estimations)
+table(is.na(data_strandings_DEL$BODY_MASS))
+# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight, with too limited sample size: see the proportion of missing values above) and more reliable in our specific case (body mass values are guess estimations)
 
@@
 
-# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight) and more reliable in our specific case
+table(is.na(data_strandings_PHO$BODY_MASS))
+# only model on size are performed because this measure is systematically performed for stranded individuals (contrary to weight, with too limited sample size: see the proportion of missing values above) and more reliable in our specific case (body mass values are guess estimations)
 

### 2025-02-05 Manuscript_data_analyses.qmd
Last figure changes

@@
   labs(y = "Body mass (kg)", x = "Mesh size (mm)")  +
-  scale_y_continuous(limits = c(75, 300)) +
+  scale_x_continuous(limits = c(75, 300)) +
   theme(text = element_text(size = 30))

### 2025-02-05 Manuscript_data_analyses.qmd
Changing figure size

@@
 plot(ggemmeans(model_BM_PHO, terms = "MESH_SIZE"), show_data = T, jitter = 2, dot_alpha = 0.2) +
-  ggtitle("Predicted by-caught individual's body mass\nas a function of used mesh size") +
+  ggtitle("Predicted by-caught individual's BM\nas a function of used mesh size") +
   labs(y = "Body mass (kg)", x = "Mesh size (mm)")  +
   scale_x_continuous(limits = c(75, 300)) +
-  theme(text = element_text(size = 30))
+  theme(text = element_text(size = 40))
 
@@
 plot(ggemmeans(model_BL_PHO, terms = "MESH_SIZE"), show_data = T, jitter = 2, dot_alpha = 0.2) +
-  ggtitle("Predicted by-caught individual's body length\nas a function of used mesh size") +
+  ggtitle("Predicted by-caught individual's BL\nas a function of used mesh size") +
   labs(y = "Body length (cm)", x = "Mesh size (mm)")  +
-  theme(text = element_text(size = 30))
+  theme(text = element_text(size = 40))
 
@@
 plot(ggemmeans(model_BL_PHO, terms = "FISHING_GEAR_grouped"), show_data = T, jitter = c(0.7, 0), dot_alpha = 0.2) +
-  ggtitle("Predicted by-caught individual's body length\nas a function of used fishing gear") +
+  ggtitle("Predicted by-caught individual's BL\nas a function of used fishing gear") +
   labs(y = "Body length (cm)", x = "Fishing gear")  +
-  theme(text = element_text(size = 30))
+  theme(text = element_text(size = 40))
 
